---
title: "Spatial modelling of wastewater data in epidemiology --- full analysis"
subtitle: "Cleaning, spatial covariates, the mixed graphical model, and spatial autocorrelation, in one run"
author: "Marquis Pretorius --- WST795 research report"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    code_folding: show
bibliography: references.bib
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150,
  ## cache is OFF deliberately. The first three sections write .rds files that
  ## every later section reads, so a cached chunk that did not re-run would
  ## leave the document reporting one pipeline while the files on disk came
  ## from another. The whole document runs in a few minutes; correctness is
  ## worth more than that here.
  cache = FALSE
)
options(width = 105)
```

# How to run this document

Everything from the raw export to the spatial results is here, in the order it
has to happen. Knit the file and the whole analysis runs; run it chunk by chunk
and it behaves the same way. The only thing it needs in the working directory
is `full_AIARMS_df.csv`, plus `references.bib`, `app.R` and `gadm41_ZAF_4.shp`
if you want the bibliography, the app listing and the provincial map.

| Section | Chunk | Writes |
|---|---|---|
| 2 | `Data Cleaning` | `output/AIARMS_mgm.rds` |
| 3 | `Spatial component` | `output/AIARMS_mgm_spatial.rds` |
| 4 | `Mixed Graphical Models` | the fit, the network, the Moran diagnostics |
| 5 onwards | spatial autocorrelation | `mgm_spatial_bundle.rds`, `mgm_spatial_core.R` |

The last two are what `app.R` loads, so knitting this document is also how the
dashboard gets rebuilt.

Expect **under a minute** for the spatial analysis, plus whatever the model fit
takes. The bivariate sections were removed because Section 2.1 of the report
defines Moran's I, Geary's C and the local Moran and nothing else --- reporting
a statistic the methodology never defines is the kind of gap an examiner finds.

The remaining lever is the **bootstrap** in Section 4. `RUN_BOOTSTRAP <- FALSE`
while you are drafting prose --- 100 refits of the full model is minutes, and it
re-runs every knit even when nothing about the model has changed. Switch it back
on for the final build.

## The decisions this document makes

Four switches change the numbers. They sit at the top of their own sections;
they are collected here so that none of them is made by accident.

| Switch | Section | Default | What it decides |
|---|---|---|---|
| `STRICT_MISSING` | 2 | `TRUE` | Whether a blank cell in a Yes/No grid is a "No" or a non-response. Moves `Flooding` and `ManureUse`. |
| `EDU_STRICT_TERTIARY` | 2 | `FALSE` | Whether a sub-matric certificate counts as tertiary. **Decides whether `Education_3` is a finding.** |
| `EXPOSURE_UNION` | 2 | `TRUE` | Whether `Months_tested` counts ESBL months only, or months with any readable result. |
| `SITE_ENCODING` | 3 | `"dummy_full"` | Seven binary site nodes, one per fieldwork code. `"categorical"` is the identified alternative. |

Set the first three to `FALSE` and `SITE_ENCODING` to `"dummy_reference"`, and
the cleaning reproduces the original pipeline's data matrix exactly; only the
`w_ii = 0` correction to the weights matrix then differs. Running it both ways
is the sensitivity analysis the methodology section needs.

---

# Cleaning the raw export {#clean}

`01_clean_AIARMS.R`. Turns the raw household survey and the monthly culture
results into an object `mgm()` will accept: a fully numeric matrix, no missing
values, categorical variables coded as consecutive integers, and a registry
declaring each variable's type and level.

```{r Data Cleaning, purl=FALSE}
###############################################################################
# 01_clean_AIARMS.R
#
# Purpose : Turn the raw AIARMS household survey + monthly culture results into
#           a MIXED GRAPHICAL MODEL (MGM) ready object, following the data
#           requirements of Haslbeck & Waldorp (2020, JSS 93(8)), the `mgm`
#           package paper.
#
# What mgm() needs (Haslbeck & Waldorp 2020, Section 3.1):
#   1. `data`  : a fully NUMERIC matrix/data.frame, NO missing values.
#   2. `type`  : character vector, one entry per column:
#                  "g" = conditional Gaussian
#                  "p" = conditional Poisson (non-negative integer counts)
#                  "c" = conditional categorical (multinomial)
#   3. `level` : integer vector, number of categories for each variable.
#                By convention level = 1 for every "g" and "p" variable.
#   4. Categorical variables must be coded as consecutive integers.
#      - Binary variables are coded 0/1 so that `binarySign = TRUE` can define
#        a SIGN for binary-binary and binary-continuous edges.
#      - Variables with m > 2 categories are coded 1..m (no sign is definable
#        for these; the paper explains why in Section 2.2).
#
# Output: ./output/AIARMS_mgm.rds  -- a list with $data, $type, $level,
#         $labels, $groups, $groups_color, plus the untouched cleaned frame.
#
# Author: prepared for the AIARMS analysis
###############################################################################

## ---------------------------------------------------------------------------
## 0. Setup
## ---------------------------------------------------------------------------
## In a combined run the earlier sections have already written their .rds
## outputs, so nothing here needs a clean slate -- and rm(list = ls()) part way
## through a knit would delete objects the later sections rely on. Run
## standalone, it behaves exactly as before.
if (!isTRUE(getOption("knitr.in.progress"))) rm(list = ls())

pkgs <- c("dplyr", "stringr", "tidyr", "tibble")
new  <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(new)) install.packages(new)
invisible(lapply(pkgs, library, character.only = TRUE))

RAW_PATH <- "full_AIARMS_df.csv"   # <-- edit if the file lives elsewhere
OUT_DIR  <- "output"
if (!dir.exists(OUT_DIR)) dir.create(OUT_DIR)

## ---------------------------------------------------------------------------
## 0a. Cleaning switches  (each one is a decision the write-up has to defend)
## ---------------------------------------------------------------------------
## STRICT_MISSING
##   FALSE reproduces the original behaviour: a blank cell anywhere in a
##   "select all that apply" grid is read as "not ticked" and becomes 0.
##   TRUE  restricts that rule to the blocks where it is actually true of the
##   questionnaire, and treats a blank as MISSING wherever the form also
##   records an explicit "No". This matters: under FALSE, 32 blank Flooding
##   cells, 63 all-blank manure rows, 8 blank standing-water cells and 82
##   partially answered handwashing rows are silently converted to negatives
##   BEFORE the missing-data report runs, so none of them ever appear in it.
STRICT_MISSING <- TRUE

## EDU_STRICT_TERTIARY
##   FALSE reproduces the original: any response matching
##   "Diploma|Certificate|degree|BTech|Post graduate" is level 3 (tertiary).
##   That sweeps in "Certificate with <Std 10/Gr.12" (n = 18) and "Diploma
##   with <Std 10/ Gr.12" (n = 7) -- 25 households holding a sub-matric
##   certificate -- alongside Bachelor's degrees.
##   TRUE sends those 25 to level 1.
##   THIS CHANGES A HEADLINE RESULT. Education_3 has global Moran's I = 0.115
##   (p = 0.006) under FALSE and I = 0.043 (p = 0.17) under TRUE. Whichever
##   you submit, report the other as a sensitivity analysis.
EDU_STRICT_TERTIARY <- FALSE

## EXPOSURE_UNION
##   FALSE reproduces the original: Months_tested counts months with an ESBL
##   result only. That denominator is wrong for 22 of 164 households when it
##   is used as the exposure for EC or KP (their columns disagree with the
##   ESBL column on which months were cultured).
##   TRUE counts a month as cultured if ANY of the three organisms has a
##   readable result.
EXPOSURE_UNION <- TRUE

## PORTABILITY. The previous version passed fileEncoding = "UTF-8", which
## tells R to RE-ENCODE the file into the session's native encoding. In a
## non-UTF-8 locale (a plain `Rscript` under LC_CTYPE=C, and the default on
## many Windows installs) that re-encoding fails on the first curly quote and
## R returns a frame with ZERO rows after only a warning:
##     "invalid input found on input connection 'full_AIARMS_df.csv'"
## The script then dies further down with an opaque rowSums() error. Passing
## encoding = "UTF-8" instead MARKS the strings as UTF-8 without re-encoding,
## which works in every locale.
##
## A second correction to the old comment: the en-dashes in the income bands
## are ALREADY mojibake in the export ("R19201 - R38400" was written by a
## double-encoding upstream). No reader argument can undo that. What matters
## is that none of the recodes below match on those characters -- the income
## bands are matched by their leading ASCII prefix, education by ASCII
## keywords -- so the mojibake is cosmetic and harmless here. Do not "fix" it
## by editing the CSV.
raw <- read.csv(RAW_PATH, stringsAsFactors = FALSE, check.names = TRUE,
                encoding = "UTF-8", na.strings = c("", "NA", "na", "N/A"))

cat("Raw dimensions:", dim(raw), "\n")
if (nrow(raw) == 0L)
  stop("Read 0 rows from ", RAW_PATH,
       " -- check the file path and that the export is not empty.")

## ---------------------------------------------------------------------------
## 1. Small helper functions
## ---------------------------------------------------------------------------

## The survey column names are enormous ("X118..D010a..Toilet.Facilities...").
## `col()` finds the single column whose name starts with a short stem, so the
## script stays readable and survives minor re-exports of the questionnaire.
col <- function(stem) {
  hit <- grep(paste0("^", stem), names(raw), value = TRUE)
  if (length(hit) == 0) stop("No column starting with: ", stem)
  raw[[hit[1]]]
}

## Yes/No -> 1/0
yn <- function(x) ifelse(x == "Yes", 1L, ifelse(x == "No", 0L, NA_integer_))

## TRUE checkbox blocks: the export writes a value only where the box was
## ticked, so a blank genuinely means "not ticked" and 0 is correct. This is
## the case for X112 (water source), where 130 of 164 cells are blank.
yn0 <- function(x) ifelse(!is.na(x) & x == "Yes", 1L, 0L)

## Yes/No ITEMS that happen to live inside a grid. Here the export writes an
## explicit "No", so a blank is a non-response, not a negative. Under
## STRICT_MISSING it becomes NA and is handled by the imputation step in
## section 9 like every other missing cell.
yn_item <- function(x) {
  if (STRICT_MISSING) ifelse(x == "Yes", 1L, ifelse(x == "No", 0L, NA_integer_))
  else                ifelse(!is.na(x) & x == "Yes", 1L, 0L)
}

## Grab every column of a checkbox block, e.g. block("X123") returns
## X123.1..Before.eating ... X123.8..When.your.hands.appear.dirty
block <- function(stem, idx) {
  nm <- paste0("^", stem, "\\.", idx, "\\.")
  do.call(cbind, lapply(nm, function(p) {
    hit <- grep(p, names(raw), value = TRUE)
    if (length(hit) == 0) rep(NA_character_, nrow(raw)) else raw[[hit[1]]]
  }))
}

## ---------------------------------------------------------------------------
## 2. De-duplicate participants
## ---------------------------------------------------------------------------
## Study_code has 164 unique values across 171 rows (repeat survey submissions).
## An MGM assumes i.i.d. rows, so duplicated participants would inflate the
## effective sample size and bias the lasso towards keeping their edges.
dup <- duplicated(raw$Study_code)
cat("Dropping", sum(dup), "duplicate Study_code rows.\n")
raw <- raw[!dup, ]

## ---------------------------------------------------------------------------
## 3. OUTCOMES: monthly stool/rectal culture results -> carriage counts
## ---------------------------------------------------------------------------
## Each participant has up to 9 monthly results per organism. Free-text entry
## produced variants: "EC_Poss", "E.coli", "No Growth", "KP_NEg",
## "ESBL_ Ecoli", "ESBL _EC + KP", ... We normalise case/underscores/spaces and
## then apply a positive / negative rule per organism.

months <- c("March","May","June","July","August","September",
            "October","November","December")

norm <- function(x) tolower(gsub("[_[:space:]]+", " ", trimws(x)))

ec_pos <- function(x) {
  s <- norm(x)
  out <- rep(NA_integer_, length(s))
  out[grepl("no growth|neg", s)]            <- 0L
  out[is.na(out) & grepl("ec|coli", s)]     <- 1L
  out
}
kp_pos <- function(x) {
  s <- norm(x)
  out <- rep(NA_integer_, length(s))
  out[grepl("neg", s)]                      <- 0L
  out[is.na(out) & grepl("kp", s)]          <- 1L
  ## cells reading only "E. coli" are E. coli results mis-filed in a KP column
  out
}
esbl_pos <- function(x) {
  s <- norm(x)
  out <- rep(NA_integer_, length(s))
  out[grepl("neg", s)]                      <- 0L
  out[is.na(out) & grepl("esbl", s)]        <- 1L
  out
}

EC <- sapply(months, function(m) ec_pos(raw[[paste0("EC_",  m)]]))
KP <- sapply(months, function(m) kp_pos(raw[[paste0("KP_",  m)]]))
ES <- sapply(months, function(m) esbl_pos(raw[[paste0("ESBL_", m)]]))

## Number of positive months (a count -> Poisson node) and number of months
## actually cultured. Because participants were sampled an unequal number of
## times, `Months_tested` is carried into the model as a Gaussian covariate:
## mgm has no offset term, so conditioning on the exposure is the closest
## available equivalent to a rate model.
## Months actually cultured. The original definition counted months with an
## ESBL result only. Checked against the export: the EC column disagrees with
## the ESBL column about which months were cultured for 71 of 164 households
## and the KP column for 74, so an ESBL-only denominator is not a valid
## exposure for EC_pos_n or KP_pos_n. The union rule below counts a month as
## cultured if any of the three organisms returned a readable result; it
## differs from the ESBL-only count for 22 households, by at most 2 months.
months_any <- (!is.na(EC)) | (!is.na(KP)) | (!is.na(ES))

D <- data.frame(
  Study_code    = raw$Study_code,
  EC_pos_n      = rowSums(EC, na.rm = TRUE),
  KP_pos_n      = rowSums(KP, na.rm = TRUE),
  ESBL_pos_n    = rowSums(ES, na.rm = TRUE),
  Months_tested = if (EXPOSURE_UNION) rowSums(months_any) else rowSums(!is.na(ES)),
  stringsAsFactors = FALSE
)

## Four households have no readable culture result in any month. Their
## carriage counts are structural zeros, not observed zeros, and they drag
## every carriage statistic towards zero. They are KEPT so that n matches the
## survey sample, but they are flagged here so the write-up can say so and so
## that a rate analysis can exclude them.
cat("Households with zero cultured months:", sum(D$Months_tested == 0), "\n")

## Convenience binaries (not used in the default model, handy for checks)
D$ESBL_ever <- as.integer(D$ESBL_pos_n > 0)
D$KP_ever   <- as.integer(D$KP_pos_n   > 0)

## NOTE: E. coli carriage is near-saturated -- 97.1% of cultured months are
## positive (952 of 981 readable EC cells) -- so most of the variance in
## EC_pos_n comes from how many months a household was cultured
## (r = 0.93 with Months_tested). It is nonetheless INCLUDED in the model
## alongside KP and ESBL so that all three target markers are represented;
## expect its main edge to be with Months_tested.

## ---------------------------------------------------------------------------
## 4. DEMOGRAPHY AND SOCIO-ECONOMIC POSITION
## ---------------------------------------------------------------------------
D$Age <- as.numeric(col("X4\\.\\."))                                   # "g"

D$Sex <- ifelse(col("X5\\.\\.") == "Female", 1L, 0L)                   # "c", 2

## X8 asks for people EXCLUDING the respondent, so add 1 for household size.
D$HHsize <- as.numeric(col("X8\\.\\.")) + 1                            # "p"

rooms_map <- c("One"=1,"Two"=2,"Three"=3,"Four"=4,"Five"=5,
               "Six"=6,"Seven"=7,"More than seven"=8)
D$Rooms <- unname(rooms_map[col("X99\\.\\.")])                         # "p"

## Crowding = persons per room. A classic transmission-relevant exposure and
## a genuinely continuous quantity, so it enters as Gaussian.
D$Crowding <- D$HHsize / D$Rooms                                       # "g"

## Highest household education, collapsed to 3 ordered-but-nominal levels.
## MGM treats "c" variables as unordered; collapsing avoids 18 sparse states.
edu <- col("X52\\.\\.")
## A qualification whose own label says "<Std 10/Gr.12" was obtained without
## matric. Under EDU_STRICT_TERTIARY it is level 1, not level 3. See the note
## on the switch: this moves 25 households and changes the Education_3
## spatial result.
tert <- grepl("Diploma|Certificate|degree|BTech|Post graduate", edu)
if (EDU_STRICT_TERTIARY) tert <- tert & !grepl("<\\s*Std 10", edu)
D$Education <- ifelse(
  tert, 3L,
  ifelse(grepl("Grade 12|Grade 13", edu) & !grepl("Attend Grade 12", edu), 2L,
         ifelse(!is.na(edu), 1L, NA_integer_)))                        # "c", 3
# 1 = primary / incomplete secondary, 2 = completed secondary, 3 = tertiary
cat("Education levels (EDU_STRICT_TERTIARY =", EDU_STRICT_TERTIARY, "):\n")
print(table(D$Education, useNA = "ifany"))

ws <- col("X53\\.\\.")
D$WorkStatus <- ifelse(ws %in% c("Employer","Paid employee","Self-employed",
                                 "Paid domestic worker"), 1L,
                ifelse(ws == "Unemployed", 2L, 3L))                    # "c", 3
# 1 = employed, 2 = unemployed, 3 = pensioner / student / other

## Income bands are ordered, so they are far more informative as a single
## ordinal score (Gaussian) than as 11 categorical states we cannot estimate.
inc_levels <- c("No income","R1-R400","R401 - R800","R801 - R1600",
                "R1601 - R3200","R3201 - R6400","R6401 - R12800",
                "R12801 - R25600","R25601 - R51200","R51201 - R102400")
inc_raw <- col("X61\\.\\.")
D$IncomeBand <- sapply(inc_raw, function(s) {
  if (is.na(s) || grepl("Prefer not", s)) return(NA_real_)
  hit <- which(startsWith(s, inc_levels))
  if (length(hit)) hit[1] - 1 else NA_real_
}, USE.NAMES = FALSE)                                                  # "g"

D$Grant <- yn(col("X59\\.\\."))                                        # "c", 2

## ---------------------------------------------------------------------------
## 5. HOUSING, WATER, SANITATION, HYGIENE (the WASH block)
## ---------------------------------------------------------------------------
dw <- col("X95\\.\\.")
D$FormalDwelling <- as.integer(grepl("^House|^Room/flatlet", dw))      # "c", 2

## Two versions of the water question exist in the export: an old single-choice
## item (X110) and a newer checkbox block (X112.*). We take the old item where
## present and fall back to the checkboxes, then reduce to the distinction that
## actually matters for contamination risk: water inside the dwelling or not.
w_old <- col("X110\\.\\.")
w_in  <- as.integer(yn0(block("X112", 1)))
w_yd  <- as.integer(yn0(block("X112", 2)))
D$PipedInside <- ifelse(
  !is.na(w_old) & w_old == "Piped water inside the dwelling", 1L,
  ifelse(!is.na(w_old) & w_old %in% c("Piped water outside the dwelling (e.g., yard)",
                                      "Public tap/standpipe","Other"), 0L,
         ifelse(w_in == 1, 1L, ifelse(w_yd == 1, 0L, NA_integer_))))   # "c", 2

## Toilet: flush (to sewer or septic) versus pit latrine / public / none.
## The intermediate "pit with slab" group has only 14 members, too few to
## support its own categorical state at n = 164, so it joins the unimproved arm.
tl <- col("X118\\.\\.")
D$FlushToilet    <- as.integer(grepl("flush", tl, ignore.case = TRUE)) # "c", 2
D$ToiletShared   <- yn(col("X121\\.\\."))                              # "c", 2
D$ToiletFloods   <- yn(col("X120\\.\\."))                              # "c", 2
D$OpenDefecation <- as.integer(col("X122\\.\\.") != "Never")           # "c", 2

## X126.1 and X128.1 are Yes/No items, not tick boxes: the export records an
## explicit "No" (X126.1 is Yes 105 / No 56 / blank 8; X128.1 has 32 blanks).
## yn_item() therefore treats a blank as missing under STRICT_MISSING.
D$StandingWater  <- as.integer(yn_item(block("X126", 1)))              # "c", 2
D$Flooding       <- as.integer(yn_item(block("X128", 1)))              # "c", 2
D$RefuseCollected <- as.integer(startsWith(
  ifelse(is.na(col("X108\\.\\.")), "", col("X108\\.\\.")), "Removed")) # "c", 2

## Handwashing: 8 Yes/No items summed into a 0-8 practice score. A sum of
## indicators is exactly the kind of variable the Poisson node is for.
##
## These items also carry an explicit "No", and 82 of 164 households leave
## between 1 and 7 of them blank. Summing "== Yes" over the raw block counts
## every blank as a No and pushes those 82 scores down. Under STRICT_MISSING
## the score is prorated over the items the household actually answered, and
## set to NA (for section 9 to impute) when fewer than five were answered.
hw      <- block("X123", 1:8)
hw_yes  <- rowSums(hw == "Yes", na.rm = TRUE)
hw_ans  <- rowSums(!is.na(hw))
if (STRICT_MISSING) {
  D$HandwashScore <- ifelse(hw_ans >= 5L,
                            as.integer(round(8 * hw_yes / hw_ans)),
                            NA_integer_)                               # "p"
  cat("Handwashing items answered per household:\n"); print(table(hw_ans))
} else {
  D$HandwashScore <- hw_yes                                            # "p"
}

## ---------------------------------------------------------------------------
## 6. HEALTH SERVICE USE AND ANTIBIOTIC BEHAVIOUR
## ---------------------------------------------------------------------------
cf <- col("X62\\.\\.")
## "Never" has only 2 observations, so it is merged with "Yearly".
D$CareMonthly <- as.integer(cf %in% c("Weekly", "Montly"))             # "c", 2

D$HIVinHH <- yn(col("X82\\.\\."))                                      # "c", 2

ch <- col("X76\\.\\.")
D$ChronicAny <- ifelse(is.na(ch), NA_integer_,
                       as.integer(ch != "No chronic conditions"))      # "c", 2

sev <- col("X81\\.\\.")
D$DiarrSeverity <- ifelse(sev == "No one experiences diarrhoea", 1L,
                   ifelse(sev == "Mild (resolves in 1-2 days)",   2L,
                   ifelse(!is.na(sev), 3L, NA_integer_)))              # "c", 3
# 1 = no diarrhoea, 2 = mild, 3 = moderate or severe

D$AbxNoRx <- yn(col("X90\\.\\."))                                      # "c", 2

src <- col("X86\\.\\.")
D$AbxSource <- ifelse(is.na(src), NA_integer_,
               ifelse(src == "Not applicable", 1L,
               ifelse(src == "Government facility (Clinic; Community Heath Centre, Hospital)",
                      2L, 3L)))                                        # "c", 3
# 1 = does not source antibiotics, 2 = government facility only,
# 3 = private GP and/or pharmacy (with or without government)

crs <- col("X92\\.\\.")
## "Until the package is empty" (n = 2) is merged with "As advised", since both
## describe completing a full prescribed course.
D$AbxCourse <- ifelse(crs %in% c("As advised by a healthcare provider",
                                 "Until the package is empty"), 1L,
               ifelse(crs == "Until symptoms are resolved or cured", 2L,
               ifelse(crs == "I do not take antibiotics", 3L, NA_integer_)))
                                                                       # "c", 3
# 1 = completes course, 2 = stops when symptoms resolve, 3 = does not take

## ---------------------------------------------------------------------------
## 7. FOOD, ANIMALS, ENVIRONMENTAL EXPOSURE
## ---------------------------------------------------------------------------
D$StreetFood <- as.integer(col("X135\\.\\.") != "Don't consume street food")

## Meat frequency: the questionnaire changed mid-study (X129 old, X131 new) and
## both are checkbox grids of frequency x meat type. We take the HIGHEST
## frequency ticked for any meat type: 1 = never ... 6 = daily, minus 1 so the
## score starts at 0.
meat_blocks <- cbind(block("X129", 1:6), block("X131", 1:6))
meat_level  <- rep(NA_real_, nrow(raw))
for (j in 1:6) {
  ticked <- !is.na(meat_blocks[, j]) | !is.na(meat_blocks[, j + 6])
  meat_level[ticked] <- j
}
D$MeatFreq <- meat_level - 1                                           # "g"

D$AnimalsKept <- yn(col("X157\\.\\."))                                 # "c", 2
## Manure: five Yes/No items. 63 of 164 households leave all five blank --
## under the old rule every one of them became "does not use manure". Under
## STRICT_MISSING an all-blank row is missing; a partially answered row is an
## OR over the items that were answered.
mn      <- block("X150", 1:5)
mn_yes  <- rowSums(mn == "Yes", na.rm = TRUE)
mn_ans  <- rowSums(!is.na(mn))
## (braces, not a bare multi-line if/else: at the console R closes the `if`
## at the end of the first line and the stray `else` is a syntax error)
D$ManureUse <- if (STRICT_MISSING) {
  ifelse(mn_ans > 0L, as.integer(mn_yes > 0), NA_integer_)
} else {
  as.integer(mn_yes > 0)
}                                                                      # "c", 2

## ---------------------------------------------------------------------------
## 8. Variable registry: name -> type, level, group
## ---------------------------------------------------------------------------
## This registry IS the contract with mgm(). Everything downstream (fitting,
## plotting, the Shiny app) reads `type` and `level` from here, so a variable
## can never drift out of sync with its declared distribution.

registry <- tibble::tribble(
  ~var,             ~type, ~level, ~group,        ~label,
  # --- outcomes -------------------------------------------------------------
  "ESBL_pos_n",     "p",   1L,     "Carriage",    "ESBL+ months",
  "EC_pos_n",       "p",   1L,     "Carriage",    "E. coli+ months",
  "KP_pos_n",       "p",   1L,     "Carriage",    "K. pneumoniae+ months",
  "Months_tested",  "g",   1L,     "Carriage",    "Months cultured",
  # --- demography / SES -----------------------------------------------------
  "Age",            "g",   1L,     "Demographic", "Age",
  "Sex",            "c",   2L,     "Demographic", "Sex (1=F)",
  "HHsize",         "p",   1L,     "Demographic", "Household size",
  "Crowding",       "g",   1L,     "Demographic", "Persons per room",
  "Education",      "c",   3L,     "Socioeconomic","Highest education",
  "WorkStatus",     "c",   3L,     "Socioeconomic","Work status",
  "IncomeBand",     "g",   1L,     "Socioeconomic","Income band",
  "Grant",          "c",   2L,     "Socioeconomic","Receives grant",
  # --- WASH -----------------------------------------------------------------
  "FormalDwelling", "c",   2L,     "WASH",        "Formal dwelling",
  "PipedInside",    "c",   2L,     "WASH",        "Piped water indoors",
  "FlushToilet",    "c",   2L,     "WASH",        "Flush toilet",
  "ToiletShared",   "c",   2L,     "WASH",        "Shared toilet",
  "ToiletFloods",   "c",   2L,     "WASH",        "Toilet floods",
  "OpenDefecation", "c",   2L,     "WASH",        "Open defecation",
  "StandingWater",  "c",   2L,     "WASH",        "Standing water",
  "Flooding",       "c",   2L,     "WASH",        "Household flooding",
  "RefuseCollected","c",   2L,     "WASH",        "Refuse collected",
  "HandwashScore",  "p",   1L,     "WASH",        "Handwashing score (0-8)",
  # --- health service use / antibiotics -------------------------------------
  "CareMonthly",    "c",   2L,     "Health",      "Care sought monthly+",
  "HIVinHH",        "c",   2L,     "Health",      "HIV in household",
  "ChronicAny",     "c",   2L,     "Health",      "Any chronic condition",
  "DiarrSeverity",  "c",   3L,     "Health",      "Diarrhoea severity",
  "AbxNoRx",        "c",   2L,     "Antibiotic",  "Antibiotics w/o Rx",
  "AbxSource",      "c",   3L,     "Antibiotic",  "Antibiotic source",
  "AbxCourse",      "c",   3L,     "Antibiotic",  "Course completion",
  # --- food / animals -------------------------------------------------------
  "StreetFood",     "c",   2L,     "Food/animal", "Eats street food",
  "MeatFreq",       "g",   1L,     "Food/animal", "Meat frequency",
  "AnimalsKept",    "c",   2L,     "Food/animal", "Keeps animals",
  "ManureUse",      "c",   2L,     "Food/animal", "Uses manure"
)

## A smaller "core" set for when you want more power per parameter. With
## n = 164 the full 32-node model is a genuinely high-dimensional problem
## (the l1 penalty keeps it identified, but sensitivity is low).
core_vars <- c("ESBL_pos_n","EC_pos_n","KP_pos_n","Months_tested","Age","Sex","HHsize",
               "Crowding","Education","IncomeBand","PipedInside","FlushToilet",
               "ToiletShared","OpenDefecation","StandingWater","HandwashScore",
               "HIVinHH","AbxNoRx","AbxSource","AbxCourse","AnimalsKept")

## ---------------------------------------------------------------------------
## 9. Missing data
## ---------------------------------------------------------------------------
## mgm() cannot take NA. Missingness here is light (<= 6 cells per variable),
## so median (continuous / count) or modal (categorical) single imputation is
## used by default. Set IMPUTE <- FALSE for complete-case analysis instead.
IMPUTE <- TRUE

miss_report <- sapply(registry$var, function(v) sum(is.na(D[[v]])))
cat("\nMissing cells per model variable (STRICT_MISSING =", STRICT_MISSING, "):\n")
if (any(miss_report > 0)) print(miss_report[miss_report > 0]) else cat("  none\n")
cat(sprintf("Total missing cells: %d across %d variables (%.2f%% of the matrix)\n",
            sum(miss_report), sum(miss_report > 0),
            100 * sum(miss_report) / (nrow(D) * nrow(registry))))

if (IMPUTE) {
  for (i in seq_len(nrow(registry))) {
    v <- registry$var[i]
    x <- D[[v]]
    if (!any(is.na(x))) next
    fill <- if (registry$type[i] == "c") {
      as.numeric(names(sort(table(x), decreasing = TRUE))[1])   # mode
    } else {
      median(x, na.rm = TRUE)                                    # median
    }
    D[[v]][is.na(x)] <- fill
  }
} else {
  D <- D[complete.cases(D[, registry$var]), ]
}
cat("Rows retained:", nrow(D), "\n")

## ---------------------------------------------------------------------------
## 10. Validate the coding against mgm's requirements
## ---------------------------------------------------------------------------
## Any violation caught here would otherwise surface as an obscure glmnet error
## deep inside mgm(), so we check explicitly:
##   - "p" variables are non-negative integers
##   - "c" variables have exactly `level` distinct consecutive values,
##     starting at 0 (binary, so binarySign works) or 1 (m > 2)
for (i in seq_len(nrow(registry))) {
  v <- registry$var[i]; ty <- registry$type[i]; lv <- registry$level[i]
  x <- D[[v]]
  if (ty == "p") {
    stopifnot(all(x >= 0), all(x == round(x)))
  }
  if (ty == "c") {
    u <- sort(unique(x))
    if (lv == 2) {
      if (!identical(as.numeric(u), c(0, 1)))
        stop("Binary variable ", v, " is not coded 0/1: ", paste(u, collapse = ","))
    } else {
      if (!identical(as.numeric(u), as.numeric(seq_len(lv))))
        stop("Categorical ", v, " must be coded 1..", lv, "; found: ",
             paste(u, collapse = ","))
    }
  }
}
cat("Coding validation passed for all", nrow(registry), "model variables.\n")

## ---------------------------------------------------------------------------
## 11. Assemble and save the MGM object
## ---------------------------------------------------------------------------
group_cols <- c(Carriage = "#B2182B", Demographic = "#EF8A62",
                Socioeconomic = "#FDDBC7", WASH = "#67A9CF",
                Health = "#2166AC", Antibiotic = "#7B3294",
                `Food/animal` = "#5AAE61")

AIARMS <- list(
  data         = as.matrix(D[, registry$var]),
  type         = registry$type,
  level        = registry$level,
  labels       = registry$label,
  vars         = registry$var,
  groups       = split(seq_len(nrow(registry)), registry$group),
  groups_color = unname(group_cols[names(split(seq_len(nrow(registry)),
                                               registry$group))]),
  registry     = registry,
  core_vars    = core_vars,
  clean_frame  = D
)

saveRDS(AIARMS, file.path(OUT_DIR, "AIARMS_mgm.rds"))
write.csv(D, file.path(OUT_DIR, "AIARMS_clean.csv"), row.names = FALSE)

cat("\nSaved:", file.path(OUT_DIR, "AIARMS_mgm.rds"), "\n")
cat("Model matrix:", nrow(AIARMS$data), "rows x", ncol(AIARMS$data), "nodes\n")
print(table(AIARMS$type))

## The sentence in Section 1 of the report that counts variable types has to
## match this, so print it in the report's own words rather than from memory.
n_bin <- sum(registry$type == "c" & registry$level == 2)
n_cat <- sum(registry$type == "c" & registry$level >  2)
n_cnt <- sum(registry$type == "p")
n_con <- sum(registry$type == "g")
cat(sprintf(
  "\nFOR THE REPORT: %d variables -- %d binary indicators, %d categorical,\n%d counts and %d continuous measurements.\n",
  nrow(registry), n_bin, n_cat, n_cnt, n_con))
cat("(Of the", n_con, "Gaussian nodes only Age and Crowding are genuinely\n",
    "continuous; Months_tested, IncomeBand and MeatFreq are bounded ordinal\n",
    "scores declared 'g'. Say so in the report rather than calling them\n",
    "continuous measurements.)\n")

## ---------------------------------------------------------------------------
## 12. Tidy up two helpers that shadow base R
## ---------------------------------------------------------------------------
## col() above shadows base::col() (the column indices of a matrix) and norm()
## shadows base::norm(). Inside this script that is harmless -- nothing here
## wants the base versions. It stops being harmless the moment this section
## shares a global environment with anything else: the spatial analysis
## unpacks a matrix with base::col(), which would reach the questionnaire
## helper instead and fail with
##     No column starting with: 0.14146335481534...
## because the matrix gets pasted into a regular expression. Both helpers have
## done their job by this point, so they do not stay in the workspace.
rm(col, norm)

```

---

# The spatial block {#spatial-block}

`01b_add_spatial.R`. Joins the coordinates back on, projects them to local
metres, drops the two geocoding outliers, derives the spatial covariates and
extends the registry. This is where the fieldwork site codes become nodes.

```{r Spatial component, purl=FALSE}
###############################################################################
# 01b_add_spatial.R
#
# Adds a SPATIAL block to the cleaned AIARMS object without touching
# 01_clean_AIARMS.R. Reads output/AIARMS_mgm.rds, joins coordinates back from
# the raw export, derives spatial covariates, and writes
# output/AIARMS_mgm_spatial.rds with the same structure plus the new nodes.
#
# Run AFTER 01_clean_AIARMS.R:
#     source("01_clean_AIARMS.R")
#     source("01b_add_spatial.R")
#
# ---------------------------------------------------------------------------
# WHAT CHANGED IN THIS VERSION
# ---------------------------------------------------------------------------
# The spatial grouping is now the FIELDWORK SITE CODE itself -- ARUE, ARUO,
# ARUT, ARUL, ARUS, ARUU, ARUF -- rather than an anonymous merged "Zone".
#
#   * Node renamed Zone -> Site, and its categories keep their real codes, so
#     the Area Profile and showInteraction() output name actual sites instead
#     of the integers 1..4.
#   * SITE_MIN_N is 0: no site is merged, so ARUF (n = 3) keeps its own
#     level and all seven codes survive.
#   * SITE_ENCODING is "dummy_full": SEVEN binary nodes, one per site code,
#     labelled "Site ARUE" ... "Site ARUU". Every site appears in the network
#     as itself, with its own edges and its own sign. This is the encoding
#     that answers "which covariates attach to ARUU?" directly.
#     The cost is real and is handled, not ignored: the seven dummies sum to
#     1 in every row, so the SITE-TO-SITE edges among them are an artefact of
#     the coding. The app suppresses that block by default and the write-up
#     must state it. Site-to-covariate edges are what this encoding is for.
#     "categorical" (one node, seven levels) is the identified alternative
#     and stays available; "dummy_reference" drops ARUE entirely. See
#     section 0.
#   * Previously SITE_MIN_N was dropped from 20 to 10. At 20 the merge loop collapsed seven
#     sites into four (ARUF -> ARUU, then ARUS -> ARUO, then ARUU -> ARUO),
#     which is exactly the site identity you wanted to keep. At 10 only ARUF
#     merges. Set SITE_MIN_N <- 0 to keep all seven codes untouched.
#   * The trend surface (East_km / North_km) is now OFF by default. See the
#     note in section 4a: with site membership in the graph the two describe
#     overlapping structure, and including both puts near-collinear nodes in
#     the same model.
#
# THE THREE NOTIONS OF SPACE, AND WHICH ONE EACH NODE ANSWERS
#
#   (a) AREA MEMBERSHIP   -- Site        [ON by default]
#       Which fieldwork area a household belongs to. Captures discrete
#       between-settlement differences: shared standpipe, shared sewer line,
#       shared refuse round. This is the one you asked for.
#
#   (b) LOCAL CONTEXT     -- NN_density  [ON by default]
#       How many other study households sit within a fixed radius. Built
#       environment density as an exposure in its own right, and the spatial
#       variable most likely to connect to crowding and WASH.
#
#   (c) LARGE-SCALE TREND -- East_km, North_km   [OFF by default]
#       A linear trend surface. An edge from East_km means a variable shifts
#       systematically along the east-west axis, conditional on the rest.
#
#   A fourth notion -- the spatially lagged OUTCOME -- is derived in section 6
#   but deliberately left out of the node set. See the note there.
###############################################################################

## Deliberately NOT rm(list = ls()) here. This script is often run a section at
## a time in the console, and wiping the workspace mid-way is how you end up
## with "object 'REG' not found" from section 7. Set RESET <- TRUE if you do
## want a clean slate.
RESET <- FALSE
if (RESET) rm(list = setdiff(ls(), "RESET"))

OUT_DIR  <- "output"
RAW_PATH <- "full_AIARMS_df.csv"

## load_aiarms() is idempotent: safe to call again from any section.
load_aiarms <- function() {
  f <- file.path(OUT_DIR, "AIARMS_mgm.rds")
  if (!file.exists(f))
    stop("Cannot find ", f, "\n",
         "Run 01_clean_AIARMS.R first -- 01b extends its output, it does not ",
         "rebuild it.")
  readRDS(f)
}

AIARMS <- load_aiarms()
REG    <- AIARMS$registry
D      <- AIARMS$clean_frame

## ---------------------------------------------------------------------------
## 0. Switches
## ---------------------------------------------------------------------------
## How the fieldwork site enters the model. All three produce valid models;
## they answer the question differently. Numbers below are from this cohort.
##
##   "categorical"     ONE node, level = m -- THE DEFAULT HERE. Every site
##                     code is a level of that node, so all seven appear and
##                     none is a reference. mgm handles it natively: the
##                     nodewise model for Site is one multinomial regression,
##                     and the m x (u-1) parameters behind each edge are
##                     aggregated into a single weight (Haslbeck & Waldorp
##                     2020, Sec. 2.2). Per-site detail is recovered with
##                     showInteraction(). One edge per partner, and no sign,
##                     because a sign is not definable for m > 2.
##                     Caveat to state in the write-up: ARUF has n = 3, so
##                     that level is estimated from three households. Read
##                     any edge that turns on ARUF as hypothesis-generating.
##
##   "dummy_reference" m-1 BINARY nodes, one site held out as reference.
##                     RECOMMENDED if you want per-site nodes: each site gets
##                     its own node, its own edges, and -- because binary nodes
##                     are coded 0/1 with binarySign = TRUE -- its own SIGN.
##                     Predicting each dummy from the others gives R^2 of
##                     0.18-0.27 here, so they are ordinary correlated
##                     predictors, nothing degenerate. Cost: the reference site
##                     has no node, and every effect is read relative to it.
##
##   "dummy_full"      ALL m binary nodes -- THE DEFAULT HERE. Every site code
##                     is a node in its own right: ARUE, ARUF, ARUL, ARUO,
##                     ARUS, ARUT, ARUU each appear in the network with their
##                     own edges and, because binary nodes are coded 0/1 with
##                     binarySign = TRUE, their own SIGN. Nothing is held out
##                     and no edge is a contrast against a reference site.
##
##                     THE COST, STATED PLAINLY. The dummies sum to 1 in every
##                     row, so predicting any one from the other m-1 gives
##                     R^2 = 1 exactly. Two consequences:
##                       (i)  each site node's own nodewise logistic regression
##                            is perfectly separated by the other six, which
##                            produces a clique of strong SITE-TO-SITE edges
##                            that is an artefact of the coding. Ignore that
##                            block; the app zeroes it by default.
##                       (ii) inside a COVARIATE's regression the seven dummies
##                            plus the intercept are rank-deficient, so the
##                            lasso picks one of several equivalent solutions.
##                            Site-to-covariate edges are therefore stable in
##                            aggregate but which individual site carries an
##                            effect can move between bootstrap samples. Check
##                            anything you intend to report against the
##                            Stability tab, and against the "categorical" fit.
##
##                     Use "categorical" when you need an identified model to
##                     defend; use this when you need to see the sites.
##
## Note: mgm's own overparameterize = TRUE is NOT this. There the indicators
## are predictors inside another node's design matrix and never receive a
## conditional model of their own, which is why the L1 penalty is enough there
## and is not enough here.
SITE_ENCODING  <- "dummy_full"    # one binary node per site; all seven kept
SITE_REFERENCE <- NULL   # NULL = use the largest site; or name one, e.g. "ARUE"

SITE_MIN_N     <- 0      # merge a site into its nearest neighbour below this n
                         #   0  -> keep all seven codes exactly as they are
                         #   10 -> only ARUF (n = 3) merges, leaving six sites
                         #   20 -> collapses to four zones (the old behaviour)
INCLUDE_TREND  <- FALSE  # add East_km / North_km alongside Site
DROP_OUTLIERS  <- TRUE
OUTLIER_KM     <- 10
RADIUS_M       <- 150    # neighbourhood radius for NN_density
K_NN           <- 8      # neighbours for the spatial weights matrix

## ---------------------------------------------------------------------------
## 1. Join coordinates back on Study_code
## ---------------------------------------------------------------------------
raw <- read.csv(RAW_PATH, stringsAsFactors = FALSE, check.names = TRUE,
                encoding = "UTF-8", na.strings = c("", "NA", "na", "N/A"))
stopifnot(nrow(raw) > 0L)                      # see the note in 01 on encoding
raw <- raw[!duplicated(raw$Study_code), ]      # same de-duplication as 01

geo <- raw[, c("Study_code", "Longitude.x", "Latitude.x")]
D   <- merge(D, geo, by = "Study_code", all.x = TRUE, sort = FALSE)
D   <- D[match(AIARMS$clean_frame$Study_code, D$Study_code), ]   # restore order
stopifnot(!any(is.na(D$Latitude.x)))

## Fieldwork site is encoded in the Study_code prefix.
D$Site_code <- sub("^([A-Za-z]+).*$", "\\1", D$Study_code)
cat("Fieldwork sites as recorded:\n"); print(table(D$Site_code))

## ---------------------------------------------------------------------------
## 2. Project longitude/latitude to local metres
## ---------------------------------------------------------------------------
## Over a study area a few km across, an equirectangular projection about the
## median point is accurate to well under a metre, so there is no need for sf
## or a formal CRS transformation. Distances are then ordinary Euclidean.
R_EARTH <- 6371000
lat0 <- median(D$Latitude.x); lon0 <- median(D$Longitude.x)
D$x_m <- R_EARTH * cos(lat0 * pi/180) * (D$Longitude.x - lon0) * pi/180
D$y_m <- R_EARTH * (D$Latitude.x - lat0) * pi/180

## ---------------------------------------------------------------------------
## 3. Outlying households
## ---------------------------------------------------------------------------
## Two ARUO households are recorded ~62 km from every other point, at identical
## coordinates. Whether that is a second site or a GPS/data-entry error, they
## would dominate any distance-based quantity: the trend surface would fit one
## line through them, and their nearest-neighbour density would be 1.
d_centre <- sqrt(D$x_m^2 + D$y_m^2) / 1000
is_out   <- d_centre > OUTLIER_KM
cat("\nHouseholds beyond", OUTLIER_KM, "km of the study centroid:", sum(is_out), "\n")
if (sum(is_out)) print(D[is_out, c("Study_code", "Longitude.x", "Latitude.x")])

if (DROP_OUTLIERS && any(is_out)) {
  D <- D[!is_out, ]
  cat("Dropped. Retained", nrow(D), "households for the spatial model.\n")
}

## ---------------------------------------------------------------------------
## 4. Spatial covariates
## ---------------------------------------------------------------------------

## (a) Trend surface -- OFF unless INCLUDE_TREND is TRUE.
##     Kilometres rather than metres so the coefficients sit on a scale
##     comparable to the other Gaussian nodes after scale = TRUE.
##
##     Why off by default now that Site is in the model: the sites ARE the
##     spatial partition of this cohort, and their centroids span barely 2 km.
##     A linear surface fitted across that span is largely a smoothed version
##     of site membership, so the two nodes compete to explain the same
##     variation. The lasso then splits the signal arbitrarily between them and
##     both edges look weaker than the structure actually is. Turn it on only
##     if you specifically want to ask whether anything varies smoothly ACROSS
##     sites rather than BETWEEN them.
D$East_km  <- (D$x_m - mean(D$x_m)) / 1000                            # "g"
D$North_km <- (D$y_m - mean(D$y_m)) / 1000                            # "g"

## (b) Local density: number of OTHER study households within RADIUS_M.
##     A count of points in a disc, so a Poisson node. Median nearest-neighbour
##     distance is about 17 m, so 150 m captures the immediate cluster.
dmat <- as.matrix(dist(cbind(D$x_m, D$y_m)))
D$NN_density <- rowSums(dmat < RADIUS_M) - 1                          # "p"

## (c) Site: the fieldwork area code, kept as the grouping.
##
##     Merging rule: while any site has fewer than SITE_MIN_N households, the
##     smallest is merged into the site with the NEAREST centroid. Merging by
##     proximity rather than by convenience keeps every resulting group
##     spatially coherent -- a merged group is still a contiguous patch of the
##     study area, not a scattering of unrelated households.
##
##     Why a floor exists at all: a categorical level is estimated from the
##     households in it. ARUF has three. Three observations cannot support the
##     parameters of their own multinomial level, and mgm will either drop the
##     level or produce an estimate driven by those three rows. Setting
##     SITE_MIN_N <- 0 keeps all seven codes if you would rather see that
##     happen than merge.
site <- D$Site_code
merge_log <- character(0)
repeat {
  tab <- table(site)
  if (all(tab >= SITE_MIN_N) || length(tab) <= 2) break
  small <- names(tab)[which.min(tab)]
  cent  <- t(sapply(split(seq_along(site), site),
                    function(i) c(mean(D$x_m[i]), mean(D$y_m[i]))))
  others <- setdiff(rownames(cent), small)
  dd <- sqrt((cent[others, 1] - cent[small, 1])^2 +
             (cent[others, 2] - cent[small, 2])^2)
  target <- others[which.min(dd)]
  merge_log <- c(merge_log,
                 sprintf("  %s (n = %d) -> %s  (centroids %.2f km apart)",
                         small, min(tab), target, min(dd) / 1000))
  site[site == small] <- target
}

site_f      <- factor(site)
site_labels <- levels(site_f)
n_site      <- nlevels(site_f)

## Build whichever node set SITE_ENCODING asked for.
site_vars <- character(0)

if (SITE_ENCODING == "categorical") {
  D$Site    <- as.integer(site_f)                                     # "c", m
  site_vars <- "Site"
  ## The integer code is what mgm needs, but nothing downstream should ever
  ## have to guess that 4 means ARUO. site_levels travels with the object and
  ## is what the spatial analysis and the plots label with.
  site_levels <- site_labels
  names(site_levels) <- seq_along(site_labels)

} else {
  ## One binary node per site, coded 0/1 so binarySign = TRUE gives each a sign.
  all_sites <- site_labels
  if (SITE_ENCODING == "dummy_reference") {
    ref <- if (is.null(SITE_REFERENCE)) names(which.max(table(site)))
           else SITE_REFERENCE
    if (!ref %in% all_sites)
      stop("SITE_REFERENCE '", ref, "' is not one of: ",
           paste(all_sites, collapse = ", "))
    kept <- setdiff(all_sites, ref)
    cat("\nReference site (absorbed, no node):", ref,
        sprintf("(n = %d)\n", sum(site == ref)))
  } else if (SITE_ENCODING == "dummy_full") {
    kept <- all_sites
    message("SITE_ENCODING = 'dummy_full': all ", length(kept), " site codes ",
            "get their own binary node.\n",
            "  The dummies sum to 1 in every row, so the SITE-TO-SITE edges ",
            "among them are an\n  artefact of the coding, not a finding. Read ",
            "site-to-covariate edges only; the app\n  suppresses the site-site ",
            "block by default. Report this in the methodology.")
  } else {
    stop("SITE_ENCODING must be 'categorical', 'dummy_reference' or 'dummy_full'.")
  }
  for (a in kept) {
    v <- paste0("Site_", a)
    D[[v]] <- as.integer(site == a)                                   # "c", 2
    site_vars <- c(site_vars, v)
  }
}
D$Site_group <- site      # merged code as text, for maps and tables

cat("\nSite grouping (SITE_MIN_N =", SITE_MIN_N, "):\n")
if (length(merge_log)) {
  cat("merges performed:\n"); cat(merge_log, sep = "\n"); cat("\n")
} else {
  cat("no merges needed; all site codes retained\n")
}
print(table(site))
cat("\nInteger code -> site:\n")
print(setNames(site_labels, seq_len(n_site)))

## (d) Distance from the centre of the study area. Correlated with the trend
##     surface by construction, so it is derived but NOT in the default set.
D$Dist_centre_km <- sqrt((D$x_m - mean(D$x_m))^2 +
                         (D$y_m - mean(D$y_m))^2) / 1000

## ---------------------------------------------------------------------------
## 5. Spatial weights matrix (k nearest neighbours, row-standardised)
## ---------------------------------------------------------------------------
## Used by the Moran's I diagnostics in 02_fit_mgm.R. Symmetrised so that
## "i is a neighbour of j" implies the reverse, then row-standardised so each
## household's neighbours carry a total weight of 1.
## CORRECTNESS NOTE. The previous version took order(dmat[i, ])[2:(K_NN+1)]
## and assumed position 1 is household i. That holds only when i is the
## unique minimiser of its own distance row. Seven pairs of households in
## this cohort sit at identical coordinates, so for 7 of the 162 rows
## order() put the OTHER household first and i itself second -- and i was
## then written into its own neighbour set. After symmetrising and row
## standardising, diag(W) was nonzero for those 7 rows, up to w_ii = 0.182.
## The report defines w_ii = 0 (Section 2.1), so this was a direct conflict
## with the written methodology. Excluding i explicitly fixes it. The effect
## on the numbers is small -- Moran's I for RefuseCollected moves from
## 0.2272 to 0.2184 -- but the matrix is now the one the report describes.
W <- matrix(0, nrow(D), nrow(D))
for (i in seq_len(nrow(D))) {
  nb <- setdiff(order(dmat[i, ]), i)[seq_len(K_NN)]
  W[i, nb] <- 1
}
W <- (W + t(W)) / 2
W <- W / rowSums(W)
stopifnot(all(diag(W) == 0), all(abs(rowSums(W) - 1) < 1e-12))

## ---------------------------------------------------------------------------
## 6. Spatially lagged carriage  (derived, but OFF by default)
## ---------------------------------------------------------------------------
## The mean carriage of a household's neighbours. As a graph node this is
## seductive but awkward, for two reasons worth being explicit about:
##
##   1. ENDOGENEITY. It is built from other rows' values of the very outcome
##      it would be joined to. In spatial econometrics this is the lagged
##      dependent variable, and it is not exogenous; the resulting edge cannot
##      be read the way the other edges are read.
##   2. INDEPENDENCE. An MGM assumes i.i.d. rows. A node constructed across
##      rows breaks that assumption directly.
##
## Moran's I on the model residuals answers the same question -- is there
## spatial structure left over? -- without either problem.
D$ESBL_lag <- as.numeric(W %*% D$ESBL_pos_n)
D$KP_lag   <- as.numeric(W %*% D$KP_pos_n)

## ---------------------------------------------------------------------------
## Prerequisite guard
## ---------------------------------------------------------------------------
## If you jumped straight here, these recover or explain what is missing rather
## than failing on an unhelpful "object not found".
if (!exists("AIARMS")) AIARMS <- load_aiarms()
if (!exists("REG"))    REG    <- AIARMS$registry
need <- c("D", "n_site", "site_labels", "site_vars", "RADIUS_M", "dmat", "W")
gone <- need[!vapply(need, exists, logical(1))]
if (length(gone))
  stop("Section 7 needs objects built earlier in this script: ",
       paste(gone, collapse = ", "),
       "\nRun the whole file instead:  source(\"01b_add_spatial.R\")")

## ---------------------------------------------------------------------------
## 7. Extend the registry
## ---------------------------------------------------------------------------
## NN_density is an ADDITION to the site nodes, not a replacement for them.
## They answer different questions: site membership is which settlement you are
## in, density is how built-up your immediate surroundings are. Set
## INCLUDE_DENSITY <- FALSE if you want the site nodes alone.
INCLUDE_DENSITY <- TRUE

site_reg <- if (SITE_ENCODING == "categorical") {
  data.frame(var = "Site", type = "c", level = as.integer(n_site),
             group = "Spatial",
             label = "Fieldwork site",
             stringsAsFactors = FALSE)
} else {
  data.frame(var = site_vars, type = "c", level = 2L, group = "Spatial",
             label = if (SITE_ENCODING == "dummy_reference")
                       sprintf("Site %s (vs %s)", sub("^Site_", "", site_vars), ref)
                     else paste("Site", sub("^Site_", "", site_vars)),
             stringsAsFactors = FALSE)
}

spatial_reg <- site_reg
if (INCLUDE_DENSITY)
  spatial_reg <- rbind(
    data.frame(var = "NN_density", type = "p", level = 1L, group = "Spatial",
               label = paste0("Neighbours within ", RADIUS_M, "m"),
               stringsAsFactors = FALSE),
    spatial_reg)
if (INCLUDE_TREND) {
  spatial_reg <- rbind(
    data.frame(var = c("East_km", "North_km"), type = "g", level = 1L,
               group = "Spatial", label = c("Easting (km)", "Northing (km)"),
               stringsAsFactors = FALSE),
    spatial_reg)
}
REG2 <- rbind(as.data.frame(REG), spatial_reg)

## Same validation as in 01: "p" non-negative integers, "c" consecutive from
## 0 (binary) or 1 (m > 2).
for (i in seq_len(nrow(spatial_reg))) {
  v <- spatial_reg$var[i]; ty <- spatial_reg$type[i]; lv <- spatial_reg$level[i]
  x <- D[[v]]
  if (anyNA(x)) stop("Spatial variable ", v, " has missing values.")
  if (ty == "p") stopifnot(all(x >= 0), all(x == round(x)))
  if (ty == "c") {
    u <- sort(unique(x))
    ok <- if (lv == 2) identical(as.numeric(u), c(0, 1)) else
                       identical(as.numeric(u), as.numeric(seq_len(lv)))
    if (!ok) stop("Categorical ", v, " badly coded: ", paste(u, collapse = ","))
  }
}
cat("\nSpatial coding validation passed.\n")

## ---------------------------------------------------------------------------
## 8. Build the model matrix, with a guard
## ---------------------------------------------------------------------------
## mgm() rejects anything that is not integer or numeric with the message
##   "Only integer and numeric values permitted."
## which does not say WHICH column is at fault. as.matrix() on a data.frame
## containing even one character or factor column silently coerces EVERY
## column to character, so a single stray column poisons the whole matrix and
## the error surfaces much later, inside mgm(). This guard names the offender
## instead, and coerces the cases that are safe to coerce.
build_model_matrix <- function(df, vars) {
  missing_vars <- setdiff(vars, names(df))
  if (length(missing_vars))
    stop("Columns missing from the frame: ", paste(missing_vars, collapse = ", "))

  cls <- vapply(df[vars], function(z) class(z)[1], character(1))
  bad <- vars[!cls %in% c("numeric", "integer", "logical", "factor")]
  if (length(bad)) {
    stop("These columns are not numeric and cannot be coerced safely:\n",
         paste(sprintf("  %s  (class %s, e.g. %s)", bad, cls[bad],
                       vapply(df[bad], function(z)
                         paste(utils::head(unique(as.character(z)), 3),
                               collapse = " / "), character(1))),
               collapse = "\n"),
         "\nRemove them from the registry, or convert them upstream.")
  }
  ## logical -> 0/1, factor -> its integer codes; both are legitimate here.
  for (v in vars) {
    if (is.logical(df[[v]])) df[[v]] <- as.integer(df[[v]])
    if (is.factor(df[[v]]))  df[[v]] <- as.integer(df[[v]])
  }
  M <- as.matrix(df[, vars, drop = FALSE])
  storage.mode(M) <- "double"
  if (anyNA(M)) {
    na_cols <- vars[colSums(is.na(M)) > 0]
    stop("NA values remain in: ", paste(na_cols, collapse = ", "),
         " -- mgm() cannot take missing data.")
  }
  M
}

Xall <- build_model_matrix(D, REG2$var)
cat("Model matrix built:", nrow(Xall), "x", ncol(Xall),
    "| storage mode:", storage.mode(Xall), "\n")

## Core set gains the spatial block, so the default fit in 02 answers the
## question this script was added for.
core_vars2 <- c(AIARMS$core_vars, spatial_reg$var)

group_cols <- c(Carriage = "#B2182B", Demographic = "#EF8A62",
                Socioeconomic = "#FDDBC7", WASH = "#67A9CF",
                Health = "#2166AC", Antibiotic = "#7B3294",
                `Food/animal` = "#5AAE61", Spatial = "#1B7837")

AIARMS_SP <- list(
  data         = Xall,
  type         = REG2$type,
  level        = REG2$level,
  labels       = REG2$label,
  vars         = REG2$var,
  groups       = split(seq_len(nrow(REG2)), REG2$group),
  groups_color = unname(group_cols[names(split(seq_len(nrow(REG2)), REG2$group))]),
  registry     = REG2,
  core_vars    = core_vars2,
  clean_frame  = D,
  ## spatial extras carried alongside for the diagnostics and the map
  coords       = as.matrix(D[, c("x_m", "y_m")]),
  lonlat       = as.matrix(D[, c("Longitude.x", "Latitude.x")]),
  W            = W,
  K_NN         = K_NN,
  radius_m     = RADIUS_M,
  site_labels  = site_labels,      # every retained site code, in level order
  site_vars    = site_vars,        # the node name(s) carrying site information
  site_encoding = SITE_ENCODING,
  site_group   = D$Site_group,     # merged code per household, as text
  site         = D$Site_code       # original, unmerged code per household
)

saveRDS(AIARMS_SP, file.path(OUT_DIR, "AIARMS_mgm_spatial.rds"))
write.csv(D, file.path(OUT_DIR, "AIARMS_clean_spatial.csv"), row.names = FALSE)

cat("\nSaved:", file.path(OUT_DIR, "AIARMS_mgm_spatial.rds"), "\n")
cat("Model matrix:", nrow(Xall), "rows x", ncol(Xall), "nodes",
    "(", nrow(spatial_reg), "spatial )\n")
print(table(AIARMS_SP$type))

## ---------------------------------------------------------------------------
## 9. Reading the Site node afterwards
## ---------------------------------------------------------------------------
## Site is one categorical node, so the network shows ONE edge weight per
## partner: the mean absolute value of the several parameters behind it, and
## no sign. The per-site parameters are still in the fit. After running
## 02_fit_mgm.R:
##
##   j <- which(AIARMS_SP$vars == "Site")
##   showInteraction(fit_mgm, int = c(j, which(AIARMS_SP$vars == "ESBL_pos_n")))
##
## and map the integer codes back with AIARMS_SP$site_labels.
if (SITE_ENCODING == "categorical") {
  cat("\nSite codes in the fitted model:",
      paste(sprintf("%d=%s", seq_along(site_labels), site_labels), collapse = "  "),
      "\n")
} else {
  cat("\nSite nodes in the fitted model:", paste(site_vars, collapse = ", "), "\n")
  cat("Each is binary 0/1, so each carries its own SIGN in the network.\n")
}
```

---

# The mixed graphical model {#mgm-fit}

`02_fit_mgm.R`. One $\ell_1$-penalised GLM per node, stitched into a graph
(Algorithm 1 of Haslbeck & Waldorp 2020), then predictability, the network
plot, Moran's I on raw variables and on model residuals, and the bootstrap
stability assessment.

```{r Mixed Graphical Models, purl=FALSE}
###############################################################################
# 02_fit_mgm.R   (spatial-enabled)
#
# Estimate, assess and visualise a Mixed Graphical Model on the cleaned AIARMS
# data, following Haslbeck & Waldorp (2020, JSS 93(8)) Section 3.1.
#
#   (1) estimate the model              -> mgm()
#   (2) compute predictions / errors    -> predict()
#   (3) visualise the network           -> qgraph()
#   (4) SPATIAL DIAGNOSTICS             -> Moran's I, raw and on residuals
#   (5) assess stability                -> resample() + plotRes()
#   (6) higher-order (k = 3) structure  -> mgm(k = 3) + FactorGraph()
#
# Set USE_SPATIAL <- FALSE to reproduce the original non-spatial analysis.
###############################################################################

## In a combined run the earlier sections have already written their .rds
## outputs, so nothing here needs a clean slate -- and rm(list = ls()) part way
## through a knit would delete objects the later sections rely on. Run
## standalone, it behaves exactly as before.
if (!isTRUE(getOption("knitr.in.progress"))) rm(list = ls())
set.seed(1)     # cross-validation and the bootstrap are both stochastic

pkgs <- c("mgm", "qgraph")
new  <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(new)) install.packages(new)
library(mgm); library(qgraph)

OUT_DIR     <- "output"
USE_SPATIAL <- TRUE     # <- requires 01b_add_spatial.R to have been run

AIARMS <- if (USE_SPATIAL) readRDS(file.path(OUT_DIR, "AIARMS_mgm_spatial.rds")) else
                           readRDS(file.path(OUT_DIR, "AIARMS_mgm.rds"))

## ---------------------------------------------------------------------------
## 0. Choose the variable set
## ---------------------------------------------------------------------------
## "core" = the main hypotheses (+ the four spatial nodes when USE_SPATIAL)
## "full" = every declared node (high-dimensional at n = 162; low sensitivity)
VAR_SET <- "core"

## Written on one line: at the R console a multi-line `if ... else` is a syntax
## error, because R evaluates the complete `if` at the end of the first line and
## then meets a stray `else`. Fine inside braces or a sourced file, but this way
## the block can also be pasted.
sel <- if (VAR_SET == "core") match(AIARMS$core_vars, AIARMS$vars) else seq_along(AIARMS$vars)

X     <- AIARMS$data[, sel, drop = FALSE]
type  <- AIARMS$type[sel]
level <- AIARMS$level[sel]
labs  <- AIARMS$labels[sel]
grp   <- AIARMS$registry$group[sel]

cat("Fitting on", nrow(X), "observations and", ncol(X), "nodes\n")
if (USE_SPATIAL) cat("Spatial nodes included:",
                     paste(AIARMS$vars[sel][grp == "Spatial"], collapse = ", "), "\n")
print(table(type))

## ---------------------------------------------------------------------------
## 1. Estimate the pairwise MGM  (Algorithm 1 of the paper)
## ---------------------------------------------------------------------------
## One l1-penalised GLM per node (neighbourhood selection, Meinshausen &
## Buhlmann 2006): each variable is regressed on all the others using the link
## belonging to ITS OWN conditional distribution -- linear for "g", log-link
## Poisson for "p", multinomial for "c". The nonzero coefficients define that
## node's neighbourhood; the p neighbourhoods are stitched into one graph.
##
##   k = 2        pairwise interactions only
##   EBIC, 0.25   the paper's default; higher gamma = sparser graph
##   ruleReg AND  an edge survives only if both nodewise regressions found it
##   threshold LW the beta-min threshold tau of Equation 8
##   binarySign   signs for binary-binary and binary-continuous edges
##   scale TRUE   z-scores Gaussian nodes so one lambda penalises them fairly

fit_mgm <- mgm(data       = X,
               type       = type,
               level      = level,
               k          = 2,
               lambdaSel  = "EBIC",
               lambdaGam  = 0.25,
               ruleReg    = "AND",
               threshold  = "LW",
               binarySign = TRUE,
               scale      = TRUE)

wadj <- fit_mgm$pairwise$wadj
dimnames(wadj) <- list(labs, labs)
cat("\nEstimated edges:", sum(wadj[upper.tri(wadj)] != 0),
    "of", choose(ncol(X), 2), "possible\n")

ut <- which(upper.tri(wadj) & wadj != 0, arr.ind = TRUE)
edges <- data.frame(from = labs[ut[, 1]], to = labs[ut[, 2]],
                    weight = wadj[ut], sign = fit_mgm$pairwise$signs[ut])
edges <- edges[order(-abs(edges$weight)), ]
print(head(edges, 15), row.names = FALSE)
write.csv(edges, file.path(OUT_DIR, "mgm_edgelist.csv"), row.names = FALSE)

## Which nodes does space actually touch? This is the question 01b was added
## to answer, so it is worth printing directly rather than hunting in the plot.
if (USE_SPATIAL) {
  sp_lab   <- labs[grp == "Spatial"]
  sp_edges <- edges[edges$from %in% sp_lab | edges$to %in% sp_lab, ]
  cat("\n--- Edges involving a spatial node ---\n")
  if (nrow(sp_edges)) print(sp_edges, row.names = FALSE) else
    cat("None. Location is conditionally independent of every other node here.\n")
}

## ---------------------------------------------------------------------------
## 2. Predictability (nodewise errors)
## ---------------------------------------------------------------------------
## An edge says a dependency exists, not how much of a node is explained.
## nCC is the normalised accuracy: gain over the intercept-only model divided
## by the maximum possible gain, so a node with a 90% majority class cannot
## look good by predicting the majority class.
pred <- predict(object   = fit_mgm,
                data     = X,
                errorCon = c("RMSE", "R2"),
                errorCat = c("CC", "nCC"))
print(pred$errors)

err_ring <- ifelse(type == "c", pred$errors[, "nCC"], pred$errors[, "R2"])
err_ring[is.na(err_ring)] <- 0

## ---------------------------------------------------------------------------
## 3. Visualise
## ---------------------------------------------------------------------------
grp_list <- split(seq_along(labs), grp)
pal <- c("Carriage"="#B2182B","Demographic"="#EF8A62","Socioeconomic"="#FDDBC7",
         "WASH"="#67A9CF","Health"="#2166AC","Antibiotic"="#7B3294",
         "Food/animal"="#5AAE61","Spatial"="#1B7837")

pdf(file.path(OUT_DIR, "mgm_network.pdf"), width = 12, height = 8)
qgraph(wadj,
       edge.color = fit_mgm$pairwise$edgecolor,
       layout     = "spring", repulsion = 1.1,
       pie        = err_ring,
       pieColor   = ifelse(type == "c", "tomato", "lightblue"),
       groups     = grp_list,
       color      = unname(pal[names(grp_list)]),
       nodeNames  = labs, labels = seq_along(labs),
       legend     = TRUE, legend.cex = 0.38,
       vsize = 4, esize = 14,
       title = paste0("AIARMS MGM (k = 2, EBIC gamma = 0.25, ", VAR_SET,
                      if (USE_SPATIAL) " + spatial)" else ")"))
dev.off()

###############################################################################
## 4. SPATIAL DIAGNOSTICS
###############################################################################
if (USE_SPATIAL) {

  W  <- AIARMS$W          # row-standardised k-nearest-neighbour weights
  xy <- AIARMS$coords

  ## Moran's I with a permutation test.
  ## I is a correlation between a variable and the average of its neighbours.
  ## I > 0 means nearby households resemble each other; I ~ 0 means location
  ## carries no information; I < 0 means neighbours differ more than chance.
  ## Its expectation under no autocorrelation is -1/(n-1), not 0, which matters
  ## when n is small. The permutation test avoids the normal approximation,
  ## which is unsafe for skewed counts like these.
  moran_test <- function(v, W, nperm = 999) {
    v <- as.numeric(v); ok <- !is.na(v)
    Wv <- W[ok, ok]; Wv <- Wv / rowSums(Wv); v <- v[ok]; n <- length(v)
    if (sd(v) == 0) return(list(I = NA, expected = NA, p = NA))
    Ifun <- function(z) (n / sum(Wv)) * as.numeric(t(z) %*% Wv %*% z) /
                        as.numeric(t(z) %*% z)
    z <- v - mean(v)
    I_obs  <- Ifun(z)
    I_perm <- replicate(nperm, Ifun(sample(z)))
    ## The previous version compared |I_perm| with |I_obs|, i.e. it tested
    ## against a null centred at ZERO. Under randomisation E[I] = -1/(n-1),
    ## not 0, and the report says so explicitly. The doubled-smaller-tail
    ## p-value below needs no centring assumption at all and is the same
    ## rule used throughout spatial_autocorrelation_MGM.Rmd, so the two
    ## documents now report the same p-value for the same statistic.
    hi <- (1 + sum(I_perm >= I_obs)) / (1 + nperm)
    lo <- (1 + sum(I_perm <= I_obs)) / (1 + nperm)
    list(I = I_obs, expected = -1/(n - 1),
         p = min(1, 2 * min(hi, lo)))
  }

  ## (a) RAW autocorrelation, variable by variable.
  cat("\n=== Moran's I, raw variables (k =", AIARMS$K_NN, "neighbours) ===\n")
  raw_I <- t(sapply(seq_along(labs), function(j) {
    r <- moran_test(X[, j], W); c(I = r$I, p = r$p)
  }))
  rownames(raw_I) <- labs
  print(round(raw_I[order(-raw_I[, "I"]), ], 3))

  ## (b) RESIDUAL autocorrelation.
  ## This is the diagnostic that matters. If a carriage variable is spatially
  ## clustered ONLY because its risk factors are clustered, then conditioning
  ## on the network should remove it and the residual I should fall towards
  ## zero. Residual I still positive => structure the covariates do not
  ## explain: unmeasured local exposure, or household-to-household
  ## transmission. Residuals are only defined for continuous and count nodes;
  ## for categorical nodes the analogue is a misclassification indicator,
  ## which is far harder to interpret, so they are skipped.
  cont <- which(type %in% c("g", "p"))
  cat("\n=== Moran's I on MGM residuals (continuous / count nodes) ===\n")
  ## Poisson nodes get PEARSON residuals. A raw count residual has variance
  ## proportional to its own mean, so Moran's I on raw Poisson residuals is a
  ## weighted statistic on heteroscedastic input and the households with the
  ## largest fitted counts dominate it. Dividing by sqrt(mu) removes that.
  res_I <- t(sapply(cont, function(j) {
    mu      <- pred$predicted[, j]
    resid_j <- if (type[j] == "p") (X[, j] - mu) / sqrt(pmax(mu, 1e-8))
               else                 X[, j] - mu
    r <- moran_test(resid_j, W)
    c(raw_I = raw_I[j, "I"], resid_I = r$I, p = r$p)
  }))
  rownames(res_I) <- labs[cont]
  print(round(res_I, 3))

  write.csv(data.frame(variable = rownames(raw_I), raw_I),
            file.path(OUT_DIR, "moran_raw.csv"), row.names = FALSE)
  write.csv(data.frame(variable = rownames(res_I), res_I),
            file.path(OUT_DIR, "moran_residual.csv"), row.names = FALSE)

  ## (c) Correlogram: how I behaves across distance bands. A single
  ## k-nearest-neighbour I can miss clustering that operates at a coarser
  ## scale, so look across bands before concluding there is none.
  bands <- c(0, 100, 200, 400, 800, 1600, 3200)
  dmat  <- as.matrix(dist(xy))
  correlogram <- function(v) {
    sapply(seq_len(length(bands) - 1), function(b) {
      Wb <- (dmat > bands[b]) & (dmat <= bands[b + 1])
      keep <- rowSums(Wb) > 0
      if (sum(keep) < 10) return(NA_real_)
      Wb <- Wb[keep, keep]; Wb <- Wb / rowSums(Wb)
      z <- v[keep] - mean(v[keep]); n <- length(z)
      (n / sum(Wb)) * as.numeric(t(z) %*% Wb %*% z) / as.numeric(t(z) %*% z)
    })
  }

  targets <- intersect(c("ESBL+ months", "K. pneumoniae+ months",
                         "Persons per room", "Household size"), labs)
  mids <- (bands[-1] + bands[-length(bands)]) / 2
  pdf(file.path(OUT_DIR, "spatial_correlogram.pdf"), width = 8, height = 6)
  plot(NA, xlim = range(mids), ylim = c(-0.4, 0.8), log = "x",
       xlab = "Distance band (m, log scale)", ylab = "Moran's I",
       main = "Spatial correlogram")
  abline(h = 0, lty = 2, col = "grey50")
  for (i in seq_along(targets)) {
    lines(mids, correlogram(X[, match(targets[i], labs)]),
          col = i, lwd = 2, type = "b", pch = 16)
  }
  legend("topright", legend = targets, col = seq_along(targets),
         lwd = 2, bty = "n", cex = 0.8)
  dev.off()

  ## (d) Map of carriage: colour by ESBL months, point size by local density.
  esbl_j <- match("ESBL+ months", labs)
  dens_j <- grep("^Neighbours within", labs)[1]
  pdf(file.path(OUT_DIR, "spatial_map.pdf"), width = 10, height = 8)
  v     <- X[, esbl_j]
  pal_v <- colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(10)
  cols  <- pal_v[cut(v, breaks = 10, labels = FALSE)]
  cx    <- if (!is.na(dens_j)) 0.7 + 1.6 * X[, dens_j] / max(X[, dens_j]) else 1.2
  plot(xy[, 1], xy[, 2], asp = 1, pch = 21, bg = cols, col = "grey30", cex = cx,
       xlab = "Easting (m from centroid)", ylab = "Northing (m from centroid)",
       main = "ESBL-positive months by household\n(point size = local density)")
  legend("topleft", legend = c("low", "", "", "", "high"),
         pt.bg = pal_v[c(1, 3, 5, 7, 10)], pch = 21, bty = "n",
         title = "ESBL+ months", cex = 0.8)
  dev.off()

  cat("\nWrote spatial_map.pdf, spatial_correlogram.pdf and the Moran tables.\n")
}

## ---------------------------------------------------------------------------
## 5. Stability via the nonparametric bootstrap
## ---------------------------------------------------------------------------
## Proportion of B bootstrap fits in which each edge was nonzero. Read this as
## STABILITY, not significance: l1 sampling distributions have a point mass at
## zero, so the quantiles are not confidence intervals.
RUN_BOOTSTRAP <- TRUE
if (RUN_BOOTSTRAP) {
  res_obj <- resample(object = fit_mgm, data = X, nB = 100,
                      quantiles = c(0.05, 0.95))
  pdf(file.path(OUT_DIR, "mgm_stability.pdf"), width = 9, height = 11)
  plotRes(res_obj, quantiles = c(0.05, 0.95), labels = labs)
  dev.off()
  saveRDS(res_obj, file.path(OUT_DIR, "mgm_resample.rds"))
}

## ---------------------------------------------------------------------------
## 6. Higher-order (k = 3) MGM and factor graph
## ---------------------------------------------------------------------------
## A three-way term reads as MODERATION. With space in the model the natural
## question becomes whether a covariate-carriage association differs by zone --
## but at n = 162 treat anything found here as hypothesis-generating.
RUN_K3 <- FALSE
if (RUN_K3) {
  fit_k3 <- mgm(data = X, type = type, level = level,
                k = 3, lambdaSel = "EBIC", lambdaGam = 0.25,
                overparameterize = TRUE)
  print(fit_k3$interactions$indicator)
  pdf(file.path(OUT_DIR, "mgm_factorgraph.pdf"), width = 11, height = 9)
  FactorGraph(object = fit_k3, labels = labs, PairwiseAsEdge = FALSE)
  dev.off()
  saveRDS(fit_k3, file.path(OUT_DIR, "fit_mgm_k3.rds"))
}

## ---------------------------------------------------------------------------
## 7. Save
## ---------------------------------------------------------------------------
saveRDS(list(fit = fit_mgm, X = X, type = type, level = level,
             labels = labs, groups = grp, errors = pred$errors,
             spatial = USE_SPATIAL),
        file.path(OUT_DIR, "fit_mgm_k2.rds"))

cat("\nWrote outputs to ./", OUT_DIR, "\n", sep = "")

```

---

## Why the spatial analysis consumes the cleaned object

The earlier version of this analysis cleaned the raw export itself. It no longer
does. It now consumes **`output/AIARMS_mgm_spatial.rds`**, the object produced by
`01_clean_AIARMS.R` and `01b_add_spatial.R` in `MGM_Research.Rmd`, so the MGM and
the spatial analysis are computed on the same 162 rows, the same recodes, and
the same variable definitions. There is one cleaning script in the project, not
two.

That object carries a **registry** (`var`, `type`, `level`, `group`, `label`),
which is a genuine advantage here rather than just convenient bookkeeping.
Declaring that `Education` is `"c"` with `level = 3` is a statement that the
integer codes 1, 2, 3 are *unordered* --- and that statement changes which
spatial statistic is admissible for it (Section \@ref(expand)). The registry
drives the analysis rather than being carried alongside it.

Three consequences of using their variable set instead of mine, all
improvements:

* **Handwashing is now a 0--8 score**, not eight separate binaries. In the previous version two handwashing indicators had to be dropped for near-zero variance (160/161 and 153/156 positive). As a Poisson node the information survives and is testable.
* **Site membership and neighbourhood density are in the frame.** They are spatial by construction, so they are excluded from the substantive results --- but they make excellent positive controls (Section \@ref(controls)).
* **Carriage is a count over a varying number of cultured months.** `Months_tested` is in the registry, so the confounding can be tested directly rather than assumed away (Section \@ref(carriage)).

```{r libs}
## Only these two are needed. Every figure is base graphics, so the tangled
## core -- and therefore the Shiny app -- carries no plotting dependency.
suppressPackageStartupMessages({
  library(sf)      # only for the projection cross-check in Section 3
  library(spdep)   # neighbours, weights, reference implementations
})
```

```{r paths, purl=FALSE}
MGM_RDS <- "output/AIARMS_mgm_spatial.rds"
CC_RDS  <- "output_cc/AIARMS_mgm_spatial.rds"   # complete-case build, Section 12
K_DEF   <- 8   # matches K_NN in 01b_add_spatial.R
```

```{r load, purl=FALSE}
stopifnot(file.exists(MGM_RDS))
A <- readRDS(MGM_RDS)
cat(sprintf("%d households x %d nodes\n", nrow(A$data), ncol(A$data)))
print(table(type = A$type))
knitr::kable(A$registry, caption = "The registry, as built by 01/01b")
```

---

## Registry-driven variable expansion {#expand}

Moran's I is a covariance-like quantity: it needs a variable whose numeric
values mean something. For a **binary** or **numeric** node that is fine. For a
node the registry declares as `"c"` with `level > 2`, it is not. `WorkStatus`
is coded 1 = employed, 2 = unemployed, 3 = pensioner/other, and mgm is told to
treat those as unordered precisely because the numbering is arbitrary. Running
Moran's I on that code would silently assert that "unemployed" sits midway
between "employed" and "pensioner", and the answer would change if the levels
were renumbered.

So every `"c"` node with `level > 2` is expanded into its `level` indicator
variables, and each indicator is tested on its own. The question becomes the
one that survives renumbering: *do households with tertiary education cluster?*
rather than *is the education code spatially autocorrelated?*

Three of the five expanded nodes --- `Education`, `DiarrSeverity`, `AbxCourse`
--- happen to be substantively ordered even though mgm treats them as nominal.
The indicator decomposition is still the right default, because it is the
registry that governs; Section \@ref(joincounts) adds the whole-variable test
for all five.

```{r expand-core}
## Their equirectangular x_m / y_m are already in metres, so coordinates are
## taken straight from the object rather than re-projected.
mgm_coords <- function(A) A$coords

## Site is a categorical node whose integer codes stand for ARUE, ARUF, ARUL,
## ARUO, ARUS, ARUT, ARUU. level_name() turns the code back into the code the
## fieldwork used, so nothing downstream is ever labelled "Site = 4".
level_name <- function(A, v, l) {
  if (identical(v, "Site") && !is.null(A$site_labels) &&
      l >= 1 && l <= length(A$site_labels)) A$site_labels[l] else as.character(l)
}

expand_registry <- function(A) {
  R <- A$registry; D <- A$clean_frame
  out <- list(); meta <- list()
  for (i in seq_len(nrow(R))) {
    v <- R$var[i]; ty <- R$type[i]; lv <- R$level[i]
    if (ty == "c" && lv > 2) {
      for (l in sort(unique(D[[v]]))) {
        ln <- level_name(A, v, l)
        nm <- sprintf("%s_%s", v, ln)
        out[[nm]]  <- as.numeric(D[[v]] == l)
        meta[[nm]] <- data.frame(var = nm, parent = v, kind = "indicator",
                                 group = R$group[i],
                                 label = sprintf("%s = %s", R$label[i], ln),
                                 stringsAsFactors = FALSE)
      }
    } else {
      out[[v]]  <- as.numeric(D[[v]])
      meta[[v]] <- data.frame(var = v, parent = NA_character_,
                              kind = if (ty == "c") "binary" else ty,
                              group = R$group[i], label = R$label[i],
                              stringsAsFactors = FALSE)
    }
  }
  list(X = as.data.frame(out), meta = do.call(rbind, meta))
}
```

```{r expand-run, purl=FALSE}
E <- expand_registry(A)
cat(sprintf("%d registry nodes -> %d analysis variables\n",
            nrow(A$registry), ncol(E$X)))
print(table(kind = E$meta$kind))
knitr::kable(subset(E$meta, !is.na(parent))[, c("var", "parent", "label")],
             row.names = FALSE, caption = "Nodes expanded into level indicators")
```

---

## Coordinates and weights

`01b_add_spatial.R` uses an equirectangular projection about the median point
rather than a formal CRS, on the argument that over a study area a couple of
kilometres across the error is negligible. That is worth checking rather than
accepting, because every distance downstream depends on it.

```{r projcheck, purl=FALSE}
ll  <- setNames(as.data.frame(A$lonlat), c("lon", "lat"))
utm <- sf::st_coordinates(
  sf::st_transform(sf::st_as_sf(ll, coords = c("lon","lat"), crs = 4326), 32736))
d_eq  <- as.matrix(dist(mgm_coords(A)))
d_utm <- as.matrix(dist(utm))
cat(sprintf("max absolute difference: %.2f m over a maximum separation of %.0f m\n",
            max(abs(d_eq - d_utm)), max(d_utm)))
cat(sprintf("max relative difference: %.3f %%\n",
            100 * max(abs(d_eq - d_utm)[d_utm > 0] / d_utm[d_utm > 0])))
```

The claim holds: the largest disagreement is 4.93 m on a maximum separation of
2286 m, i.e. 0.27% in the worst case and far less typically. The equirectangular
coordinates are used throughout so that these results and the Moran diagnostics
inside `02_fit_mgm.R` sit on identical geometry. (Note for the write-up: the
extent of the retained cohort is about 2.3 km, not 2.9 km. The 2.9 km figure in
an earlier draft predates the outlier drop.)

**What the report has to say about $\mathbf{W}$.** Equation (1) of the report
defines $w_{ij}$ as a binary critical-distance rule with $d$ chosen at the first
peak of a Moran's I correlogram. That is not what runs here, and it is not what
should run: on this cohort Moran's I decays monotonically with distance, so a
"first peak" lands at 50-75 m and leaves 14-28 of the 162 households with no
neighbours at all. What runs, in `01b` and here, is $k = 8$ nearest neighbours,
row-standardised, with $w_{ii} = 0$. Section 2.1 of the report is written to
match that, and the distance-band matrices appear in Section
\@ref(sensitivity) as the sensitivity check they are.

`01b` builds the same neighbour sets but symmetrises by averaging the indicator
matrix with its transpose rather than by union. The two give the same neighbour
*set* (1662 nonzero cells either way); averaging only halves the weight on a
one-directional link before standardising, so the matrices differ slightly in
their entries. Both are defensible; the difference is quantified in Section
\@ref(sensitivity) and the app lets either be used. `k = 8` is carried over as
the default so nothing silently diverges from `01b`.

```{r weights-core}
AIARMS_CRS <- 32736   # only used if lon/lat are supplied instead of metric coords

as_points <- function(d, crs = AIARMS_CRS) {
  p <- sf::st_as_sf(d, coords = c("lon", "lat"), crs = 4326, remove = FALSE)
  sf::st_transform(p, crs)
}

## Accepts a coordinate matrix in metres, or a frame with lon/lat columns.
coords_of <- function(d, coords = NULL) {
  if (!is.null(coords)) return(as.matrix(coords))
  sf::st_coordinates(as_points(d))
}

make_listw <- function(coords, type = c("knn", "dist", "idw"),
                       k = 6, d = 400, alpha = 1, style = "W", d0 = 1) {
  type <- match.arg(type)
  n <- nrow(coords)
  if (type == "knn") {
    k  <- min(k, n - 1L)
    nb <- spdep::knn2nb(spdep::knearneigh(coords, k = k), sym = FALSE)
    lw <- spdep::nb2listw(nb, style = style, zero.policy = TRUE)
  } else if (type == "dist") {
    nb <- spdep::dnearneigh(coords, 0, d)
    lw <- spdep::nb2listw(nb, style = style, zero.policy = TRUE)
  } else {
    nb    <- spdep::dnearneigh(coords, 0, d)
    dl    <- spdep::nbdists(nb, coords)
    ## d0 is a distance floor: some households share exact coordinates and
    ## 1/0 would be Inf.
    glist <- lapply(dl, function(x) 1 / (pmax(x, d0)^alpha))
    lw    <- spdep::nb2listw(nb, glist = glist, style = style, zero.policy = TRUE)
  }
  attr(lw, "n_isolated") <- sum(spdep::card(lw$neighbours) == 0L)
  lw
}
```

---

## The estimators, written in the report's notation

The three statistics below are coded directly from equations (2), (3) and (4)
of Section 2.1 of the research report, in the report's own notation, so that
what runs and what is written down are the same object. (Equation (1) in the
report is the definition of $w_{ij}$, so Moran's I is (2), Geary's C is (3)
and the local Moran is (4). Earlier drafts of this file referred to them as
(1)-(3); the numbering below is the one in the submitted report.) Let $Z(\mathbf{s}_i)$
be the attribute at location $\mathbf{s}_i$, let $\bar{Z}$ be its mean, let
$w_{ij}$ be the spatial weight linking $i$ and $j$, let
$w_{\cdot\cdot} = \sum_i\sum_j w_{ij}$, and let
$S^2 = \frac{1}{n-1}\sum_{i=1}^n (Z(\mathbf{s}_i) - \bar{Z})^2$.

**Moran's I**, report equation (2):

$$I = \frac{n}{(n-1)S^2 w_{\cdot\cdot}}
      \sum_{i=1}^n\sum_{j=1}^n w_{ij}
      \bigl(Z(\mathbf{s}_i)-\bar{Z}\bigr)\bigl(Z(\mathbf{s}_j)-\bar{Z}\bigr)$$

Under the null of no spatial autocorrelation $E[I] = -\frac{1}{n-1}$, so that
is the reference line on every plot below, not zero. $I > -\frac{1}{n-1}$ means
a location tends to be connected to locations with similar values;
$I < -\frac{1}{n-1}$ means connected locations tend to hold dissimilar values.

**Geary's C**, report equation (3):

$$C = \frac{1}{2S^2 w_{\cdot\cdot}}
      \sum_{i=1}^n\sum_{j=1}^n w_{ij}
      \bigl(Z(\mathbf{s}_i)-Z(\mathbf{s}_j)\bigr)^2$$

$C > 1$ means locations are connected to locations with dissimilar values;
$C < 1$ means connected locations hold similar values. Both are reported
together for the reason the report gives: $C$ is the more sensitive of the two
to differences between immediate neighbours while $I$ responds to the broader
pattern, so agreement between them is stronger evidence than either alone.

**Local Moran (LISA)**, report equation (4):

$$I_i = \frac{n\,(Z_i-\bar{Z})\sum_j w_{ij}(Z_j-\bar{Z})}
              {\sum_i (Z_i-\bar{Z})^2}$$

This is the scaling that gives Anselin's defining LISA property --- the sum
over all observations is proportional to the global measure. Here it holds
exactly: $\frac{1}{n}\sum_i I_i = I$ under row standardisation, and that is
one of the checks in Section \@ref(verify).

Two notes on how each is coded. Since $(n-1)S^2 = \sum_i(Z_i-\bar{Z})^2$,
$I$ and $C$ are evaluated with that substitution, which is the same quantity
written without forming $S^2$ separately. And the double sum in $C$ would be an
$O(n^2)$ loop, so it is expanded as

$$\sum_i\sum_j w_{ij}(Z_i-Z_j)^2
 = \sum_i\Bigl[Z_i^2\textstyle\sum_j w_{ij} - 2Z_i\sum_j w_{ij}Z_j
   + \sum_j w_{ij}Z_j^2\Bigr]$$

which is three spatial lag operations. Both rearrangements are algebraic
identities, and Section \@ref(verify) checks the coded values against
`spdep`'s independent implementations rather than taking that on trust.

```{r estimators-core}
## Report equation (2). n / [(n-1) S^2 w..] * sum_i sum_j w_ij (Z_i - Zbar)(Z_j - Zbar),
## evaluated using (n-1) S^2 = sum_i (Z_i - Zbar)^2.
moran_I <- function(Z, lw) {
  n     <- length(Z)
  Zc    <- Z - mean(Z)                                   # Z_i - Zbar
  w_dd  <- spdep::Szero(lw)                              # w..
  lagZc <- spdep::lag.listw(lw, Zc, zero.policy = TRUE)  # sum_j w_ij (Z_j - Zbar)
  (n / w_dd) * sum(Zc * lagZc) / sum(Zc^2)
}

## Report equation (3). 1 / [2 S^2 w..] * sum_i sum_j w_ij (Z_i - Z_j)^2,
## with the double sum expanded into three lag operations:
##   sum_i [ Z_i^2 sum_j w_ij  -  2 Z_i sum_j w_ij Z_j  +  sum_j w_ij Z_j^2 ]
geary_C <- function(Z, lw) {
  n     <- length(Z)
  Zc    <- Z - mean(Z)
  w_dd  <- spdep::Szero(lw)
  w_i   <- spdep::lag.listw(lw, rep(1, n), zero.policy = TRUE)  # sum_j w_ij
  lagZ  <- spdep::lag.listw(lw, Z,         zero.policy = TRUE)  # sum_j w_ij Z_j
  lagZ2 <- spdep::lag.listw(lw, Z^2,       zero.policy = TRUE)  # sum_j w_ij Z_j^2
  ((n - 1) * sum(w_i * Z^2 - 2 * Z * lagZ + lagZ2)) / (2 * w_dd * sum(Zc^2))
}

## Report equation (4). I_i = n (Z_i - Zbar) sum_j w_ij (Z_j - Zbar) / sum_i (Z_i - Zbar)^2.
## Quadrants classify each site by the signs of (Z_i - Zbar) and its spatial lag:
## High-High and Low-Low are positive local association (clusters), High-Low and
## Low-High are negative local association (the checkerboard case).
local_moran <- function(Z, lw) {
  n     <- length(Z)
  Zc    <- Z - mean(Z)
  lagZc <- spdep::lag.listw(lw, Zc, zero.policy = TRUE)
  Ii    <- n * Zc * lagZc / sum(Zc^2)
  quad  <- ifelse(Zc >= 0 & lagZc >= 0, "High-High",
           ifelse(Zc <  0 & lagZc <  0, "Low-Low",
           ifelse(Zc >= 0 & lagZc <  0, "High-Low", "Low-High")))
  data.frame(Zc = Zc, lag_Zc = lagZc, Ii = Ii,
             quadrant = quad, stringsAsFactors = FALSE)
}
```

---

## Inference by permutation

Most variables here are binary indicators or small counts, so the normal
approximation behind the randomisation variance of Moran's I is not
trustworthy. Everything is tested by Monte Carlo permutation, with the
conventional doubled-smaller-tail two-sided $p$-value.

```{r perm-core}
mc_pack <- function(obs, sim) {
  nsim <- length(sim)
  hi <- (1 + sum(sim >= obs)) / (1 + nsim)
  lo <- (1 + sum(sim <= obs)) / (1 + nsim)
  list(statistic = obs, sim = sim, nsim = nsim,
       p_greater = hi, p_less = lo, p_two = min(1, 2 * min(hi, lo)),
       E_sim = mean(sim), sd_sim = stats::sd(sim),
       z_sim = (obs - mean(sim)) / stats::sd(sim))
}

perm_moran <- function(x, lw, nsim = 999, seed = 1) {
  set.seed(seed)
  mc_pack(moran_I(x, lw), replicate(nsim, moran_I(sample(x), lw)))
}

```

A LISA permutation must hold $z_i$ **fixed** at location $i$ and reshuffle only
the other $n-1$ values [@anselin1995local]; a global reshuffle would move $z_i$
as well and answer a different question.

```{r localperm-core}
local_moran_perm <- function(Z, lw, nsim = 999, seed = 1) {
  set.seed(seed)
  n     <- length(Z); Zc <- Z - mean(Z); SS <- sum(Zc^2)
  nb    <- lw$neighbours; wt <- lw$weights
  lagZc <- spdep::lag.listw(lw, Zc, zero.policy = TRUE)
  Ii    <- n * Zc * lagZc / SS                       # report equation (4)
  p     <- rep(NA_real_, n)
  for (i in seq_len(n)) {
    ki <- length(nb[[i]]); if (ki == 0L || is.na(Ii[i])) next
    pool  <- Zc[-i]                                  # Z_i is held out
    ## WITHOUT replacement: the conditional null of Anselin (1995) draws k_i
    ## of the remaining n-1 values, it does not resample them. With
    ## replacement (the previous version) the draw variance is inflated by
    ## a factor (n-1)/(n-1-k_i+1) and the local p-values come out slightly
    ## conservative.
    draws <- vapply(seq_len(nsim), function(b) pool[sample.int(n - 1L, ki)],
                    numeric(ki))
    if (ki == 1L) draws <- matrix(draws, 1L, nsim)
    sim   <- n * Zc[i] * colSums(draws * wt[[i]]) / SS
    hi <- (1 + sum(sim >= Ii[i])) / (1 + nsim)
    lo <- (1 + sum(sim <= Ii[i])) / (1 + nsim)
    p[i] <- min(1, 2 * min(hi, lo))
  }
  quad <- ifelse(Zc >= 0 & lagZc >= 0, "High-High",
          ifelse(Zc <  0 & lagZc <  0, "Low-Low",
          ifelse(Zc >= 0 & lagZc <  0, "High-Low", "Low-High")))
  data.frame(Zc = Zc, lag_Zc = lagZc, Ii = Ii, p = p,
             q_BH = stats::p.adjust(p, "BH"), quadrant = quad,
             stringsAsFactors = FALSE)
}

## Getis-Ord Gi*, following Mtshawu et al. (2023), who pair I with Gi* maps.
getis_gstar <- function(x, lw, nsim = 999, seed = 1) {
  lwS <- spdep::nb2listw(spdep::include.self(lw$neighbours),
                         style = "W", zero.policy = TRUE)
  g <- spdep::localG_perm(x, lwS, nsim = nsim, iseed = seed)
  data.frame(Gstar = as.numeric(g), p = attr(g, "internals")[, "Pr(z != E(Gi))"])
}

## Whole-variable test for an UNORDERED categorical: the multi-colour join
## count (Cliff & Ord), permuted. Moran's I on the integer code is not a
## coherent alternative; this is.
## NOTE ON WEIGHTS. A join count counts joins, so it is defined on BINARY
## weights. Passing a row-standardised listw runs without error but returns a
## weighted sum, not a count -- the "same_colour" column then reads 10.24
## rather than 81.96 for the same configuration. The p-values are identical
## (the permutation ranks are unchanged by the row scaling), but the estimate
## column is not a count and must not be reported as one. binary_listw()
## rebuilds the same neighbour graph with style = "B".
binary_listw <- function(lw)
  spdep::nb2listw(spdep::make.sym.nb(lw$neighbours), style = "B",
                  zero.policy = TRUE)

joincount_cat <- function(x, lw, nsim = 999, seed = 1) {
  set.seed(seed)
  f  <- factor(x)
  lw <- binary_listw(lw)
  jc <- spdep::joincount.mc(f, lw, nsim = nsim, zero.policy = TRUE)
  ## BUG FIX. joincount.mc() returns an htest/mc.sim object in which
  ##     $statistic   is the OBSERVED same-colour join count
  ##     $estimate[1] is "mean of simulation"
  ##     $estimate[2] is "variance of simulation"
  ## The previous version read $estimate[1] into the "same_colour" column and
  ## $estimate[2] into "expected", so the table printed the permutation mean
  ## as the observed count and the permutation VARIANCE as the expected
  ## count. For Education level 1 that is 98.4 and 55.8 where the true values
  ## are 96 observed against 98.4 expected -- i.e. the old table showed a
  ## level with FEWER same-colour joins than chance as though it had more.
  ## The p-values were never affected. Cross-checked against
  ## joincount.test(): observed 96.00, expectation 98.13, variance 57.18.
  ##
  ## Note also that joincount.mc is one-sided ("greater"), which is the
  ## convention for a join count: clustering means MORE same-colour joins
  ## than chance. Say so in the write-up; it is not the two-sided rule used
  ## for Moran's I elsewhere in this document.
  do.call(rbind, lapply(seq_along(jc), function(i)
    data.frame(level       = levels(f)[i],
               same_colour = unname(jc[[i]]$statistic),
               expected    = unname(jc[[i]]$estimate[1]),
               sd_sim      = sqrt(unname(jc[[i]]$estimate[2])),
               p_greater   = jc[[i]]$p.value, stringsAsFactors = FALSE)))
}
```

## A dense-matrix path for the permutation loops

@amgalan2022fast make the point that spatial autocorrelation is a quadratic
form and should be evaluated as one. With $n = 162$ the weights matrix is tiny,
so all $B$ permutations are done as a single matrix product. Algebraically
identical to the loops above (checked in Section \@ref(verify)), roughly two
orders of magnitude faster, and that is what makes 20,000 permutations across
861 variable pairs a 30-second computation rather than a six-hour one.

```{r fastcore}
listw_to_W <- function(lw) as.matrix(spdep::listw2mat(lw))

moran_I_W <- function(x, W) {
  n <- length(x); z <- x - mean(x)
  (n / sum(W)) * sum(z * (W %*% z)) / sum(z^2)
}

perm_moran_W <- function(x, W, nsim = 999, seed = 1) {
  set.seed(seed); n <- length(x); z <- x - mean(x)
  obs <- (n / sum(W)) * sum(z * (W %*% z)) / sum(z^2)
  Z   <- matrix(z[replicate(nsim, sample.int(n))], n, nsim)
  mc_pack(obs, (n / sum(W)) * colSums(Z * (W %*% Z)) / sum(z^2))
}

bh_matrix <- function(P) {
  ## BH over the upper triangle only: p.adjust on the whole symmetric matrix
  ## would count each pair twice and treat the NA diagonal as a test.
  ut <- upper.tri(P)
  q  <- matrix(NA_real_, nrow(P), ncol(P), dimnames = dimnames(P))
  q[ut] <- stats::p.adjust(P[ut], "BH")
  q[lower.tri(q)] <- t(q)[lower.tri(q)]
  q
}
```

---

## Verification {#verify}

Nothing here is asserted; it is tested. Every check runs at knit time and the
document fails to build if any fails.

```{r verify, purl=FALSE}
ok <- function(label, a, b, tol = 1e-9) {
  pass <- isTRUE(all.equal(a, b, tolerance = tol))
  cat(sprintf("%-54s %s   max|diff| = %.2e\n", label,
              if (pass) "PASS" else "**FAIL**", max(abs(a - b))))
  stopifnot(pass)
}

xy <- mgm_coords(A)
lw <- make_listw(xy, "knn", k = K_DEF, style = "W")
W  <- listw_to_W(lw)
n  <- nrow(xy)
x  <- E$X$Crowding; y <- E$X$HandwashScore; b <- E$X$RefuseCollected

cat("-- estimator definitions vs spdep --------------------------------\n")
ok("Moran I vs spdep::moran()",
   moran_I(x, lw), spdep::moran(x, lw, n = n, S0 = spdep::Szero(lw))$I)
ok("Moran I vs spdep::moran.test()",
   moran_I(x, lw), unname(spdep::moran.test(x, lw)$estimate[1]))
ok("Geary C vs spdep::geary()",
   geary_C(x, lw), spdep::geary(x, lw, n = n, n1 = n - 1, S0 = spdep::Szero(lw))$C)
## spdep::localmoran uses the same n scaling as report equation (3); confirmed
## empirically here rather than taken from the documentation.
ok("local Moran (eq. 4) vs spdep::localmoran()",
   unname(local_moran(x, lw)$Ii), unname(spdep::localmoran(x, lw)[, "Ii"]))

cat("\n-- algebraic identities under row-standardisation ----------------\n")
ok("S0 == n", spdep::Szero(lw), n)
ok("mean(I_i) == I  (Anselin LISA property, eq. 4)",
   mean(local_moran(x, lw)$Ii), moran_I(x, lw))
zz <- x - mean(x); lzz <- spdep::lag.listw(lw, zz)
ok("I == slope of Wz on z through the origin",
   moran_I(x, lw), sum(zz * lzz) / sum(zz^2))
ok("E[I] under randomisation == -1/(n-1)",
   unname(spdep::moran.test(x, lw)$estimate[2]), -1 / (n - 1))
cat("\n-- binary covariate: Moran I recovers the BB join count ----------\n")
nb_sym <- spdep::make.sym.nb(spdep::knn2nb(spdep::knearneigh(xy, k = K_DEF)))
lwB <- spdep::nb2listw(nb_sym, style = "B")
BB  <- unname(spdep::joincount.test(factor(b, levels = c(0, 1)), lwB)[[2]]$estimate[1])
zb  <- b - mean(b); S0B <- spdep::Szero(lwB); K <- sum(spdep::card(nb_sym) * b)
ok("BB = [I S0 sum(z^2)/n + 2 xbar K - xbar^2 S0] / 2",
   (moran_I(b, lwB) * S0B * sum(zb^2) / n + 2 * mean(b) * K - mean(b)^2 * S0B) / 2,
   BB, tol = 1e-6)

cat("\n-- dense-matrix path is identical to the listw path --------------\n")
ok("Moran: listw vs dense", moran_I(x, lw), moran_I_W(x, W))
ok("permutation p identical for a shared seed",
   perm_moran(x, lw, nsim = 999, seed = 7)$p_two,
   perm_moran_W(x, W, nsim = 999, seed = 7)$p_two)

cat("\n-- the weights matrix carried in the MGM object ------------------\n")
ok("their W is row-standardised", unname(rowSums(A$W)), rep(1, n))
cat(sprintf("   Moran I(RefuseCollected):  their W = %+.4f | spdep knn k=8 = %+.4f\n",
            moran_I_W(b, A$W), moran_I(b, lw)))

```

---

## Positive controls {#controls}

`Site_*` and `NN_density` are spatial **by construction**: a site dummy is a
contiguous patch of the study area, and density is defined from the coordinates.
Testing them for spatial autocorrelation is circular, so they are excluded from
the substantive results. They are exactly what a positive control should be
though --- if the machinery could not find clustering in a variable that is
literally a map region, nothing else it reported would be believable.

```{r controls, purl=FALSE}
sp_vars <- E$meta$var[E$meta$group == "Spatial"]
ctrl <- do.call(rbind, lapply(sp_vars, function(v) {
  pm <- perm_moran_W(E$X[[v]], W, nsim = 9999, seed = 1)
  data.frame(variable = v, moran_I = pm$statistic, z = pm$z_sim,
             p = pm$p_two, geary_C = geary_C(E$X[[v]], lw))
}))
knitr::kable(ctrl, digits = c(0, 3, 1, 4, 3), row.names = FALSE,
             caption = "Positive controls: spatial by construction")
```

```{r controls-range, purl=FALSE}
cat(sprintf("Moran's I between %.2f and %.2f, Geary's C between %.2f and %.2f.\n",
            min(ctrl$moran_I), max(ctrl$moran_I),
            min(ctrl$geary_C), max(ctrl$geary_C)))
```

The estimator recovers what it should. The range is printed rather than
asserted because it depends on the site encoding: with all seven codes kept,
`Site_ARUF` (n = 3) sits near I = 0.25 while every other site is above 0.70, so
a hard-coded "0.61 to 0.91" -- correct for the older run in which ARUF was
merged away -- is wrong for this one. **A site with three households is a weak
positive control, not a finding.**

---

## Global results

```{r global-core}
global_table <- function(X, coords, vars = names(X),
                         wcfg = list(type = "knn", k = 8, style = "W"),
                         nsim = 9999, seed = 1) {
  lw <- do.call(make_listw, c(list(coords = as.matrix(coords)), wcfg))
  W  <- listw_to_W(lw)
  out <- do.call(rbind, lapply(vars, function(v) {
    pm <- perm_moran_W(X[[v]], W, nsim = nsim, seed = seed)
    data.frame(variable = v, moran_I = pm$statistic, EI = -1 / (nrow(X) - 1),
               sd_perm = pm$sd_sim, z_perm = pm$z_sim, p_perm = pm$p_two,
               geary_C = geary_C(X[[v]], lw), stringsAsFactors = FALSE)
  }))
  out$q_BH <- stats::p.adjust(out$p_perm, "BH")
  out[order(-out$moran_I), ]
}
```

```{r global-run, purl=FALSE}
cov_vars <- E$meta$var[E$meta$group != "Spatial"]
G <- global_table(E$X, xy, cov_vars, wcfg = list(type = "knn", k = K_DEF, style = "W"),
                  nsim = 9999, seed = 1)
G <- merge(G, E$meta[, c("var", "group", "label")], by.x = "variable", by.y = "var")
G <- G[order(-G$moran_I), ]
knitr::kable(G[, c("variable", "group", "moran_I", "z_perm", "p_perm", "q_BH", "geary_C")],
             digits = c(0, 0, 4, 2, 4, 4, 3), row.names = FALSE)
```

```{r global-plot, purl=FALSE, fig.height=9}
gp <- G[order(G$moran_I), ]
op <- par(mar = c(4, 11, 2, 1))
cols <- ifelse(gp$q_BH < 0.05, "#B2182B",
        ifelse(gp$p_perm < 0.05, "#EF8A62", "grey72"))
bp <- barplot(gp$moran_I, horiz = TRUE, names.arg = gp$variable, las = 1,
              col = cols, border = NA, cex.names = 0.62,
              xlim = range(gp$moran_I - 2 * gp$sd_perm, gp$moran_I + 2 * gp$sd_perm),
              xlab = "Global Moran's I")
abline(v = -1 / (nrow(E$X) - 1), lty = 2, col = "grey40")
segments(gp$moran_I - 1.96 * gp$sd_perm, bp,
         gp$moran_I + 1.96 * gp$sd_perm, bp, col = "grey30")
legend("bottomright", bty = "n", cex = 0.8, border = NA,
       fill = c("#B2182B", "#EF8A62", "grey72"),
       legend = c(sprintf("q < 0.05 (BH, %d tests)", nrow(G)),
                  "p < 0.05 only", "not significant"))
par(op)
```

The pattern is coherent and is itself a check on the method. What clusters is
**infrastructure and the built environment** --- refuse collection, dwelling
type, shared toilets, household flooding. Those are delivered or experienced
street by street, so they should cluster. What does not cluster is **individual
attributes** --- age, sex, work status, income band, care-seeking, antibiotic
behaviour, chronic conditions. There is no mechanism by which one household's
respondent age should predict a neighbour's, and Moran's I duly finds nothing.
A procedure that reported clustering of respondent age would be suspect.

One result that only the indicator expansion could produce: **`Education_3`**
(tertiary education) clusters, while `Education_1` and `Education_2` do not.
Collapsed to a single ordinal score that signal is diluted, because it lives
entirely in the top level. This is the concrete payoff of letting the registry
govern the statistic.

**But this one result is conditional on a coding decision, and the write-up has
to say so.** `01_clean_AIARMS.R` assigns level 3 to any response matching
`Diploma|Certificate|degree|BTech|Post graduate`. That sweeps in "Certificate
with &lt;Std 10/Gr.12" (n = 18) and "Diploma with &lt;Std 10/ Gr.12" (n = 7):
25 households whose own qualification label says it was obtained without
matric. Setting `EDU_STRICT_TERTIARY <- TRUE` moves those 25 to level 1, and
`Education_3` then falls from $I = 0.115$ ($p = 0.006$) to $I = 0.043$
($p = 0.17$) -- it stops being a finding. Nothing else in the table moves. Pick
the definition you can defend in the methodology, and report the other as a
sensitivity analysis; do not report the significant version alone.

`HIVinHH` also survives correction ($I = 0.112$, $q = 0.034$) and should be
named alongside the WASH variables. It complicates the tidy "infrastructure
clusters, individual attributes do not" reading, which is a reason to report it
rather than a reason to leave it out.

---

## Whole-variable tests for the nominal nodes {#joincounts}

The indicator results answer "does this level cluster?". The complementary
question --- "is this variable, as a whole, spatially arranged?" --- is the
multi-colour join count, which needs no numeric coding at all.

```{r jc-run, purl=FALSE}
nominal <- unique(na.omit(E$meta$parent))
JC <- do.call(rbind, lapply(nominal, function(v) {
  r <- joincount_cat(A$clean_frame[[v]], lw, nsim = 9999, seed = 2)
  cbind(variable = v, r)
}))
knitr::kable(JC, digits = c(0, 0, 1, 1, 2, 4), row.names = FALSE,
             caption = paste("Multi-colour join counts on binary weights,",
                             "9999 permutations, one-sided (more same-colour",
                             "joins than chance)"))
```

The join counts agree with the indicator analysis: the only nominal node with
spatial structure is education, and it sits in the tertiary level.

---

## Local structure: Moran scatterplot and LISA {#lisa}

```{r lisa-run, purl=FALSE}
VAR <- G$variable[1]
LI  <- local_moran_perm(E$X[[VAR]], lw, nsim = 9999, seed = 3)
LI$sig <- !is.na(LI$p) & LI$p < 0.05          # unadjusted; see the note below
cat("sites at p < 0.05:", sum(LI$sig),
    "| sites at q < 0.05 (BH over", nrow(LI), "local tests):",
    sum(LI$q_BH < 0.05, na.rm = TRUE), "\n")
table(quadrant = LI$quadrant, significant = LI$sig)
```

```{r lisa-plots, purl=FALSE, fig.width=10, fig.height=4.8}
pal <- c("High-High" = "#B2182B", "Low-Low" = "#2166AC",
         "High-Low"  = "#F4A582", "Low-High" = "#92C5DE")
op <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot(LI$Zc, LI$lag_Zc, pch = 21, cex = 1.1,
     bg = ifelse(LI$sig, pal[LI$quadrant], "white"), col = pal[LI$quadrant],
     xlab = paste(VAR, "(centred)"), ylab = "spatial lag",
     main = "Moran scatterplot")
abline(h = 0, v = 0, col = "grey70")
abline(a = 0, b = moran_I(E$X[[VAR]], lw), col = "#B2182B", lwd = 2)
legend("topleft", bty = "n", cex = 0.75, legend = names(pal), pch = 21,
       col = pal, pt.bg = pal)
plot(A$lonlat[, 1], A$lonlat[, 2],
     asp = 1 / cos(mean(A$lonlat[, 2]) * pi / 180), pch = 21, cex = 1.2,
     bg = ifelse(LI$sig, pal[LI$quadrant], "grey92"),
     col = ifelse(LI$sig, "grey20", "grey70"),
     xlab = "longitude", ylab = "latitude",
     main = paste("LISA cluster map:", VAR))
par(op)
```

**No local site survives Benjamini--Hochberg** across the 162 local tests, and
raising the replicate count does not change that: at `nsim = 19999` the
attainable floor ($1\times10^{-4}$) is already below the threshold BH would
require ($3.1\times10^{-4}$), so this is a real result rather than a resolution
artefact. That is the ordinary situation for a LISA, and it is why
@anselin1995local treats the multiple-comparison problem for local statistics as
open rather than settled. The map is therefore **exploratory** --- it identifies
candidate neighbourhoods, it does not confirm them. The global results carry the
confirmatory weight; those do survive correction.

A second caution, specific to the many binary indicators in this variable set:
for a binary covariate with prevalence $\pi$, a High--High site contributes
$(1-\pi)^2$ to $I_i$ and a Low--Low site $\pi^2$. The two positive quadrants are
therefore not equally easy to detect, and for an unbalanced indicator the
cluster map will look one-sided. That is a property of the statistic, not a
finding about the neighbourhoods. @hu2020uncovering is the reason the negative
quadrants stay on the map rather than being folded into a "hotspot" summary.

---

## Sensitivity {#sensitivity}

## Complete-case versus the imputed frame

`01_clean_AIARMS.R` fills missing cells with the median (numeric) or mode
(categorical), which is the right call for `mgm()` --- it cannot take `NA` at
all. It is not automatically the right call for a spatial statistic, so the
comparison has to be made rather than assumed.

The comparison needs a second build of the same pipeline with `IMPUTE <- FALSE`,
written to `output_cc/` so it sits beside the imputed one rather than replacing
it. The chunk below does that by re-running the cleaning and spatial sections
out of this document with the two switches overridden, in their own
environment so nothing in the current workspace is disturbed. It runs once ---
if `output_cc/` is already there it does nothing --- and takes about fifteen
seconds.

```{r cc-build, purl=FALSE}
## Set to FALSE if you would rather not have the second build happen
## automatically; the section below then reports that it was skipped.
BUILD_CC <- TRUE

## Defined here as well as in the paths chunk, so this section stands on its
## own when you run it by hand rather than from the top of the document.
CC_RDS <- "output_cc/AIARMS_mgm_spatial.rds"

## Pull one named chunk out of an .Rmd as plain text.
rmd_chunk_text <- function(rmd, label) {
  x      <- readLines(rmd, warn = FALSE)
  opens  <- grep("^```\\{r", x)
  fences <- grep("^```\\s*$", x)
  for (o in opens) {
    if (identical(trimws(sub("^```\\{r[ ,]*([^,}]*).*$", "\\1", x[o])), label)) {
      cl <- fences[fences > o][1]
      if (!is.na(cl) && cl > o + 1) return(paste(x[(o + 1):(cl - 1)], collapse = "\n"))
    }
  }
  NULL
}

## The cleaning sections live in this file when it is the combined document,
## and in MGM_Research.Rmd when this is the standalone spatial document, so
## both are searched.
chunk_source <- function(label) {
  here <- tryCatch(knitr::current_input(), error = function(e) NULL)
  cands <- unique(c(here, "WST795_analysis.Rmd", "MGM_Research.Rmd"))
  for (f in cands) {
    if (is.null(f) || !nzchar(f) || !file.exists(f)) next
    b <- rmd_chunk_text(f, label)
    if (!is.null(b)) return(b)
  }
  NULL
}

if (BUILD_CC && !file.exists(CC_RDS)) {
  msg <- tryCatch({
    s1 <- chunk_source("Data Cleaning")
    s2 <- chunk_source("Spatial component")
    if (is.null(s1) || is.null(s2))
      stop("could not find the 'Data Cleaning' and 'Spatial component' chunks")
    code <- paste(s1, s2, sep = "\n")
    ## The two switches, overridden in the source rather than set afterwards,
    ## because both sections assign them near the top of their own bodies.
    code <- gsub('OUT_DIR  <- "output"', 'OUT_DIR  <- "output_cc"', code, fixed = TRUE)
    code <- sub('IMPUTE <- TRUE',        'IMPUTE <- FALSE',         code, fixed = TRUE)

    ## Run it in a SEPARATE R PROCESS, which is how you would run it by hand.
    ## Two reasons it cannot simply be eval()ed here:
    ##   * the cleaning section opens with rm(list = ls()), which would wipe
    ##     the objects the imputed analysis above is still using;
    ##   * the spatial section checks its prerequisites with
    ##     vapply(need, exists, logical(1)), and exists() resolves against the
    ##     global environment rather than a private one -- so evaluating the
    ##     code into a new.env() fails with "Section 7 needs objects built
    ##     earlier in this script". Its own comment says as much.
    tmp <- file.path(tempdir(), "cc_build.R")
    writeLines(code, tmp)
    rscript <- file.path(R.home("bin"),
      if (.Platform$OS.type == "windows") "Rscript.exe" else "Rscript")
    out <- suppressWarnings(system2(rscript,
      c("-e", shQuote(sprintf('setwd("%s"); source("%s")',
                              normalizePath(getwd(), winslash = "/"),
                              normalizePath(tmp,      winslash = "/")))),
      stdout = TRUE, stderr = TRUE))
    st <- attr(out, "status")
    if (!is.null(st) && st != 0)
      stop("the build exited with status ", st, ":\n",
           paste(utils::tail(out, 8), collapse = "\n"))
    if (!file.exists(CC_RDS))
      stop("the build ran but did not write ", CC_RDS)
    Acc <- readRDS(CC_RDS)
    sprintf("built %s -- n = %d complete cases, %d nodes, %d site codes retained",
            CC_RDS, nrow(Acc$data), ncol(Acc$data), length(Acc$site_labels))
  }, error = function(e) paste("complete-case build failed:", conditionMessage(e)))
  cat(msg, "\n")
} else if (!BUILD_CC) {
  cat("BUILD_CC is FALSE -- the complete-case comparison below is skipped.\n")
} else {
  cat("using the existing", CC_RDS, "\n")
}
```

```{r cc, purl=FALSE, eval=file.exists("output_cc/AIARMS_mgm_spatial.rds")}
## Guarded in the body as well as in the chunk option, so running this chunk on
## its own prints an explanation instead of failing inside readRDS().
if (!file.exists(CC_RDS)) {
  cat("No complete-case build found at", CC_RDS,
      "\nRun the cc-build chunk above with BUILD_CC <- TRUE.\n")
} else {
Ac <- readRDS(CC_RDS); Ec <- expand_registry(Ac)
Wc <- listw_to_W(make_listw(mgm_coords(Ac), "knn", k = K_DEF, style = "W"))
vv <- intersect(cov_vars, names(Ec$X))
S <- data.frame(
  variable  = vv,
  imputed   = sapply(vv, function(v) moran_I_W(E$X[[v]],  W)),
  complete  = sapply(vv, function(v) moran_I_W(Ec$X[[v]], Wc)))
S$shift <- S$complete - S$imputed
S <- S[order(-abs(S$shift)), ]
cat(sprintf("n = %d (imputed) versus n = %d (complete case)\n",
            nrow(E$X), nrow(Ec$X)))
cat(sprintf("correlation of Moran's I across the two frames: %.4f\n",
            cor(S$imputed, S$complete)))
print(knitr::kable(head(S, 10), digits = 4, row.names = FALSE,
                   caption = "Largest shifts in Moran's I"))
}
```

### What this comparison actually shows, and it is not what an earlier draft said

The honest handling of blank survey items (`STRICT_MISSING`, Section 2) changed
this section's conclusion, and the change is worth reporting rather than
quietly absorbing.

```{r cc-cost, purl=FALSE, eval=file.exists("output_cc/AIARMS_mgm_spatial.rds")}
Aimp <- readRDS("output/AIARMS_mgm.rds")
Acc  <- readRDS("output_cc/AIARMS_mgm.rds")
Simp <- readRDS(MGM_RDS); Scc <- readRDS(CC_RDS)
cat(sprintf("listwise deletion: %d of %d households retained, %d lost (%.0f%%)\n",
            nrow(Acc$clean_frame), nrow(Aimp$clean_frame),
            nrow(Aimp$clean_frame) - nrow(Acc$clean_frame),
            100 * (1 - nrow(Acc$clean_frame) / nrow(Aimp$clean_frame))))
cat(sprintf("site codes: %d retained of %d%s\n",
            length(Scc$site_labels), length(Simp$site_labels),
            if (length(setdiff(Simp$site_labels, Scc$site_labels)))
              paste0(" -- lost entirely: ",
                     paste(setdiff(Simp$site_labels, Scc$site_labels), collapse = ", "))
            else ""))
knitr::kable(as.data.frame(rbind(
  imputed       = table(Simp$site_group)[Simp$site_labels],
  `complete case` = table(Scc$site_group)[Simp$site_labels])),
  caption = "Households per site under each frame")
```

Under the blank-as-no coding, listwise deletion cost 22 households and the two
frames agreed almost perfectly, which made it easy to report the comparison as
reassuring. Once the blanks are treated as the non-responses they are, the
picture is different: the missingness is concentrated in `ManureUse` (63
all-blank rows) and `Flooding` (32), so **listwise deletion now removes more
than half the sample**, and the correlation of Moran's $I$ across the two frames
falls to roughly 0.72.

The right conclusion is therefore **not** "the two agree, so imputation is
harmless". It is that at this level of item non-response a complete-case
analysis is **not a viable comparison at all** --- it is a different and much
smaller study, on a neighbour graph rebuilt from a third of the households, with
one fieldwork site (`ARUF`) disappearing entirely and `ARUE` falling from 46
households to 14. Differences between the two frames are dominated by **sample
composition**, not by imputation bias, and they were before as well; what has
changed is that the effect is now too large to describe as a minor shift.

That is an argument *for* single imputation in this setting, and it is a
stronger one than the old paragraph made. State it that way: `mgm()` cannot
accept `NA` at all, so some completion is mandatory, and the alternative here
would discard 57% of the sample and a whole settlement. The imputed frame is
reported as primary and the complete-case frame as a bound, with its sample size
given every time it is mentioned.

What survives that framing: the variables whose Moran's $I$ moves least between
the two frames are the ones to lead with, and the table above ranks them. Treat
the complete-case column as indicative rather than as a competing estimate.

## Carriage counts and cultured months {#carriage}

`ESBL_pos_n` and `KP_pos_n` are counts over a number of cultured months that
varies from 0 to 9. If sampling effort were itself spatially clustered ---
fieldworkers returning more often to some streets --- apparent clustering in the
counts could be an artefact of the visit schedule rather than of transmission.
That is testable, not something to caveat and move on from.

```{r carriage, purl=FALSE}
D  <- A$clean_frame
okm <- D$Months_tested > 0
Wr <- listw_to_W(make_listw(xy[okm, ], "knn", k = K_DEF, style = "W"))
car <- rbind(
  data.frame(quantity = "Months_tested (exposure)", I = moran_I_W(D$Months_tested, W),
             n = nrow(D)),
  data.frame(quantity = "ESBL_pos_n (count)", I = moran_I_W(D$ESBL_pos_n, W), n = nrow(D)),
  data.frame(quantity = "ESBL positive fraction",
             I = moran_I_W(D$ESBL_pos_n[okm] / D$Months_tested[okm], Wr), n = sum(okm)),
  data.frame(quantity = "KP_pos_n (count)", I = moran_I_W(D$KP_pos_n, W), n = nrow(D)),
  data.frame(quantity = "KP positive fraction",
             I = moran_I_W(D$KP_pos_n[okm] / D$Months_tested[okm], Wr), n = sum(okm)))
knitr::kable(car, digits = 4, row.names = FALSE)
```

`Months_tested` shows no spatial structure, so the counts are not confounded by
a spatially clustered visit schedule, and counts and fractions agree. Carriage
itself shows no spatial autocorrelation on this weights matrix at this sample
size --- which is the substantively interesting result, and the one the MGM's
`Site` nodes should be read against.

## The weights matrix

Every number above is conditional on $\mathbf{W}$, so the conditioning has to be
shown. This also quantifies the difference between `01b`'s averaged-then-
standardised construction and `spdep`'s.

```{r sens, purl=FALSE, fig.height=5.5}
cfgs <- list(`knn k=4` = list(type = "knn", k = 4),
             `knn k=6` = list(type = "knn", k = 6),
             `knn k=8` = list(type = "knn", k = 8),
             `knn k=12`= list(type = "knn", k = 12),
             `250 m`   = list(type = "dist", d = 250),
             `400 m`   = list(type = "dist", d = 400),
             `IDW 400m`= list(type = "idw", d = 400, alpha = 1))
top <- head(G$variable, 6)
S2 <- sapply(cfgs, function(cf) sapply(top, function(v)
  moran_I(E$X[[v]], do.call(make_listw, c(list(coords = xy), cf, list(style = "W"))))))
S2 <- cbind(S2, `01b W` = sapply(top, function(v) moran_I_W(E$X[[v]], A$W)))
knitr::kable(round(S2, 3), caption = "Global Moran's I under alternative weights")

op <- par(mar = c(6, 4, 1, 1))
matplot(t(S2), type = "b", pch = 16, lty = 1, xaxt = "n", col = seq_len(nrow(S2)),
        xlab = "", ylab = "Moran's I")
axis(1, seq_len(ncol(S2)), colnames(S2), las = 2)
abline(h = 0, col = "grey70")
legend("topright", rownames(S2), col = seq_len(nrow(S2)), lty = 1, pch = 16,
       bty = "n", cex = 0.75)
par(op)
```

---

## Export and tangle

```{r export, purl=FALSE}
## The Shiny app validates this object and REFUSES one it does not recognise,
## so the fields it checks -- and the schema stamp it compares against its own
## BUNDLE_SCHEMA -- have to be written here too. Keep this number and the one
## at the top of app.R in step; raise both whenever a change to the cleaning
## scripts or to the core functions alters the numbers.
BUNDLE_SCHEMA <- 3L

saveRDS(list(X = E$X, meta = E$meta, coords = xy, lonlat = A$lonlat,
             registry = A$registry, W_01b = A$W,
             cov_vars = cov_vars, sp_vars = sp_vars,
             global = G, joincounts = JC,
             site_group = A$site_group, k = K_DEF,
             site_labels = A$site_labels, site_encoding = A$site_encoding,
             schema = BUNDLE_SCHEMA,
             built = Sys.time()),
        "mgm_spatial_bundle.rds")
cat("wrote mgm_spatial_bundle.rds (schema", BUNDLE_SCHEMA, ") --",
    length(cov_vars), "covariates,", length(sp_vars), "spatial variables\n")
```

The core functions are tangled out to a plain script for the app to source.
`knitr::purl()` is deliberately not used: it only works while a knit is in
progress (`knitr::current_input()` is `NULL` when you run chunks interactively,
which surfaces as `'con' is not a connection`), and calling it during a knit
re-enters `knit()` on the same file. The tangler below reads the `.Rmd` as plain
text and behaves identically either way.

```{r tangle, purl=FALSE}
## CORE_CHUNKS is an explicit whitelist, not "everything without purl=FALSE".
## In the combined document the cleaning, spatial and model-fitting sections
## are ordinary chunks too, and tangling those into mgm_spatial_core.R would
## make the app re-run the entire pipeline every time it sources the file.
## Naming the core chunks is the only version of this rule that stays correct
## when sections are added.
CORE_CHUNKS <- c("libs", "expand-core", "weights-core", "estimators-core",
                 "perm-core", "localperm-core", "fastcore", "global-core")

write_core <- function(rmd, out = "mgm_spatial_core.R", keep = CORE_CHUNKS) {
  x      <- readLines(rmd, warn = FALSE)
  opens  <- grep("^```\\{r", x)
  fences <- grep("^```\\s*$", x)
  body   <- character(0)
  seen   <- character(0)
  for (o in opens) {
    cl <- fences[fences > o][1]
    if (is.na(cl)) stop("unclosed chunk at line ", o)
    label <- trimws(sub("^```\\{r[ ,]*([^,}]*).*$", "\\1", x[o]))
    if (!label %in% keep) next
    seen  <- c(seen, label)
    body  <- c(body, "", sprintf("## ---- %s ----", label), x[(o + 1):(cl - 1)])
  }
  missing <- setdiff(keep, seen)
  if (length(missing))
    stop("these core chunks were not found in ", basename(rmd), ": ",
         paste(missing, collapse = ", "), "\n(renamed a chunk? update CORE_CHUNKS)")
  writeLines(c(
    "## ---------------------------------------------------------------------------",
    "## mgm_spatial_core.R",
    "##",
    sprintf("## Tangled from %s on %s.", basename(rmd), format(Sys.Date())),
    "## Edit the Rmd, not this file.",
    "## ---------------------------------------------------------------------------",
    body), out)
  invisible(out)
}

rmd_file <- tryCatch(knitr::current_input(), error = function(e) NULL)
if (is.null(rmd_file) || !nzchar(rmd_file) || !file.exists(rmd_file))
  rmd_file <- if (file.exists("WST795_analysis.Rmd")) {
    "WST795_analysis.Rmd"
  } else {
    "spatial_autocorrelation_MGM.Rmd"
  }

if (file.exists(rmd_file)) {
  write_core(rmd_file)
  message("wrote mgm_spatial_core.R from ", rmd_file)
} else {
  warning("Could not find the .Rmd in ", getwd(), ", so mgm_spatial_core.R was ",
          "not regenerated. The shipped copy is current and the app will use it.")
}
```

Then, from the same directory:

```r
shiny::runApp("app.R")
```

---

## The Shiny app

```{r app-listing, eval=FALSE, purl=FALSE, code=readLines("app.R")}
```

---

## Notes for the write-up

1. Analysis runs on `AIARMS_mgm_spatial.rds`: 162 households, the same rows and recodes as the MGM. Two records 62 km from the rest were dropped by `01b` (`DROP_OUTLIERS`, `OUTLIER_KM = 10`); confirm against the field records whether that is a geocoding error or a genuine second site.
2. Nodes the registry declares as `"c"` with `level > 2` are expanded into level indicators, `Site` included --- so each of ARUE, ARUF, ARUL, ARUO, ARUS, ARUT and ARUU is tested as itself, under its own code, never as a contrast against a reference site. Moran's I on the 1..m integer codes would be an artefact of an arbitrary numbering that `mgm` explicitly treats as unordered. Whole-variable behaviour is covered by multi-colour join counts on binary weights.
3. `Site_*` and `NN_density` are excluded from the covariate results as spatial by construction, and reported separately as positive controls and as a between-settlement block. The control range is printed, not asserted; `Site_ARUF` (n = 3) sits well below the others and is a weak control.
4. Weights: 8 nearest neighbours, row-standardised, $w_{ii} = 0$, on `01b`'s equirectangular metric coordinates (verified against EPSG:32736 to within 4.93 m over a 2286 m extent). Section 2.1 of the report must describe *this*, not a critical-distance band --- a distance band chosen at the first correlogram peak leaves 14-28 households with no neighbours on these data. `01b`'s own averaged-then-standardised $\mathbf{W}$ is reported alongside.
5. Inference is by permutation throughout, using the doubled smaller-tail two-sided $p$-value, because most analysis variables are binary indicators or small counts for which the normal approximation to the randomisation variance is unreliable at $n = 162$.
6. Benjamini--Hochberg within each family. The family sizes are computed, not hard-coded: with `EC_pos_n` in the registry there are 43 univariate tests and 903 pairs (42 and 861 in the older build that excluded it). Two-sided doubled-tail $p$-values floor at $2/(B+1)$, so at `nsim = 19999` no single pair clears the rank-1 BH threshold; rejections come from ties at the floor. No LISA site survives correction at any replicate count; the local map is exploratory.
7. Complete-case deletion is **not** a viable sensitivity analysis once blank survey items are treated as non-responses: it retains 70 of 164 households, rebuilds the neighbour graph from a third of the sample, and loses the `ARUF` site entirely. Report the imputed frame as primary and the complete-case frame as a bound, always with its sample size attached; differences between them are sample composition, not imputation bias. `Education_3` separately depends on whether a sub-matric certificate counts as tertiary (see Section 8); report that sensitivity explicitly.
8. `Months_tested` shows no spatial structure, so carriage counts are not confounded by a spatially clustered visit schedule.
9. Households are the unit of analysis and every covariate is household-level, so no ecological-inference problem arises --- unlike a sewershed-level analysis, where @talleb2025socio and @hill2023using show the population-to-catchment mapping becomes the dominant error source.


---


