---
title: "Spatial modelling of wastewater data in epidemiology --- full analysis"
subtitle: "Cleaning, spatial covariates and the mixed graphical model, in one run"
author: "Marquis Pretorius --- WST795 research report"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150,
  ## cache is OFF deliberately. The first three sections write .rds files that
  ## every later section reads, so a cached chunk that did not re-run would
  ## leave the document reporting one pipeline while the files on disk came
  ## from another. The whole document runs in a few minutes; correctness is
  ## worth more than that here.
  cache = FALSE
)
options(width = 105)
```

# How to run this document

Everything from the raw export to the mixed graphical model is here, in the
order it has to happen. Knit the file and the whole analysis runs; run it chunk
by chunk and it behaves the same way. It needs two files in the same folder as
this document: `full_AIARMS_df.csv` (the data) and `app.R` (for the app
listing at the end).

| Section | Chunk | Writes |
|---|---|---|
| 2 | `Data Cleaning` | `output/AIARMS_mgm.rds` |
| 3 | `Spatial component` | `output/AIARMS_mgm_spatial.rds` |
| 4 | `Mixed Graphical Models` | the fit the app shows: edge list, predictability and the network figure |

`output/AIARMS_mgm_spatial.rds` is the object `app.R` (CLIQUE) loads. When the
app is run locally it can also rebuild that object itself, by running the
`Data Cleaning` and `Spatial component` chunks of this document, so those two
chunk labels must not be renamed.

The model reported in the research report is the one the app fits when it
opens, with the settings described in its Methodology tab: all covariates,
$k = 2$, $\lambda_s$ chosen by 10-fold cross-validation for each nodewise
regression, the OR-rule and the default threshold of the `mgm` package
(version 1.2-15). Section 4 of this document fits **that same model**: the same
data matrix, the same single categorical `Site` node, the same arguments and
the same random seed (`MGM_SEED` in `app.R`). The specification block it prints
should therefore agree line for line with the Specification panel of the app.
If a setting or the seed is changed in the app, change it in Section 4 too.

Both read `output/AIARMS_mgm_spatial.rds`, and the app does **not** rebuild that
file when one already exists. After changing anything in Sections 2 or 3, knit
this document (or delete the file) before opening the app, and copy the new
file into any deployed version of the app; otherwise the app keeps showing the
previous data.

The bootstrap in Section 4 is optional and off by default (`RUN_BOOTSTRAP`);
the app does not compute it.

## The decisions this document makes

Four switches change the numbers. They sit at the top of their own sections;
they are collected here so that none of them is made by accident.

| Switch | Section | Default | What it decides |
|---|---|---|---|
| `STRICT_MISSING` | 2 | `TRUE` | Whether a blank cell in a Yes/No grid is a "No" or a non-response. Moves `Flooding` and `ManureUse`. |
| `EDU_STRICT_TERTIARY` | 2 | `FALSE` | Whether a sub-matric certificate counts as tertiary. Moves 25 households between education levels. |
| `EXPOSURE_UNION` | 2 | `TRUE` | Whether `Months_tested` counts ESBL months only, or months with any readable result. |
| `SITE_ENCODING` | 3 | `"dummy_full"` | Seven binary site nodes, one per fieldwork code. `"categorical"` is the identified alternative. |

Set the first three to `FALSE` and `SITE_ENCODING` to `"dummy_reference"`, and
the cleaning reproduces the original pipeline's data matrix exactly. Running it
both ways is a sensitivity analysis for the cleaning decisions.

---

# Cleaning the raw export {#clean}

`01_clean_AIARMS.R`. Turns the raw household survey and the monthly culture
results into an object `mgm()` will accept: a fully numeric matrix, no missing
values, categorical variables coded as consecutive integers, and a registry
declaring each variable's type and level.

```{r Data Cleaning, purl=FALSE}
###############################################################################
# 01_clean_AIARMS.R
#
# Purpose : Turn the raw AIARMS household survey + monthly culture results into
#           a MIXED GRAPHICAL MODEL (MGM) ready object, following the data
#           requirements of Haslbeck & Waldorp (2020, JSS 93(8)), the `mgm`
#           package paper.
#
# What mgm() needs (Haslbeck & Waldorp 2020, Section 3.1):
#   1. `data`  : a fully NUMERIC matrix/data.frame, NO missing values.
#   2. `type`  : character vector, one entry per column:
#                  "g" = conditional Gaussian
#                  "p" = conditional Poisson (non-negative integer counts)
#                  "c" = conditional categorical (multinomial)
#   3. `level` : integer vector, number of categories for each variable.
#                By convention level = 1 for every "g" and "p" variable.
#   4. Categorical variables must be coded as consecutive integers.
#      - Binary variables are coded 0/1 so that `binarySign = TRUE` can define
#        a SIGN for binary-binary and binary-continuous edges.
#      - Variables with m > 2 categories are coded 1..m (no sign is definable
#        for these; the paper explains why in Section 2.2).
#
# Output: ./output/AIARMS_mgm.rds  -- a list with $data, $type, $level,
#         $labels, $groups, $groups_color, plus the untouched cleaned frame.
#
# Author: prepared for the AIARMS analysis
###############################################################################

## ---------------------------------------------------------------------------
## 0. Setup
## ---------------------------------------------------------------------------
## In a combined run the earlier sections have already written their .rds
## outputs, so nothing here needs a clean slate -- and rm(list = ls()) part way
## through a knit would delete objects the later sections rely on. Run
## standalone, it behaves exactly as before.
if (!isTRUE(getOption("knitr.in.progress"))) rm(list = ls())

pkgs <- c("dplyr", "stringr", "tidyr", "tibble")
new  <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(new)) install.packages(new)
invisible(lapply(pkgs, library, character.only = TRUE))

RAW_PATH <- "full_AIARMS_df.csv"   # <-- edit if the file lives elsewhere
OUT_DIR  <- "output"
if (!dir.exists(OUT_DIR)) dir.create(OUT_DIR)

## ---------------------------------------------------------------------------
## 0a. Cleaning switches  (each one is a decision the write-up has to defend)
## ---------------------------------------------------------------------------
## STRICT_MISSING
##   FALSE reproduces the original behaviour: a blank cell anywhere in a
##   "select all that apply" grid is read as "not ticked" and becomes 0.
##   TRUE  restricts that rule to the blocks where it is actually true of the
##   questionnaire, and treats a blank as MISSING wherever the form also
##   records an explicit "No". This matters: under FALSE, 32 blank Flooding
##   cells, 63 all-blank manure rows, 8 blank standing-water cells and 82
##   partially answered handwashing rows are silently converted to negatives
##   BEFORE the missing-data report runs, so none of them ever appear in it.
STRICT_MISSING <- TRUE

## EDU_STRICT_TERTIARY
##   FALSE reproduces the original: any response matching
##   "Diploma|Certificate|degree|BTech|Post graduate" is level 3 (tertiary).
##   That sweeps in "Certificate with <Std 10/Gr.12" (n = 18) and "Diploma
##   with <Std 10/ Gr.12" (n = 7) -- 25 households holding a sub-matric
##   certificate -- alongside Bachelor's degrees.
##   TRUE sends those 25 to level 1.
##   This moves 25 households between education levels, so whichever setting
##   you submit, report the other as a sensitivity analysis.
EDU_STRICT_TERTIARY <- FALSE

## EXPOSURE_UNION
##   FALSE reproduces the original: Months_tested counts months with an ESBL
##   result only. That denominator is wrong for 22 of 164 households when it
##   is used as the exposure for EC or KP (their columns disagree with the
##   ESBL column on which months were cultured).
##   TRUE counts a month as cultured if ANY of the three organisms has a
##   readable result.
EXPOSURE_UNION <- TRUE

## PORTABILITY. The previous version passed fileEncoding = "UTF-8", which
## tells R to RE-ENCODE the file into the session's native encoding. In a
## non-UTF-8 locale (a plain `Rscript` under LC_CTYPE=C, and the default on
## many Windows installs) that re-encoding fails on the first curly quote and
## R returns a frame with ZERO rows after only a warning:
##     "invalid input found on input connection 'full_AIARMS_df.csv'"
## The script then dies further down with an opaque rowSums() error. Passing
## encoding = "UTF-8" instead MARKS the strings as UTF-8 without re-encoding,
## which works in every locale.
##
## A second correction to the old comment: the en-dashes in the income bands
## are ALREADY mojibake in the export ("R19201 - R38400" was written by a
## double-encoding upstream). No reader argument can undo that. What matters
## is that none of the recodes below match on those characters -- the income
## bands are matched by their leading ASCII prefix, education by ASCII
## keywords -- so the mojibake is cosmetic and harmless here. Do not "fix" it
## by editing the CSV.
raw <- read.csv(RAW_PATH, stringsAsFactors = FALSE, check.names = TRUE,
                encoding = "UTF-8", na.strings = c("", "NA", "na", "N/A"))

cat("Raw dimensions:", dim(raw), "\n")
if (nrow(raw) == 0L)
  stop("Read 0 rows from ", RAW_PATH,
       " -- check the file path and that the export is not empty.")

## ---------------------------------------------------------------------------
## 1. Small helper functions
## ---------------------------------------------------------------------------

## The survey column names are enormous ("X118..D010a..Toilet.Facilities...").
## `col()` finds the single column whose name starts with a short stem, so the
## script stays readable and survives minor re-exports of the questionnaire.
col <- function(stem) {
  hit <- grep(paste0("^", stem), names(raw), value = TRUE)
  if (length(hit) == 0) stop("No column starting with: ", stem)
  raw[[hit[1]]]
}

## Yes/No -> 1/0
yn <- function(x) ifelse(x == "Yes", 1L, ifelse(x == "No", 0L, NA_integer_))

## TRUE checkbox blocks: the export writes a value only where the box was
## ticked, so a blank genuinely means "not ticked" and 0 is correct. This is
## the case for X112 (water source), where 130 of 164 cells are blank.
yn0 <- function(x) ifelse(!is.na(x) & x == "Yes", 1L, 0L)

## Yes/No ITEMS that happen to live inside a grid. Here the export writes an
## explicit "No", so a blank is a non-response, not a negative. Under
## STRICT_MISSING it becomes NA and is handled by the imputation step in
## section 9 like every other missing cell.
yn_item <- function(x) {
  if (STRICT_MISSING) ifelse(x == "Yes", 1L, ifelse(x == "No", 0L, NA_integer_))
  else                ifelse(!is.na(x) & x == "Yes", 1L, 0L)
}

## Grab every column of a checkbox block, e.g. block("X123") returns
## X123.1..Before.eating ... X123.8..When.your.hands.appear.dirty
block <- function(stem, idx) {
  nm <- paste0("^", stem, "\\.", idx, "\\.")
  do.call(cbind, lapply(nm, function(p) {
    hit <- grep(p, names(raw), value = TRUE)
    if (length(hit) == 0) rep(NA_character_, nrow(raw)) else raw[[hit[1]]]
  }))
}

## ---------------------------------------------------------------------------
## 2. De-duplicate participants
## ---------------------------------------------------------------------------
## Study_code has 164 unique values across 171 rows (repeat survey submissions).
## An MGM assumes i.i.d. rows, so duplicated participants would inflate the
## effective sample size and bias the lasso towards keeping their edges.
dup <- duplicated(raw$Study_code)
cat("Dropping", sum(dup), "duplicate Study_code rows.\n")
raw <- raw[!dup, ]

## ---------------------------------------------------------------------------
## 3. OUTCOMES: monthly stool/rectal culture results -> carriage counts
## ---------------------------------------------------------------------------
## Each participant has up to 9 monthly results per organism. Free-text entry
## produced variants: "EC_Poss", "E.coli", "No Growth", "KP_NEg",
## "ESBL_ Ecoli", "ESBL _EC + KP", ... We normalise case/underscores/spaces and
## then apply a positive / negative rule per organism.

months <- c("March","May","June","July","August","September",
            "October","November","December")

norm <- function(x) tolower(gsub("[_[:space:]]+", " ", trimws(x)))

ec_pos <- function(x) {
  s <- norm(x)
  out <- rep(NA_integer_, length(s))
  out[grepl("no growth|neg", s)]            <- 0L
  out[is.na(out) & grepl("ec|coli", s)]     <- 1L
  out
}
kp_pos <- function(x) {
  s <- norm(x)
  out <- rep(NA_integer_, length(s))
  out[grepl("neg", s)]                      <- 0L
  out[is.na(out) & grepl("kp", s)]          <- 1L
  ## cells reading only "E. coli" are E. coli results mis-filed in a KP column
  out
}
esbl_pos <- function(x) {
  s <- norm(x)
  out <- rep(NA_integer_, length(s))
  out[grepl("neg", s)]                      <- 0L
  out[is.na(out) & grepl("esbl", s)]        <- 1L
  out
}

EC <- sapply(months, function(m) ec_pos(raw[[paste0("EC_",  m)]]))
KP <- sapply(months, function(m) kp_pos(raw[[paste0("KP_",  m)]]))
ES <- sapply(months, function(m) esbl_pos(raw[[paste0("ESBL_", m)]]))

## Number of positive months (a count -> Poisson node) and number of months
## actually cultured. Because participants were sampled an unequal number of
## times, `Months_tested` is carried into the model as a Gaussian covariate:
## mgm has no offset term, so conditioning on the exposure is the closest
## available equivalent to a rate model.
## Months actually cultured. The original definition counted months with an
## ESBL result only. Checked against the export: the EC column disagrees with
## the ESBL column about which months were cultured for 71 of 164 households
## and the KP column for 74, so an ESBL-only denominator is not a valid
## exposure for EC_pos_n or KP_pos_n. The union rule below counts a month as
## cultured if any of the three organisms returned a readable result; it
## differs from the ESBL-only count for 22 households, by at most 2 months.
months_any <- (!is.na(EC)) | (!is.na(KP)) | (!is.na(ES))

D <- data.frame(
  Study_code    = raw$Study_code,
  EC_pos_n      = rowSums(EC, na.rm = TRUE),
  KP_pos_n      = rowSums(KP, na.rm = TRUE),
  ESBL_pos_n    = rowSums(ES, na.rm = TRUE),
  Months_tested = if (EXPOSURE_UNION) rowSums(months_any) else rowSums(!is.na(ES)),
  stringsAsFactors = FALSE
)

## Four households have no readable culture result in any month. Their
## carriage counts are structural zeros, not observed zeros, and they drag
## every carriage statistic towards zero. They are KEPT so that n matches the
## survey sample, but they are flagged here so the write-up can say so and so
## that a rate analysis can exclude them.
cat("Households with zero cultured months:", sum(D$Months_tested == 0), "\n")

## Convenience binaries (not used in the default model, handy for checks)
D$ESBL_ever <- as.integer(D$ESBL_pos_n > 0)
D$KP_ever   <- as.integer(D$KP_pos_n   > 0)

## NOTE: E. coli carriage is near-saturated -- 97.1% of cultured months are
## positive (952 of 981 readable EC cells) -- so most of the variance in
## EC_pos_n comes from how many months a household was cultured
## (r = 0.93 with Months_tested). It is nonetheless INCLUDED in the model
## alongside KP and ESBL so that all three target markers are represented;
## expect its main edge to be with Months_tested.

## ---------------------------------------------------------------------------
## 4. DEMOGRAPHY AND SOCIO-ECONOMIC POSITION
## ---------------------------------------------------------------------------
D$Age <- as.numeric(col("X4\\.\\."))                                   # "g"

D$Sex <- ifelse(col("X5\\.\\.") == "Female", 1L, 0L)                   # "c", 2

## X8 asks for people EXCLUDING the respondent, so add 1 for household size.
D$HHsize <- as.numeric(col("X8\\.\\.")) + 1                            # "p"

rooms_map <- c("One"=1,"Two"=2,"Three"=3,"Four"=4,"Five"=5,
               "Six"=6,"Seven"=7,"More than seven"=8)
D$Rooms <- unname(rooms_map[col("X99\\.\\.")])                         # "p"

## Crowding = persons per room. A classic transmission-relevant exposure and
## a genuinely continuous quantity, so it enters as Gaussian.
D$Crowding <- D$HHsize / D$Rooms                                       # "g"

## Highest household education, collapsed to 3 ordered-but-nominal levels.
## MGM treats "c" variables as unordered; collapsing avoids 18 sparse states.
edu <- col("X52\\.\\.")
## A qualification whose own label says "<Std 10/Gr.12" was obtained without
## matric. Under EDU_STRICT_TERTIARY it is level 1, not level 3. See the note
## on the switch: this moves 25 households and changes the Education_3
## spatial result.
tert <- grepl("Diploma|Certificate|degree|BTech|Post graduate", edu)
if (EDU_STRICT_TERTIARY) tert <- tert & !grepl("<\\s*Std 10", edu)
D$Education <- ifelse(
  tert, 3L,
  ifelse(grepl("Grade 12|Grade 13", edu) & !grepl("Attend Grade 12", edu), 2L,
         ifelse(!is.na(edu), 1L, NA_integer_)))                        # "c", 3
# 1 = primary / incomplete secondary, 2 = completed secondary, 3 = tertiary
cat("Education levels (EDU_STRICT_TERTIARY =", EDU_STRICT_TERTIARY, "):\n")
print(table(D$Education, useNA = "ifany"))

ws <- col("X53\\.\\.")
D$WorkStatus <- ifelse(ws %in% c("Employer","Paid employee","Self-employed",
                                 "Paid domestic worker"), 1L,
                ifelse(ws == "Unemployed", 2L, 3L))                    # "c", 3
# 1 = employed, 2 = unemployed, 3 = pensioner / student / other

## Income bands are ordered, so they are far more informative as a single
## ordinal score (Gaussian) than as 11 categorical states we cannot estimate.
inc_levels <- c("No income","R1-R400","R401 - R800","R801 - R1600",
                "R1601 - R3200","R3201 - R6400","R6401 - R12800",
                "R12801 - R25600","R25601 - R51200","R51201 - R102400")
inc_raw <- col("X61\\.\\.")
D$IncomeBand <- sapply(inc_raw, function(s) {
  if (is.na(s) || grepl("Prefer not", s)) return(NA_real_)
  hit <- which(startsWith(s, inc_levels))
  if (length(hit)) hit[1] - 1 else NA_real_
}, USE.NAMES = FALSE)                                                  # "g"

D$Grant <- yn(col("X59\\.\\."))                                        # "c", 2

## ---------------------------------------------------------------------------
## 5. HOUSING, WATER, SANITATION, HYGIENE (the WASH block)
## ---------------------------------------------------------------------------
dw <- col("X95\\.\\.")
D$FormalDwelling <- as.integer(grepl("^House|^Room/flatlet", dw))      # "c", 2

## Two versions of the water question exist in the export: an old single-choice
## item (X110) and a newer checkbox block (X112.*). We take the old item where
## present and fall back to the checkboxes, then reduce to the distinction that
## actually matters for contamination risk: water inside the dwelling or not.
w_old <- col("X110\\.\\.")
w_in  <- as.integer(yn0(block("X112", 1)))
w_yd  <- as.integer(yn0(block("X112", 2)))
D$PipedInside <- ifelse(
  !is.na(w_old) & w_old == "Piped water inside the dwelling", 1L,
  ifelse(!is.na(w_old) & w_old %in% c("Piped water outside the dwelling (e.g., yard)",
                                      "Public tap/standpipe","Other"), 0L,
         ifelse(w_in == 1, 1L, ifelse(w_yd == 1, 0L, NA_integer_))))   # "c", 2

## Toilet: flush (to sewer or septic) versus pit latrine / public / none.
## The intermediate "pit with slab" group has only 14 members, too few to
## support its own categorical state at n = 164, so it joins the unimproved arm.
tl <- col("X118\\.\\.")
D$FlushToilet    <- as.integer(grepl("flush", tl, ignore.case = TRUE)) # "c", 2
D$ToiletShared   <- yn(col("X121\\.\\."))                              # "c", 2
D$ToiletFloods   <- yn(col("X120\\.\\."))                              # "c", 2
D$OpenDefecation <- as.integer(col("X122\\.\\.") != "Never")           # "c", 2

## X126.1 and X128.1 are Yes/No items, not tick boxes: the export records an
## explicit "No" (X126.1 is Yes 105 / No 56 / blank 8; X128.1 has 32 blanks).
## yn_item() therefore treats a blank as missing under STRICT_MISSING.
D$StandingWater  <- as.integer(yn_item(block("X126", 1)))              # "c", 2
D$Flooding       <- as.integer(yn_item(block("X128", 1)))              # "c", 2
D$RefuseCollected <- as.integer(startsWith(
  ifelse(is.na(col("X108\\.\\.")), "", col("X108\\.\\.")), "Removed")) # "c", 2

## Handwashing: 8 Yes/No items summed into a 0-8 practice score. A sum of
## indicators is exactly the kind of variable the Poisson node is for.
##
## These items also carry an explicit "No", and 82 of 164 households leave
## between 1 and 7 of them blank. Summing "== Yes" over the raw block counts
## every blank as a No and pushes those 82 scores down. Under STRICT_MISSING
## the score is prorated over the items the household actually answered, and
## set to NA (for section 9 to impute) when fewer than five were answered.
hw      <- block("X123", 1:8)
hw_yes  <- rowSums(hw == "Yes", na.rm = TRUE)
hw_ans  <- rowSums(!is.na(hw))
if (STRICT_MISSING) {
  D$HandwashScore <- ifelse(hw_ans >= 5L,
                            as.integer(round(8 * hw_yes / hw_ans)),
                            NA_integer_)                               # "p"
  cat("Handwashing items answered per household:\n"); print(table(hw_ans))
} else {
  D$HandwashScore <- hw_yes                                            # "p"
}

## ---------------------------------------------------------------------------
## 6. HEALTH SERVICE USE AND ANTIBIOTIC BEHAVIOUR
## ---------------------------------------------------------------------------
cf <- col("X62\\.\\.")
## "Never" has only 2 observations, so it is merged with "Yearly".
D$CareMonthly <- as.integer(cf %in% c("Weekly", "Montly"))             # "c", 2

D$HIVinHH <- yn(col("X82\\.\\."))                                      # "c", 2

ch <- col("X76\\.\\.")
D$ChronicAny <- ifelse(is.na(ch), NA_integer_,
                       as.integer(ch != "No chronic conditions"))      # "c", 2

sev <- col("X81\\.\\.")
D$DiarrSeverity <- ifelse(sev == "No one experiences diarrhoea", 1L,
                   ifelse(sev == "Mild (resolves in 1-2 days)",   2L,
                   ifelse(!is.na(sev), 3L, NA_integer_)))              # "c", 3
# 1 = no diarrhoea, 2 = mild, 3 = moderate or severe

D$AbxNoRx <- yn(col("X90\\.\\."))                                      # "c", 2

src <- col("X86\\.\\.")
D$AbxSource <- ifelse(is.na(src), NA_integer_,
               ifelse(src == "Not applicable", 1L,
               ifelse(src == "Government facility (Clinic; Community Heath Centre, Hospital)",
                      2L, 3L)))                                        # "c", 3
# 1 = does not source antibiotics, 2 = government facility only,
# 3 = private GP and/or pharmacy (with or without government)

crs <- col("X92\\.\\.")
## "Until the package is empty" (n = 2) is merged with "As advised", since both
## describe completing a full prescribed course.
D$AbxCourse <- ifelse(crs %in% c("As advised by a healthcare provider",
                                 "Until the package is empty"), 1L,
               ifelse(crs == "Until symptoms are resolved or cured", 2L,
               ifelse(crs == "I do not take antibiotics", 3L, NA_integer_)))
                                                                       # "c", 3
# 1 = completes course, 2 = stops when symptoms resolve, 3 = does not take

## ---------------------------------------------------------------------------
## 7. FOOD, ANIMALS, ENVIRONMENTAL EXPOSURE
## ---------------------------------------------------------------------------
D$StreetFood <- as.integer(col("X135\\.\\.") != "Don't consume street food")

## Meat frequency: the questionnaire changed mid-study (X129 old, X131 new) and
## both are checkbox grids of frequency x meat type. We take the HIGHEST
## frequency ticked for any meat type: 1 = never ... 6 = daily, minus 1 so the
## score starts at 0.
meat_blocks <- cbind(block("X129", 1:6), block("X131", 1:6))
meat_level  <- rep(NA_real_, nrow(raw))
for (j in 1:6) {
  ticked <- !is.na(meat_blocks[, j]) | !is.na(meat_blocks[, j + 6])
  meat_level[ticked] <- j
}
D$MeatFreq <- meat_level - 1                                           # "g"

D$AnimalsKept <- yn(col("X157\\.\\."))                                 # "c", 2
## Manure: five Yes/No items. 63 of 164 households leave all five blank --
## under the old rule every one of them became "does not use manure". Under
## STRICT_MISSING an all-blank row is missing; a partially answered row is an
## OR over the items that were answered.
mn      <- block("X150", 1:5)
mn_yes  <- rowSums(mn == "Yes", na.rm = TRUE)
mn_ans  <- rowSums(!is.na(mn))
## (braces, not a bare multi-line if/else: at the console R closes the `if`
## at the end of the first line and the stray `else` is a syntax error)
D$ManureUse <- if (STRICT_MISSING) {
  ifelse(mn_ans > 0L, as.integer(mn_yes > 0), NA_integer_)
} else {
  as.integer(mn_yes > 0)
}                                                                      # "c", 2

## ---------------------------------------------------------------------------
## 8. Variable registry: name -> type, level, group
## ---------------------------------------------------------------------------
## This registry IS the contract with mgm(). Everything downstream (fitting,
## plotting, the Shiny app) reads `type` and `level` from here, so a variable
## can never drift out of sync with its declared distribution.

registry <- tibble::tribble(
  ~var,             ~type, ~level, ~group,        ~label,
  # --- outcomes -------------------------------------------------------------
  "ESBL_pos_n",     "p",   1L,     "Carriage",    "ESBL+ months",
  "EC_pos_n",       "p",   1L,     "Carriage",    "E. coli+ months",
  "KP_pos_n",       "p",   1L,     "Carriage",    "K. pneumoniae+ months",
  "Months_tested",  "g",   1L,     "Carriage",    "Months cultured",
  # --- demography / SES -----------------------------------------------------
  "Age",            "g",   1L,     "Demographic", "Age",
  "Sex",            "c",   2L,     "Demographic", "Sex (1=F)",
  "HHsize",         "p",   1L,     "Demographic", "Household size",
  "Crowding",       "g",   1L,     "Demographic", "Persons per room",
  "Education",      "c",   3L,     "Socioeconomic","Highest education",
  "WorkStatus",     "c",   3L,     "Socioeconomic","Work status",
  "IncomeBand",     "g",   1L,     "Socioeconomic","Income band",
  "Grant",          "c",   2L,     "Socioeconomic","Receives grant",
  # --- WASH -----------------------------------------------------------------
  "FormalDwelling", "c",   2L,     "WASH",        "Formal dwelling",
  "PipedInside",    "c",   2L,     "WASH",        "Piped water indoors",
  "FlushToilet",    "c",   2L,     "WASH",        "Flush toilet",
  "ToiletShared",   "c",   2L,     "WASH",        "Shared toilet",
  "ToiletFloods",   "c",   2L,     "WASH",        "Toilet floods",
  "OpenDefecation", "c",   2L,     "WASH",        "Open defecation",
  "StandingWater",  "c",   2L,     "WASH",        "Standing water",
  "Flooding",       "c",   2L,     "WASH",        "Household flooding",
  "RefuseCollected","c",   2L,     "WASH",        "Refuse collected",
  "HandwashScore",  "p",   1L,     "WASH",        "Handwashing score (0-8)",
  # --- health service use / antibiotics -------------------------------------
  "CareMonthly",    "c",   2L,     "Health",      "Care sought monthly+",
  "HIVinHH",        "c",   2L,     "Health",      "HIV in household",
  "ChronicAny",     "c",   2L,     "Health",      "Any chronic condition",
  "DiarrSeverity",  "c",   3L,     "Health",      "Diarrhoea severity",
  "AbxNoRx",        "c",   2L,     "Antibiotic",  "Antibiotics w/o Rx",
  "AbxSource",      "c",   3L,     "Antibiotic",  "Antibiotic source",
  "AbxCourse",      "c",   3L,     "Antibiotic",  "Course completion",
  # --- food / animals -------------------------------------------------------
  "StreetFood",     "c",   2L,     "Food/animal", "Eats street food",
  "MeatFreq",       "g",   1L,     "Food/animal", "Meat frequency",
  "AnimalsKept",    "c",   2L,     "Food/animal", "Keeps animals",
  "ManureUse",      "c",   2L,     "Food/animal", "Uses manure"
)

## A smaller "core" set for when you want more power per parameter. With
## n = 164 the full 32-node model is a genuinely high-dimensional problem
## (the l1 penalty keeps it identified, but sensitivity is low).
## WHY THIS SUBSET, AND WHAT IT COSTS. The registry carries 41 variables; the
## model is fitted on these 21 plus the spatial nodes. The subset is the set of
## variables the study was designed around -- carriage, demography, WASH,
## antibiotic use and animal contact -- and it is held down because n = 162 and
## a nodewise LASSO over 41 nodes on 162 households has very little power.
## The cost has to be stated, not left implicit: refuse collection, household
## flooding, formal dwelling and manure use are NOT nodes in the core set, so a
## core-set network is silent about them by construction rather than by
## finding. The reported model uses all covariates, not the core set.
core_vars <- c("ESBL_pos_n","EC_pos_n","KP_pos_n","Months_tested","Age","Sex","HHsize",
               "Crowding","Education","IncomeBand","PipedInside","FlushToilet",
               "ToiletShared","OpenDefecation","StandingWater","HandwashScore",
               "HIVinHH","AbxNoRx","AbxSource","AbxCourse","AnimalsKept")

## ---------------------------------------------------------------------------
## 9. Missing data
## ---------------------------------------------------------------------------
## mgm() cannot take NA. Missingness here is light (<= 6 cells per variable),
## so median (continuous / count) or modal (categorical) single imputation is
## used by default. Set IMPUTE <- FALSE for complete-case analysis instead.
IMPUTE <- TRUE

miss_report <- sapply(registry$var, function(v) sum(is.na(D[[v]])))
cat("\nMissing cells per model variable (STRICT_MISSING =", STRICT_MISSING, "):\n")
if (any(miss_report > 0)) print(miss_report[miss_report > 0]) else cat("  none\n")
cat(sprintf("Total missing cells: %d across %d variables (%.2f%% of the matrix)\n",
            sum(miss_report), sum(miss_report > 0),
            100 * sum(miss_report) / (nrow(D) * nrow(registry))))

if (IMPUTE) {
  for (i in seq_len(nrow(registry))) {
    v <- registry$var[i]
    x <- D[[v]]
    if (!any(is.na(x))) next
    fill <- if (registry$type[i] == "c") {
      as.numeric(names(sort(table(x), decreasing = TRUE))[1])   # mode
    } else {
      median(x, na.rm = TRUE)                                    # median
    }
    D[[v]][is.na(x)] <- fill
  }
} else {
  D <- D[complete.cases(D[, registry$var]), ]
}
cat("Rows retained:", nrow(D), "\n")

## ---------------------------------------------------------------------------
## 10. Validate the coding against mgm's requirements
## ---------------------------------------------------------------------------
## Any violation caught here would otherwise surface as an obscure glmnet error
## deep inside mgm(), so we check explicitly:
##   - "p" variables are non-negative integers
##   - "c" variables have exactly `level` distinct consecutive values,
##     starting at 0 (binary, so binarySign works) or 1 (m > 2)
for (i in seq_len(nrow(registry))) {
  v <- registry$var[i]; ty <- registry$type[i]; lv <- registry$level[i]
  x <- D[[v]]
  if (ty == "p") {
    stopifnot(all(x >= 0), all(x == round(x)))
  }
  if (ty == "c") {
    u <- sort(unique(x))
    if (lv == 2) {
      if (!identical(as.numeric(u), c(0, 1)))
        stop("Binary variable ", v, " is not coded 0/1: ", paste(u, collapse = ","))
    } else {
      if (!identical(as.numeric(u), as.numeric(seq_len(lv))))
        stop("Categorical ", v, " must be coded 1..", lv, "; found: ",
             paste(u, collapse = ","))
    }
  }
}
cat("Coding validation passed for all", nrow(registry), "model variables.\n")

## ---------------------------------------------------------------------------
## 11. Assemble and save the MGM object
## ---------------------------------------------------------------------------
group_cols <- c(Carriage = "#B2182B", Demographic = "#EF8A62",
                Socioeconomic = "#FDDBC7", WASH = "#67A9CF",
                Health = "#2166AC", Antibiotic = "#7B3294",
                `Food/animal` = "#5AAE61")

AIARMS <- list(
  data         = as.matrix(D[, registry$var]),
  type         = registry$type,
  level        = registry$level,
  labels       = registry$label,
  vars         = registry$var,
  groups       = split(seq_len(nrow(registry)), registry$group),
  groups_color = unname(group_cols[names(split(seq_len(nrow(registry)),
                                               registry$group))]),
  registry     = registry,
  core_vars    = core_vars,
  clean_frame  = D
)

saveRDS(AIARMS, file.path(OUT_DIR, "AIARMS_mgm.rds"))
write.csv(D, file.path(OUT_DIR, "AIARMS_clean.csv"), row.names = FALSE)

cat("\nSaved:", file.path(OUT_DIR, "AIARMS_mgm.rds"), "\n")
cat("Model matrix:", nrow(AIARMS$data), "rows x", ncol(AIARMS$data), "nodes\n")
print(table(AIARMS$type))

## The sentence in Section 1 of the report that counts variable types has to
## match this, so print it in the report's own words rather than from memory.
n_bin <- sum(registry$type == "c" & registry$level == 2)
n_cat <- sum(registry$type == "c" & registry$level >  2)
n_cnt <- sum(registry$type == "p")
n_con <- sum(registry$type == "g")
cat(sprintf(
  "\nFOR THE REPORT: %d variables -- %d binary indicators, %d categorical,\n%d counts and %d continuous measurements.\n",
  nrow(registry), n_bin, n_cat, n_cnt, n_con))
cat("(Of the", n_con, "Gaussian nodes only Age and Crowding are genuinely\n",
    "continuous; Months_tested, IncomeBand and MeatFreq are bounded ordinal\n",
    "scores declared 'g'. Say so in the report rather than calling them\n",
    "continuous measurements.)\n")

## ---------------------------------------------------------------------------
## 12. Tidy up two helpers that shadow base R
## ---------------------------------------------------------------------------
## col() above shadows base::col() (the column indices of a matrix) and norm()
## shadows base::norm(). Inside this script that is harmless -- nothing here
## wants the base versions. It stops being harmless the moment this section
## shares a global environment with anything else: the spatial analysis
## unpacks a matrix with base::col(), which would reach the questionnaire
## helper instead and fail with
##     No column starting with: 0.14146335481534...
## because the matrix gets pasted into a regular expression. Both helpers have
## done their job by this point, so they do not stay in the workspace.
rm(col, norm)

```

---

# The spatial block {#spatial-block}

`01b_add_spatial.R`. Joins the coordinates back on, projects them to local
metres, drops the two geocoding outliers, derives the spatial covariates and
extends the registry. This is where the fieldwork site codes become nodes.

```{r Spatial component, purl=FALSE}
###############################################################################
# 01b_add_spatial.R
#
# Adds a SPATIAL block to the cleaned AIARMS object without touching
# 01_clean_AIARMS.R. Reads output/AIARMS_mgm.rds, joins coordinates back from
# the raw export, derives spatial covariates, and writes
# output/AIARMS_mgm_spatial.rds with the same structure plus the new nodes.
#
# Run AFTER 01_clean_AIARMS.R:
#     source("01_clean_AIARMS.R")
#     source("01b_add_spatial.R")
#
# ---------------------------------------------------------------------------
# WHAT CHANGED IN THIS VERSION
# ---------------------------------------------------------------------------
# The spatial grouping is now the FIELDWORK SITE CODE itself -- ARUE, ARUO,
# ARUT, ARUL, ARUS, ARUU, ARUF -- rather than an anonymous merged "Zone".
#
#   * Node renamed Zone -> Site, and its categories keep their real codes, so
#     the Area Profile and showInteraction() output name actual sites instead
#     of the integers 1..4.
#   * SITE_MIN_N is 0: no site is merged, so ARUF (n = 3) keeps its own
#     level and all seven codes survive.
#   * SITE_ENCODING is "dummy_full": SEVEN binary nodes, one per site code,
#     labelled "Site ARUE" ... "Site ARUU". Every site appears in the network
#     as itself, with its own edges and its own sign. This is the encoding
#     that answers "which covariates attach to ARUU?" directly.
#     The cost is real and is handled, not ignored: the seven dummies sum to
#     1 in every row, so the SITE-TO-SITE edges among them are an artefact of
#     the coding. The app suppresses that block by default and the write-up
#     must state it. Site-to-covariate edges are what this encoding is for.
#     "categorical" (one node, seven levels) is the identified alternative
#     and stays available; "dummy_reference" drops ARUE entirely. See
#     section 0.
#   * Previously SITE_MIN_N was dropped from 20 to 10. At 20 the merge loop collapsed seven
#     sites into four (ARUF -> ARUU, then ARUS -> ARUO, then ARUU -> ARUO),
#     which is exactly the site identity you wanted to keep. At 10 only ARUF
#     merges. Set SITE_MIN_N <- 0 to keep all seven codes untouched.
#   * The trend surface (East_km / North_km) is now OFF by default. See the
#     note in section 4a: with site membership in the graph the two describe
#     overlapping structure, and including both puts near-collinear nodes in
#     the same model.
#
# THE THREE NOTIONS OF SPACE, AND WHICH ONE EACH NODE ANSWERS
#
#   (a) AREA MEMBERSHIP   -- Site        [ON by default]
#       Which fieldwork area a household belongs to. Captures discrete
#       between-settlement differences: shared standpipe, shared sewer line,
#       shared refuse round. This is the one you asked for.
#
#   (b) LOCAL CONTEXT     -- NN_density  [ON by default]
#       How many other study households sit within a fixed radius. Built
#       environment density as an exposure in its own right, and the spatial
#       variable most likely to connect to crowding and WASH.
#
#   (c) LARGE-SCALE TREND -- East_km, North_km   [OFF by default]
#       A linear trend surface. An edge from East_km means a variable shifts
#       systematically along the east-west axis, conditional on the rest.
#
#   A fourth notion -- the spatially lagged OUTCOME -- is derived in section 6
#   but deliberately left out of the node set. See the note there.
###############################################################################

## Deliberately NOT rm(list = ls()) here. This script is often run a section at
## a time in the console, and wiping the workspace mid-way is how you end up
## with "object 'REG' not found" from section 7. Set RESET <- TRUE if you do
## want a clean slate.
RESET <- FALSE
if (RESET) rm(list = setdiff(ls(), "RESET"))

OUT_DIR  <- "output"
RAW_PATH <- "full_AIARMS_df.csv"

## load_aiarms() is idempotent: safe to call again from any section.
load_aiarms <- function() {
  f <- file.path(OUT_DIR, "AIARMS_mgm.rds")
  if (!file.exists(f))
    stop("Cannot find ", f, "\n",
         "Run 01_clean_AIARMS.R first -- 01b extends its output, it does not ",
         "rebuild it.")
  readRDS(f)
}

AIARMS <- load_aiarms()
REG    <- AIARMS$registry
D      <- AIARMS$clean_frame

## ---------------------------------------------------------------------------
## 0. Switches
## ---------------------------------------------------------------------------
## How the fieldwork site enters the model. All three produce valid models;
## they answer the question differently. Numbers below are from this cohort.
##
##   "categorical"     ONE node, level = m. Every site
##                     code is a level of that node, so all seven appear and
##                     none is a reference. mgm handles it natively: the
##                     nodewise model for Site is one multinomial regression,
##                     and the m x (u-1) parameters behind each edge are
##                     aggregated into a single weight (Haslbeck & Waldorp
##                     2020, Sec. 2.2). Per-site detail is recovered with
##                     showInteraction(). One edge per partner, and no sign,
##                     because a sign is not definable for m > 2.
##                     Caveat to state in the write-up: ARUF has n = 3, so
##                     that level is estimated from three households. Read
##                     any edge that turns on ARUF as hypothesis-generating.
##
##   "dummy_reference" m-1 BINARY nodes, one site held out as reference.
##                     RECOMMENDED if you want per-site nodes: each site gets
##                     its own node, its own edges, and -- because binary nodes
##                     are coded 0/1 with binarySign = TRUE -- its own SIGN.
##                     Predicting each dummy from the others gives R^2 of
##                     0.18-0.27 here, so they are ordinary correlated
##                     predictors, nothing degenerate. Cost: the reference site
##                     has no node, and every effect is read relative to it.
##
##   "dummy_full"      ALL m binary nodes -- THE DEFAULT HERE. Every site code
##                     is a node in its own right: ARUE, ARUF, ARUL, ARUO,
##                     ARUS, ARUT, ARUU each appear in the network with their
##                     own edges and, because binary nodes are coded 0/1 with
##                     binarySign = TRUE, their own SIGN. Nothing is held out
##                     and no edge is a contrast against a reference site.
##
##                     THE COST, STATED PLAINLY. The dummies sum to 1 in every
##                     row, so predicting any one from the other m-1 gives
##                     R^2 = 1 exactly. Two consequences:
##                       (i)  each site node's own nodewise logistic regression
##                            is perfectly separated by the other six, which
##                            produces a clique of strong SITE-TO-SITE edges
##                            that is an artefact of the coding. Ignore that
##                            block; the app zeroes it by default.
##                       (ii) inside a COVARIATE's regression the seven dummies
##                            plus the intercept are rank-deficient, so the
##                            lasso picks one of several equivalent solutions.
##                            Site-to-covariate edges are therefore stable in
##                            aggregate but which individual site carries an
##                            effect can move between bootstrap samples. Check
##                            anything you intend to report against the
##                            Stability tab, and against the "categorical" fit.
##
##                     Use "categorical" when you need an identified model to
##                     defend; use this when you need to see the sites.
##
## Note: mgm's own overparameterize = TRUE is NOT this. There the indicators
## are predictors inside another node's design matrix and never receive a
## conditional model of their own, which is why the L1 penalty is enough there
## and is not enough here.
SITE_ENCODING  <- "dummy_full"    # one binary node per site; all seven kept
SITE_REFERENCE <- NULL   # NULL = use the largest site; or name one, e.g. "ARUE"

SITE_MIN_N     <- 0      # merge a site into its nearest neighbour below this n
                         #   0  -> keep all seven codes exactly as they are
                         #   10 -> only ARUF (n = 3) merges, leaving six sites
                         #   20 -> collapses to four zones (the old behaviour)
INCLUDE_TREND  <- FALSE  # add East_km / North_km alongside Site
DROP_OUTLIERS  <- TRUE
OUTLIER_KM     <- 10
RADIUS_M       <- 150    # neighbourhood radius for NN_density
K_NN           <- 8      # neighbours for the spatial weights matrix

## ---------------------------------------------------------------------------
## 1. Join coordinates back on Study_code
## ---------------------------------------------------------------------------
raw <- read.csv(RAW_PATH, stringsAsFactors = FALSE, check.names = TRUE,
                encoding = "UTF-8", na.strings = c("", "NA", "na", "N/A"))
stopifnot(nrow(raw) > 0L)                      # see the note in 01 on encoding
raw <- raw[!duplicated(raw$Study_code), ]      # same de-duplication as 01

geo <- raw[, c("Study_code", "Longitude.x", "Latitude.x")]
D   <- merge(D, geo, by = "Study_code", all.x = TRUE, sort = FALSE)
D   <- D[match(AIARMS$clean_frame$Study_code, D$Study_code), ]   # restore order
stopifnot(!any(is.na(D$Latitude.x)))

## Fieldwork site is encoded in the Study_code prefix.
D$Site_code <- sub("^([A-Za-z]+).*$", "\\1", D$Study_code)
cat("Fieldwork sites as recorded:\n"); print(table(D$Site_code))

## ---------------------------------------------------------------------------
## 2. Project longitude/latitude to local metres
## ---------------------------------------------------------------------------
## Over a study area a few km across, an equirectangular projection about the
## median point is accurate to well under a metre, so there is no need for sf
## or a formal CRS transformation. Distances are then ordinary Euclidean.
R_EARTH <- 6371000
lat0 <- median(D$Latitude.x); lon0 <- median(D$Longitude.x)
D$x_m <- R_EARTH * cos(lat0 * pi/180) * (D$Longitude.x - lon0) * pi/180
D$y_m <- R_EARTH * (D$Latitude.x - lat0) * pi/180

## ---------------------------------------------------------------------------
## 3. Outlying households
## ---------------------------------------------------------------------------
## Two ARUO households are recorded ~62 km from every other point, at identical
## coordinates. Whether that is a second site or a GPS/data-entry error, they
## would dominate any distance-based quantity: the trend surface would fit one
## line through them, and their nearest-neighbour density would be 1.
d_centre <- sqrt(D$x_m^2 + D$y_m^2) / 1000
is_out   <- d_centre > OUTLIER_KM
cat("\nHouseholds beyond", OUTLIER_KM, "km of the study centroid:", sum(is_out), "\n")
if (sum(is_out)) print(D[is_out, c("Study_code", "Longitude.x", "Latitude.x")])

if (DROP_OUTLIERS && any(is_out)) {
  D <- D[!is_out, ]
  cat("Dropped. Retained", nrow(D), "households for the spatial model.\n")
}

## ---------------------------------------------------------------------------
## 4. Spatial covariates
## ---------------------------------------------------------------------------

## (a) Trend surface -- OFF unless INCLUDE_TREND is TRUE.
##     Kilometres rather than metres so the coefficients sit on a scale
##     comparable to the other Gaussian nodes after scale = TRUE.
##
##     Why off by default now that Site is in the model: the sites ARE the
##     spatial partition of this cohort, and their centroids span barely 2 km.
##     A linear surface fitted across that span is largely a smoothed version
##     of site membership, so the two nodes compete to explain the same
##     variation. The lasso then splits the signal arbitrarily between them and
##     both edges look weaker than the structure actually is. Turn it on only
##     if you specifically want to ask whether anything varies smoothly ACROSS
##     sites rather than BETWEEN them.
D$East_km  <- (D$x_m - mean(D$x_m)) / 1000                            # "g"
D$North_km <- (D$y_m - mean(D$y_m)) / 1000                            # "g"

## (b) Local density: number of OTHER study households within RADIUS_M.
##     A count of points in a disc, so a Poisson node. Median nearest-neighbour
##     distance is about 17 m, so 150 m captures the immediate cluster.
dmat <- as.matrix(dist(cbind(D$x_m, D$y_m)))
D$NN_density <- rowSums(dmat < RADIUS_M) - 1                          # "p"

## (c) Site: the fieldwork area code, kept as the grouping.
##
##     Merging rule: while any site has fewer than SITE_MIN_N households, the
##     smallest is merged into the site with the NEAREST centroid. Merging by
##     proximity rather than by convenience keeps every resulting group
##     spatially coherent -- a merged group is still a contiguous patch of the
##     study area, not a scattering of unrelated households.
##
##     Why a floor exists at all: a categorical level is estimated from the
##     households in it. ARUF has three. Three observations cannot support the
##     parameters of their own multinomial level, and mgm will either drop the
##     level or produce an estimate driven by those three rows. Setting
##     SITE_MIN_N <- 0 keeps all seven codes if you would rather see that
##     happen than merge.
site <- D$Site_code
merge_log <- character(0)
repeat {
  tab <- table(site)
  if (all(tab >= SITE_MIN_N) || length(tab) <= 2) break
  small <- names(tab)[which.min(tab)]
  cent  <- t(sapply(split(seq_along(site), site),
                    function(i) c(mean(D$x_m[i]), mean(D$y_m[i]))))
  others <- setdiff(rownames(cent), small)
  dd <- sqrt((cent[others, 1] - cent[small, 1])^2 +
             (cent[others, 2] - cent[small, 2])^2)
  target <- others[which.min(dd)]
  merge_log <- c(merge_log,
                 sprintf("  %s (n = %d) -> %s  (centroids %.2f km apart)",
                         small, min(tab), target, min(dd) / 1000))
  site[site == small] <- target
}

site_f      <- factor(site)
site_labels <- levels(site_f)
n_site      <- nlevels(site_f)

## Build whichever node set SITE_ENCODING asked for.
site_vars <- character(0)

if (SITE_ENCODING == "categorical") {
  D$Site    <- as.integer(site_f)                                     # "c", m
  site_vars <- "Site"
  ## The integer code is what mgm needs, but nothing downstream should ever
  ## have to guess that 4 means ARUO. site_levels travels with the object and
  ## is what the spatial analysis and the plots label with.
  site_levels <- site_labels
  names(site_levels) <- seq_along(site_labels)

} else {
  ## One binary node per site, coded 0/1 so binarySign = TRUE gives each a sign.
  all_sites <- site_labels
  if (SITE_ENCODING == "dummy_reference") {
    ref <- if (is.null(SITE_REFERENCE)) names(which.max(table(site)))
           else SITE_REFERENCE
    if (!ref %in% all_sites)
      stop("SITE_REFERENCE '", ref, "' is not one of: ",
           paste(all_sites, collapse = ", "))
    kept <- setdiff(all_sites, ref)
    cat("\nReference site (absorbed, no node):", ref,
        sprintf("(n = %d)\n", sum(site == ref)))
  } else if (SITE_ENCODING == "dummy_full") {
    kept <- all_sites
    message("SITE_ENCODING = 'dummy_full': all ", length(kept), " site codes ",
            "get their own binary node.\n",
            "  The dummies sum to 1 in every row, so the SITE-TO-SITE edges ",
            "among them are an\n  artefact of the coding, not a finding. Read ",
            "site-to-covariate edges only; the app\n  suppresses the site-site ",
            "block by default. Report this in the methodology.")
  } else {
    stop("SITE_ENCODING must be 'categorical', 'dummy_reference' or 'dummy_full'.")
  }
  for (a in kept) {
    v <- paste0("Site_", a)
    D[[v]] <- as.integer(site == a)                                   # "c", 2
    site_vars <- c(site_vars, v)
  }
}
D$Site_group <- site      # merged code as text, for maps and tables

cat("\nSite grouping (SITE_MIN_N =", SITE_MIN_N, "):\n")
if (length(merge_log)) {
  cat("merges performed:\n"); cat(merge_log, sep = "\n"); cat("\n")
} else {
  cat("no merges needed; all site codes retained\n")
}
print(table(site))
cat("\nInteger code -> site:\n")
print(setNames(site_labels, seq_len(n_site)))

## (d) Distance from the centre of the study area. Correlated with the trend
##     surface by construction, so it is derived but NOT in the default set.
D$Dist_centre_km <- sqrt((D$x_m - mean(D$x_m))^2 +
                         (D$y_m - mean(D$y_m))^2) / 1000

## ---------------------------------------------------------------------------
## 5. Spatial weights matrix (k nearest neighbours, row-standardised)
## ---------------------------------------------------------------------------
## Kept in the saved object for spatial checks; not used by the MGM itself,
## apart from building the lagged carriage below. Symmetrised so that
## "i is a neighbour of j" implies the reverse, then row-standardised so each
## household's neighbours carry a total weight of 1.
## CORRECTNESS NOTE. The previous version took order(dmat[i, ])[2:(K_NN+1)]
## and assumed position 1 is household i. That holds only when i is the
## unique minimiser of its own distance row. Seven pairs of households in
## this cohort sit at identical coordinates, so for 7 of the 162 rows
## order() put the OTHER household first and i itself second -- and i was
## then written into its own neighbour set. After symmetrising and row
## standardising, diag(W) was nonzero for those 7 rows, up to w_ii = 0.182.
## A household should never be its own neighbour (w_ii = 0), so excluding i
## explicitly fixes it.
W <- matrix(0, nrow(D), nrow(D))
for (i in seq_len(nrow(D))) {
  nb <- setdiff(order(dmat[i, ]), i)[seq_len(K_NN)]
  W[i, nb] <- 1
}
W <- (W + t(W)) / 2
W <- W / rowSums(W)
stopifnot(all(diag(W) == 0), all(abs(rowSums(W) - 1) < 1e-12))

## ---------------------------------------------------------------------------
## 6. Spatially lagged carriage  (derived, but OFF by default)
## ---------------------------------------------------------------------------
## The mean carriage of a household's neighbours. As a graph node this is
## seductive but awkward, for two reasons worth being explicit about:
##
##   1. ENDOGENEITY. It is built from other rows' values of the very outcome
##      it would be joined to. In spatial econometrics this is the lagged
##      dependent variable, and it is not exogenous; the resulting edge cannot
##      be read the way the other edges are read.
##   2. INDEPENDENCE. An MGM assumes i.i.d. rows. A node constructed across
##      rows breaks that assumption directly.

D$ESBL_lag <- as.numeric(W %*% D$ESBL_pos_n)
D$KP_lag   <- as.numeric(W %*% D$KP_pos_n)

## ---------------------------------------------------------------------------
## Prerequisite guard
## ---------------------------------------------------------------------------
## If you jumped straight here, these recover or explain what is missing rather
## than failing on an unhelpful "object not found".
if (!exists("AIARMS")) AIARMS <- load_aiarms()
if (!exists("REG"))    REG    <- AIARMS$registry
need <- c("D", "n_site", "site_labels", "site_vars", "RADIUS_M", "dmat", "W")
gone <- need[!vapply(need, exists, logical(1))]
if (length(gone))
  stop("Section 7 needs objects built earlier in this script: ",
       paste(gone, collapse = ", "),
       "\nRun the whole file instead:  source(\"01b_add_spatial.R\")")

## ---------------------------------------------------------------------------
## 7. Extend the registry
## ---------------------------------------------------------------------------
## NN_density is an ADDITION to the site nodes, not a replacement for them.
## They answer different questions: site membership is which settlement you are
## in, density is how built-up your immediate surroundings are. Set
## INCLUDE_DENSITY <- FALSE if you want the site nodes alone.
INCLUDE_DENSITY <- TRUE

site_reg <- if (SITE_ENCODING == "categorical") {
  data.frame(var = "Site", type = "c", level = as.integer(n_site),
             group = "Spatial",
             label = "Fieldwork site",
             stringsAsFactors = FALSE)
} else {
  data.frame(var = site_vars, type = "c", level = 2L, group = "Spatial",
             label = if (SITE_ENCODING == "dummy_reference")
                       sprintf("Site %s (vs %s)", sub("^Site_", "", site_vars), ref)
                     else paste("Site", sub("^Site_", "", site_vars)),
             stringsAsFactors = FALSE)
}

spatial_reg <- site_reg
if (INCLUDE_DENSITY)
  spatial_reg <- rbind(
    data.frame(var = "NN_density", type = "p", level = 1L, group = "Spatial",
               label = paste0("Neighbours within ", RADIUS_M, "m"),
               stringsAsFactors = FALSE),
    spatial_reg)
if (INCLUDE_TREND) {
  spatial_reg <- rbind(
    data.frame(var = c("East_km", "North_km"), type = "g", level = 1L,
               group = "Spatial", label = c("Easting (km)", "Northing (km)"),
               stringsAsFactors = FALSE),
    spatial_reg)
}
REG2 <- rbind(as.data.frame(REG), spatial_reg)

## Same validation as in 01: "p" non-negative integers, "c" consecutive from
## 0 (binary) or 1 (m > 2).
for (i in seq_len(nrow(spatial_reg))) {
  v <- spatial_reg$var[i]; ty <- spatial_reg$type[i]; lv <- spatial_reg$level[i]
  x <- D[[v]]
  if (anyNA(x)) stop("Spatial variable ", v, " has missing values.")
  if (ty == "p") stopifnot(all(x >= 0), all(x == round(x)))
  if (ty == "c") {
    u <- sort(unique(x))
    ok <- if (lv == 2) identical(as.numeric(u), c(0, 1)) else
                       identical(as.numeric(u), as.numeric(seq_len(lv)))
    if (!ok) stop("Categorical ", v, " badly coded: ", paste(u, collapse = ","))
  }
}
cat("\nSpatial coding validation passed.\n")

## ---------------------------------------------------------------------------
## 8. Build the model matrix, with a guard
## ---------------------------------------------------------------------------
## mgm() rejects anything that is not integer or numeric with the message
##   "Only integer and numeric values permitted."
## which does not say WHICH column is at fault. as.matrix() on a data.frame
## containing even one character or factor column silently coerces EVERY
## column to character, so a single stray column poisons the whole matrix and
## the error surfaces much later, inside mgm(). This guard names the offender
## instead, and coerces the cases that are safe to coerce.
build_model_matrix <- function(df, vars) {
  missing_vars <- setdiff(vars, names(df))
  if (length(missing_vars))
    stop("Columns missing from the frame: ", paste(missing_vars, collapse = ", "))

  cls <- vapply(df[vars], function(z) class(z)[1], character(1))
  bad <- vars[!cls %in% c("numeric", "integer", "logical", "factor")]
  if (length(bad)) {
    stop("These columns are not numeric and cannot be coerced safely:\n",
         paste(sprintf("  %s  (class %s, e.g. %s)", bad, cls[bad],
                       vapply(df[bad], function(z)
                         paste(utils::head(unique(as.character(z)), 3),
                               collapse = " / "), character(1))),
               collapse = "\n"),
         "\nRemove them from the registry, or convert them upstream.")
  }
  ## logical -> 0/1, factor -> its integer codes; both are legitimate here.
  for (v in vars) {
    if (is.logical(df[[v]])) df[[v]] <- as.integer(df[[v]])
    if (is.factor(df[[v]]))  df[[v]] <- as.integer(df[[v]])
  }
  M <- as.matrix(df[, vars, drop = FALSE])
  storage.mode(M) <- "double"
  if (anyNA(M)) {
    na_cols <- vars[colSums(is.na(M)) > 0]
    stop("NA values remain in: ", paste(na_cols, collapse = ", "),
         " -- mgm() cannot take missing data.")
  }
  M
}

Xall <- build_model_matrix(D, REG2$var)
cat("Model matrix built:", nrow(Xall), "x", ncol(Xall),
    "| storage mode:", storage.mode(Xall), "\n")

## Core set gains the spatial block, so the default fit in 02 answers the
## question this script was added for.
core_vars2 <- c(AIARMS$core_vars, spatial_reg$var)

group_cols <- c(Carriage = "#B2182B", Demographic = "#EF8A62",
                Socioeconomic = "#FDDBC7", WASH = "#67A9CF",
                Health = "#2166AC", Antibiotic = "#7B3294",
                `Food/animal` = "#5AAE61", Spatial = "#1B7837")

AIARMS_SP <- list(
  data         = Xall,
  type         = REG2$type,
  level        = REG2$level,
  labels       = REG2$label,
  vars         = REG2$var,
  groups       = split(seq_len(nrow(REG2)), REG2$group),
  groups_color = unname(group_cols[names(split(seq_len(nrow(REG2)), REG2$group))]),
  registry     = REG2,
  core_vars    = core_vars2,
  clean_frame  = D,
  ## spatial extras carried alongside for the diagnostics and the map
  coords       = as.matrix(D[, c("x_m", "y_m")]),
  lonlat       = as.matrix(D[, c("Longitude.x", "Latitude.x")]),
  W            = W,
  K_NN         = K_NN,
  radius_m     = RADIUS_M,
  site_labels  = site_labels,      # every retained site code, in level order
  site_vars    = site_vars,        # the node name(s) carrying site information
  site_encoding = SITE_ENCODING,
  site_group   = D$Site_group,     # merged code per household, as text
  site         = D$Site_code       # original, unmerged code per household
)

saveRDS(AIARMS_SP, file.path(OUT_DIR, "AIARMS_mgm_spatial.rds"))
write.csv(D, file.path(OUT_DIR, "AIARMS_clean_spatial.csv"), row.names = FALSE)

cat("\nSaved:", file.path(OUT_DIR, "AIARMS_mgm_spatial.rds"), "\n")
cat("Model matrix:", nrow(Xall), "rows x", ncol(Xall), "nodes",
    "(", nrow(spatial_reg), "spatial )\n")
print(table(AIARMS_SP$type))

## ---------------------------------------------------------------------------
## 9. Reading the Site node afterwards
## ---------------------------------------------------------------------------
## Site is one categorical node, so the network shows ONE edge weight per
## partner: the mean absolute value of the several parameters behind it, and
## no sign. The per-site parameters are still in the fit. After running
## 02_fit_mgm.R:
##
##   j <- which(AIARMS_SP$vars == "Site")
##   showInteraction(fit_mgm, int = c(j, which(AIARMS_SP$vars == "ESBL_pos_n")))
##
## and map the integer codes back with AIARMS_SP$site_labels.
if (SITE_ENCODING == "categorical") {
  cat("\nSite codes in the fitted model:",
      paste(sprintf("%d=%s", seq_along(site_labels), site_labels), collapse = "  "),
      "\n")
} else {
  cat("\nSite nodes in the fitted model:", paste(site_vars, collapse = ", "), "\n")
  cat("Each is binary 0/1, so each carries its own SIGN in the network.\n")
}
```

---

# The mixed graphical model {#mgm-fit}

`02_fit_mgm.R`. One $\ell_1$-penalised GLM per node, stitched into a graph
(Algorithm 1 of Haslbeck & Waldorp 2020), then predictability and the network
plot. The specification is the one CLIQUE (`app.R`) loads with, so this section
and the app report the same model.

```{r Mixed Graphical Models, purl=FALSE}
###############################################################################
# 02_fit_mgm.R   (mirrors CLIQUE / app.R)
#
# Fits the mixed graphical model that the research report presents, with the
# same data matrix, the same arguments and the same random seed as the app, so
# that the numbers printed here and the numbers shown in the app are the same
# numbers (Haslbeck & Waldorp 2020, JSS 93(8), Section 3.1).
#
#   (1) estimate the model              -> mgm()
#   (2) compute predictions / errors    -> predict()
#   (3) visualise the network           -> qgraph()
#   (4) assess stability (optional)     -> resample() + plotRes()
#
# WHAT HAS TO STAY IN STEP WITH app.R
#   MGM_SEED, MGM_CV_FOLDS            the two constants at the top of app.R
#   variable set                      "All variables" (the app's default)
#   location                          the seven Site_* indicators are replaced
#                                     by ONE categorical Site node, 7 levels
#   lambdaSel / ruleReg / threshold   "CV" / "OR" / "LW" (the app's defaults)
# If any of these is changed in the app, change it here as well.
###############################################################################

## In a combined run the earlier sections have already written their .rds
## outputs, so nothing here needs a clean slate -- and rm(list = ls()) part way
## through a knit would delete objects the later sections rely on. Run
## standalone, it behaves exactly as before.
if (!isTRUE(getOption("knitr.in.progress"))) rm(list = ls())

pkgs <- c("mgm", "qgraph")
new  <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(new)) install.packages(new)
library(mgm); library(qgraph)

OUT_DIR <- "output"

## Copied from app.R. The app seeds the random number generator with MGM_SEED
## immediately before mgm(), because the cross-validation folds are random.
MGM_SEED     <- 20260913L
MGM_CV_FOLDS <- 10L

AIARMS <- readRDS(file.path(OUT_DIR, "AIARMS_mgm_spatial.rds"))

## ---------------------------------------------------------------------------
## 0. The variable set: all variables, as in the app
## ---------------------------------------------------------------------------
v     <- AIARMS$vars
X     <- as.matrix(AIARMS$data)
type  <- AIARMS$type
level <- AIARMS$level
labs  <- AIARMS$labels
if (anyDuplicated(labs)) labs <- make.unique(as.character(labs), sep = " ")
colnames(X) <- labs
grp   <- AIARMS$registry$group[match(v, AIARMS$registry$var)]
grp[is.na(grp)] <- "Other"

## ---------------------------------------------------------------------------
## 0a. Location: one categorical Site node
## ---------------------------------------------------------------------------
## The object ships seven binary Site_* indicators. They sum to one in every
## row and are therefore not identified, so -- exactly as app.R does before
## every fit -- they are replaced by a single categorical node whose levels are
## the seven study codes in alphabetical order. The node is appended as the
## last column, which is also where the app puts it.
site_levels <- character(0)
site_cols   <- grep("^Site_", v)
if (length(site_cols) > 0) {
  keep  <- setdiff(seq_along(v), site_cols)
  v     <- v[keep]; X <- X[, keep, drop = FALSE]
  type  <- type[keep]; level <- level[keep]; labs <- labs[keep]; grp <- grp[keep]
  sg    <- if (length(AIARMS$site_group)) AIARMS$site_group else AIARMS$site
  sg    <- as.character(sg)
  if (length(sg) != nrow(X))
    stop("The object carries no per-household site code, so location cannot be encoded.")
  f     <- factor(sg)
  X     <- cbind(X, Site = as.integer(f))
  v     <- c(v, "Site")
  type  <- c(type, "c")
  level <- c(level, nlevels(f))
  labs  <- c(labs, "Site")
  grp   <- c(grp, "Spatial")
  colnames(X) <- labs
  site_levels <- levels(f)
} else if ("Site" %in% v && length(AIARMS$site_labels)) {
  site_levels <- AIARMS$site_labels      # object built with "categorical"
}
stopifnot(!anyNA(X), length(type) == ncol(X), length(level) == ncol(X))

cat("Fitting on", nrow(X), "observations and", ncol(X), "nodes\n")
cat("Site levels:", paste(sprintf("%d=%s", seq_along(site_levels), site_levels),
                          collapse = "  "), "\n")
print(table(type))

## ---------------------------------------------------------------------------
## 1. Estimate the pairwise MGM  (Algorithm 1 of the paper)
## ---------------------------------------------------------------------------
## One l1-penalised GLM per node (neighbourhood selection, Meinshausen &
## Buhlmann 2006): each variable is regressed on all the others using the link
## belonging to ITS OWN conditional distribution -- linear for "g", log-link
## Poisson for "p", multinomial for "c". The nonzero coefficients define that
## node's neighbourhood; the p neighbourhoods are stitched into one graph.
##
##   k = 2          pairwise interactions only
##   CV, 10 folds   lambda_s chosen by cross-validation in each regression
##   ruleReg OR     the two nodewise estimates are averaged unconditionally
##   threshold LW   the beta-min threshold tau of Equation 8
##   binarySign     signs for binary-binary and binary-continuous edges
##   scale TRUE     z-scores Gaussian nodes so one lambda penalises them fairly
##
## The argument list is the one app.R builds, in the same order.
args <- list(data = X, type = type, level = level,
             k = 2,
             lambdaSel = "CV",
             ruleReg = "OR",
             threshold = "LW",
             binarySign = TRUE, scale = TRUE,
             overparameterize = FALSE,
             pbar = FALSE)
args$lambdaFolds <- MGM_CV_FOLDS

## CV folds are random. ARUF has only 3 households, so some draws leave a
## training fold with 0 or 1 of them and glmnet refuses that node's regression.
## The app redraws the folds only when that happens, seeding each attempt with
## MGM_SEED + attempt - 1, so the first fit is always MGM_SEED. The same loop
## is used here, which is what makes the two fits identical.
seed_used <- NA_integer_
fit_once <- function() {
  for (attempt in 1:10) {
    set.seed(MGM_SEED + attempt - 1L)
    f <- tryCatch(do.call(mgm, args), error = function(e) e)
    if (!inherits(f, "error")) { seed_used <<- MGM_SEED + attempt - 1L; return(f) }
    if (!grepl("1 or 0 observations", conditionMessage(f))) stop(f)
  }
  stop(f)
}
fit_mgm <- fit_once()

wadj <- fit_mgm$pairwise$wadj
dimnames(wadj) <- list(labs, labs)
n_edges <- sum(wadj[upper.tri(wadj)] != 0)
n_pairs <- choose(ncol(X), 2)

## The same lines the app prints in its Specification panel. Compare them
## with the app before quoting any number in the report.
cat("\nSPECIFICATION (compare with the app)\n")
cat("  Observations (n):   ", nrow(X), "\n")
cat("  Nodes (p):          ", ncol(X), "\n")
cat("  lambda selected by: ", MGM_CV_FOLDS, "-fold cross-validation\n", sep = "")
cat("  Edge rule:           OR-rule\n")
cat("  Threshold:           tau (LW)\n")
cat("  Random seed used:   ", seed_used,
    if (!identical(seed_used, MGM_SEED)) "  (folds were redrawn)" else "", "\n")
cat("  Possible pairs:     ", n_pairs, "\n")
cat("  Edges estimated:    ", n_edges, "\n")
cat("  Density:            ", sprintf("%.1f%%", 100 * n_edges / n_pairs), "\n")
if ("Site" %in% labs)
  cat("  Edges on Site:      ", sum(wadj["Site", ] != 0), "of", ncol(X) - 1, "\n")
cat("  mgm version:        ", as.character(utils::packageVersion("mgm")), "\n")

## Edge list, in the format of the app's "Edge list (CSV)" download:
## weight rounded to 4 decimals; sign is "undefined" for any edge that involves
## a categorical node with more than two levels.
sign_label <- function(x) {
  out <- rep("undefined", length(x))
  out[!is.na(x) & x > 0] <- "positive"
  out[!is.na(x) & x < 0] <- "negative"
  out
}
ut <- which(upper.tri(wadj) & wadj != 0, arr.ind = TRUE)
edges <- data.frame(From = labs[ut[, 1]], To = labs[ut[, 2]],
                    Weight = round(wadj[ut], 4),
                    Sign = sign_label(fit_mgm$pairwise$signs[ut]),
                    stringsAsFactors = FALSE)
edges <- edges[order(-abs(edges$Weight)), ]
print(head(edges, 15), row.names = FALSE)
write.csv(edges, file.path(OUT_DIR, "mgm_edgelist.csv"), row.names = FALSE)

## Which nodes does location touch?
sp_lab   <- labs[grp == "Spatial"]
sp_edges <- edges[edges$From %in% sp_lab | edges$To %in% sp_lab, ]
cat("\n--- Edges involving a spatial node ---\n")
if (nrow(sp_edges)) print(sp_edges, row.names = FALSE) else
  cat("None. Location is conditionally independent of every other node here.\n")

## ---------------------------------------------------------------------------
## 2. Predictability (nodewise errors)
## ---------------------------------------------------------------------------
## An edge says a dependency exists, not how much of a node is explained.
## nCC is the normalised accuracy: gain over the intercept-only model divided
## by the maximum possible gain, so a node with a 90% majority class cannot
## look good by predicting the majority class.
pred <- predict(object   = fit_mgm,
                data     = X,
                errorCon = c("RMSE", "R2"),
                errorCat = c("CC", "nCC"))
print(pred$errors)

## R2 for continuous and count nodes, nCC for categorical ones, clamped to
## [0, 1] because nCC is negative when the model classifies worse than the
## majority-class rule. Same function as in app.R.
pred_vector <- function(errors, type) {
  pick <- function(nm) {
    hit <- which(colnames(errors) %in% c(nm, paste0("Error.", nm)))
    if (length(hit)) errors[[hit[1]]] else rep(NA_real_, nrow(errors))
  }
  r <- ifelse(type == "c", pick("nCC"), pick("R2"))
  r <- suppressWarnings(as.numeric(r))
  r[!is.finite(r)] <- 0
  pmin(pmax(r, 0), 1)
}
err_ring <- pred_vector(pred$errors, type)

## ---------------------------------------------------------------------------
## 3. Visualise
## ---------------------------------------------------------------------------
## Drawn as the app draws it: same palette, same layout seed.
grp_list <- split(seq_along(labs), grp)
pal <- c("Carriage" = "#D6455B", "Demographic" = "#F58A5E",
         "Socioeconomic" = "#FBD9C4", "WASH" = "#5AB8D4",
         "Health" = "#1F7A99", "Antibiotic" = "#8E5BC4",
         "Food/animal" = "#4FB477", "Spatial" = "#128C7E")
node_cols <- unname(pal[names(grp_list)]); node_cols[is.na(node_cols)] <- "#7E9A88"

pdf(file.path(OUT_DIR, "mgm_network.pdf"), width = 11, height = 8.5)
set.seed(MGM_SEED)
qgraph(wadj,
       edge.color = fit_mgm$pairwise$edgecolor,
       layout     = "spring", repulsion = 1.1,
       pie        = err_ring,
       pieColor   = ifelse(type == "c", "#F58A5E", "#5AB8D4"),
       groups     = grp_list,
       color      = node_cols,
       nodeNames  = labs, labels = as.character(seq_along(labs)),
       legend     = TRUE, legend.cex = 0.30,
       vsize = 4.2, esize = 14)
dev.off()

## ---------------------------------------------------------------------------
## 4. Stability via the nonparametric bootstrap  (optional; not in the app)
## ---------------------------------------------------------------------------
## Proportion of B bootstrap fits in which each edge was nonzero. Read this as
## STABILITY, not significance: l1 sampling distributions have a point mass at
## zero, so the quantiles are not confidence intervals.
##
## OFF by default. The app does not compute it, so nothing in the report
## depends on it. With lambda chosen by cross-validation and a three-household
## site (ARUF), a bootstrap sample can contain too few ARUF households for the
## Site regression, in which case resample() stops; the call is therefore
## wrapped so that a failure is reported instead of ending the knit.
RUN_BOOTSTRAP <- FALSE
if (RUN_BOOTSTRAP) {
  set.seed(MGM_SEED)
  res_obj <- tryCatch(
    resample(object = fit_mgm, data = X, nB = 100, quantiles = c(0.05, 0.95)),
    error = function(e) {
      message("Bootstrap not completed: ", conditionMessage(e)); NULL })
  if (!is.null(res_obj)) {
    pdf(file.path(OUT_DIR, "mgm_stability.pdf"), width = 9, height = 11)
    plotRes(res_obj, quantiles = c(0.05, 0.95), labels = labs)
    dev.off()
    saveRDS(res_obj, file.path(OUT_DIR, "mgm_resample.rds"))
  }
}

## ---------------------------------------------------------------------------
## 5. Save
## ---------------------------------------------------------------------------
saveRDS(list(fit = fit_mgm, X = X, vars = v, type = type, level = level,
             labels = labs, groups = grp, errors = pred$errors,
             site_levels = site_levels,
             spec = list(k = 2, lamSel = "CV", folds = MGM_CV_FOLDS,
                         rule = "OR", threshold = "LW",
                         seed = seed_used, preset = "full")),
        file.path(OUT_DIR, "fit_mgm_k2.rds"))

cat("\nWrote outputs to ./", OUT_DIR, "\n", sep = "")

```

## The Shiny app

```{r app-listing, eval=FALSE, purl=FALSE, code=readLines("app.R")}
```

---

## Notes for the write-up

1. The cleaned data contain 162 households: 171 raw rows, 7 repeat submissions removed (164), and two `ARUO` records about 62 km from the rest removed by the outlier rule (`DROP_OUTLIERS`, `OUTLIER_KM = 10`).
2. Blank answers in true tick-box blocks are read as "not ticked"; blanks in Yes/No items are missing (`STRICT_MISSING`). Missing values are filled by single imputation: the mode for categorical covariates and the median otherwise. The handwashing score is prorated over the items answered.
3. A sub-matric certificate or diploma is counted as tertiary education (`EDU_STRICT_TERTIARY = FALSE`, 25 households).
4. A month counts as cultured if any of the three markers has a readable result (`EXPOSURE_UNION`). *E. coli* is positive in about 97% of cultured months, so its count mostly reflects how many months were cultured.
5. The saved object carries seven `Site_*` indicators (`SITE_ENCODING = "dummy_full"`). The app replaces them with one categorical `Site` node with seven levels, one per study code, before fitting, which is the `Site` node used in the report. Leave the switch as `"dummy_full"`: the app expects that object. `ARUF` has only 3 households, so any edge that turns on that level should be read with caution. `NN_density` counts the other study households within 150 m.
6. Three of the five Gaussian covariates (`Months_tested`, `IncomeBand`, `MeatFreq`) are ordinal; they are declared Gaussian because `mgm` has no ordinal type.

---

