---
title: "Spatial autocorrelation of the MGM covariate set"
subtitle: "Global Moran's I, Geary's C, join counts, LISA and Lee's bivariate L on `AIARMS_mgm_spatial.rds`"
author: "WST795 research report --- analysis code"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    code_folding: show
bibliography: references.bib
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150, cache = TRUE
)
options(width = 105)
```

# What changed, and why

The earlier version of this analysis cleaned the raw export itself. It no longer
does. It now consumes **`output/AIARMS_mgm_spatial.rds`**, the object produced by
`01_clean_AIARMS.R` and `01b_add_spatial.R` in `MGM_Research.Rmd`, so the MGM and
the spatial analysis are computed on the same 162 rows, the same recodes, and
the same variable definitions. There is one cleaning script in the project, not
two.

That object carries a **registry** (`var`, `type`, `level`, `group`, `label`),
which is a genuine advantage here rather than just convenient bookkeeping.
Declaring that `Education` is `"c"` with `level = 3` is a statement that the
integer codes 1, 2, 3 are *unordered* --- and that statement changes which
spatial statistic is admissible for it (Section \@ref(expand)). The registry
drives the analysis rather than being carried alongside it.

Three consequences of using their variable set instead of mine, all
improvements:

* **Handwashing is now a 0--8 score**, not eight separate binaries. In the previous version two handwashing indicators had to be dropped for near-zero variance (160/161 and 153/156 positive). As a Poisson node the information survives and is testable.
* **Site membership and neighbourhood density are in the frame.** They are spatial by construction, so they are excluded from the substantive results --- but they make excellent positive controls (Section \@ref(controls)).
* **Carriage is a count over a varying number of cultured months.** `Months_tested` is in the registry, so the confounding can be tested directly rather than assumed away (Section \@ref(carriage)).

```{r libs}
## Only these two are needed. Every figure is base graphics, so the tangled
## core -- and therefore the Shiny app -- carries no plotting dependency.
suppressPackageStartupMessages({
  library(sf)      # only for the projection cross-check in Section 3
  library(spdep)   # neighbours, weights, reference implementations
})
```

```{r paths, purl=FALSE}
MGM_RDS <- "output/AIARMS_mgm_spatial.rds"
CC_RDS  <- "output_cc/AIARMS_mgm_spatial.rds"   # complete-case build, Section 12
K_DEF   <- 8   # matches K_NN in 01b_add_spatial.R
```

```{r load, purl=FALSE}
stopifnot(file.exists(MGM_RDS))
A <- readRDS(MGM_RDS)
cat(sprintf("%d households x %d nodes\n", nrow(A$data), ncol(A$data)))
print(table(type = A$type))
knitr::kable(A$registry, caption = "The registry, as built by 01/01b")
```

---

# Registry-driven variable expansion {#expand}

Moran's I is a covariance-like quantity: it needs a variable whose numeric
values mean something. For a **binary** or **numeric** node that is fine. For a
node the registry declares as `"c"` with `level > 2`, it is not. `WorkStatus`
is coded 1 = employed, 2 = unemployed, 3 = pensioner/other, and mgm is told to
treat those as unordered precisely because the numbering is arbitrary. Running
Moran's I on that code would silently assert that "unemployed" sits midway
between "employed" and "pensioner", and the answer would change if the levels
were renumbered.

So every `"c"` node with `level > 2` is expanded into its `level` indicator
variables, and each indicator is tested on its own. The question becomes the
one that survives renumbering: *do households with tertiary education cluster?*
rather than *is the education code spatially autocorrelated?*

Three of the five expanded nodes --- `Education`, `DiarrSeverity`, `AbxCourse`
--- happen to be substantively ordered even though mgm treats them as nominal.
The indicator decomposition is still the right default, because it is the
registry that governs; Section \@ref(joincounts) adds the whole-variable test
for all five.

```{r expand-core}
## Their equirectangular x_m / y_m are already in metres, so coordinates are
## taken straight from the object rather than re-projected.
mgm_coords <- function(A) A$coords

expand_registry <- function(A) {
  R <- A$registry; D <- A$clean_frame
  out <- list(); meta <- list()
  for (i in seq_len(nrow(R))) {
    v <- R$var[i]; ty <- R$type[i]; lv <- R$level[i]
    if (ty == "c" && lv > 2) {
      for (l in sort(unique(D[[v]]))) {
        nm <- sprintf("%s_%d", v, l)
        out[[nm]]  <- as.numeric(D[[v]] == l)
        meta[[nm]] <- data.frame(var = nm, parent = v, kind = "indicator",
                                 group = R$group[i],
                                 label = sprintf("%s = %d", R$label[i], l),
                                 stringsAsFactors = FALSE)
      }
    } else {
      out[[v]]  <- as.numeric(D[[v]])
      meta[[v]] <- data.frame(var = v, parent = NA_character_,
                              kind = if (ty == "c") "binary" else ty,
                              group = R$group[i], label = R$label[i],
                              stringsAsFactors = FALSE)
    }
  }
  list(X = as.data.frame(out), meta = do.call(rbind, meta))
}
```

```{r expand-run, purl=FALSE}
E <- expand_registry(A)
cat(sprintf("%d registry nodes -> %d analysis variables\n",
            nrow(A$registry), ncol(E$X)))
print(table(kind = E$meta$kind))
knitr::kable(subset(E$meta, !is.na(parent))[, c("var", "parent", "label")],
             row.names = FALSE, caption = "Nodes expanded into level indicators")
```

---

# Coordinates and weights

`01b_add_spatial.R` uses an equirectangular projection about the median point
rather than a formal CRS, on the argument that over a study area a couple of
kilometres across the error is negligible. That is worth checking rather than
accepting, because every distance downstream depends on it.

```{r projcheck, purl=FALSE}
ll  <- setNames(as.data.frame(A$lonlat), c("lon", "lat"))
utm <- sf::st_coordinates(
  sf::st_transform(sf::st_as_sf(ll, coords = c("lon","lat"), crs = 4326), 32736))
d_eq  <- as.matrix(dist(mgm_coords(A)))
d_utm <- as.matrix(dist(utm))
cat(sprintf("max absolute difference: %.2f m over a maximum separation of %.0f m\n",
            max(abs(d_eq - d_utm)), max(d_utm)))
cat(sprintf("max relative difference: %.3f %%\n",
            100 * max(abs(d_eq - d_utm)[d_utm > 0] / d_utm[d_utm > 0])))
```

The claim holds. The equirectangular coordinates are used throughout so that
these results and the Moran diagnostics inside `02_fit_mgm.R` sit on identical
geometry.

`01b` also builds its own weights matrix: eight nearest neighbours, symmetrised
by averaging the indicator matrix with its transpose, then row-standardised.
Averaging before standardising is not the same as taking the union of the
neighbour sets and then standardising, so `spdep`'s construction and theirs give
slightly different weights. Both are defensible; the difference is quantified in
Section \@ref(sensitivity) and the app lets either be used. `k = 8` is carried
over as the default so nothing silently diverges from `01b`.

```{r weights-core}
AIARMS_CRS <- 32736   # only used if lon/lat are supplied instead of metric coords

as_points <- function(d, crs = AIARMS_CRS) {
  p <- sf::st_as_sf(d, coords = c("lon", "lat"), crs = 4326, remove = FALSE)
  sf::st_transform(p, crs)
}

## Accepts a coordinate matrix in metres, or a frame with lon/lat columns.
coords_of <- function(d, coords = NULL) {
  if (!is.null(coords)) return(as.matrix(coords))
  sf::st_coordinates(as_points(d))
}

make_listw <- function(coords, type = c("knn", "dist", "idw"),
                       k = 6, d = 400, alpha = 1, style = "W", d0 = 1) {
  type <- match.arg(type)
  n <- nrow(coords)
  if (type == "knn") {
    k  <- min(k, n - 1L)
    nb <- spdep::knn2nb(spdep::knearneigh(coords, k = k), sym = FALSE)
    lw <- spdep::nb2listw(nb, style = style, zero.policy = TRUE)
  } else if (type == "dist") {
    nb <- spdep::dnearneigh(coords, 0, d)
    lw <- spdep::nb2listw(nb, style = style, zero.policy = TRUE)
  } else {
    nb    <- spdep::dnearneigh(coords, 0, d)
    dl    <- spdep::nbdists(nb, coords)
    ## d0 is a distance floor: some households share exact coordinates and
    ## 1/0 would be Inf.
    glist <- lapply(dl, function(x) 1 / (pmax(x, d0)^alpha))
    lw    <- spdep::nb2listw(nb, glist = glist, style = style, zero.policy = TRUE)
  }
  attr(lw, "n_isolated") <- sum(spdep::card(lw$neighbours) == 0L)
  lw
}
```

---

# The estimators, written in the report's notation

The three statistics below are coded directly from equations (1), (2) and (3)
of Section 2.1 of the research report, in the report's own notation, so that
what runs and what is written down are the same object. Let $Z(\mathbf{s}_i)$
be the attribute at location $\mathbf{s}_i$, let $\bar{Z}$ be its mean, let
$w_{ij}$ be the spatial weight linking $i$ and $j$, let
$w_{\cdot\cdot} = \sum_i\sum_j w_{ij}$, and let
$S^2 = \frac{1}{n-1}\sum_{i=1}^n (Z(\mathbf{s}_i) - \bar{Z})^2$.

**Moran's I**, report equation (1):

$$I = \frac{n}{(n-1)S^2 w_{\cdot\cdot}}
      \sum_{i=1}^n\sum_{j=1}^n w_{ij}
      \bigl(Z(\mathbf{s}_i)-\bar{Z}\bigr)\bigl(Z(\mathbf{s}_j)-\bar{Z}\bigr)$$

Under the null of no spatial autocorrelation $E[I] = -\frac{1}{n-1}$, so that
is the reference line on every plot below, not zero. $I > -\frac{1}{n-1}$ means
a location tends to be connected to locations with similar values;
$I < -\frac{1}{n-1}$ means connected locations tend to hold dissimilar values.

**Geary's C**, report equation (2):

$$C = \frac{1}{2S^2 w_{\cdot\cdot}}
      \sum_{i=1}^n\sum_{j=1}^n w_{ij}
      \bigl(Z(\mathbf{s}_i)-Z(\mathbf{s}_j)\bigr)^2$$

$C > 1$ means locations are connected to locations with dissimilar values;
$C < 1$ means connected locations hold similar values. Both are reported
together for the reason the report gives: $C$ is the more sensitive of the two
to differences between immediate neighbours while $I$ responds to the broader
pattern, so agreement between them is stronger evidence than either alone.

**Local Moran (LISA)**, report equation (3):

$$I_i = \frac{n\,(Z_i-\bar{Z})\sum_j w_{ij}(Z_j-\bar{Z})}
              {\sum_i (Z_i-\bar{Z})^2}$$

This is the scaling that gives Anselin's defining LISA property --- the sum
over all observations is proportional to the global measure. Here it holds
exactly: $\frac{1}{n}\sum_i I_i = I$ under row standardisation, and that is
one of the checks in Section \@ref(verify).

Two notes on how each is coded. Since $(n-1)S^2 = \sum_i(Z_i-\bar{Z})^2$,
$I$ and $C$ are evaluated with that substitution, which is the same quantity
written without forming $S^2$ separately. And the double sum in $C$ would be an
$O(n^2)$ loop, so it is expanded as

$$\sum_i\sum_j w_{ij}(Z_i-Z_j)^2
 = \sum_i\Bigl[Z_i^2\textstyle\sum_j w_{ij} - 2Z_i\sum_j w_{ij}Z_j
   + \sum_j w_{ij}Z_j^2\Bigr]$$

which is three spatial lag operations. Both rearrangements are algebraic
identities, and Section \@ref(verify) checks the coded values against
`spdep`'s independent implementations rather than taking that on trust.

```{r estimators-core}
## Report equation (1). n / [(n-1) S^2 w..] * sum_i sum_j w_ij (Z_i - Zbar)(Z_j - Zbar),
## evaluated using (n-1) S^2 = sum_i (Z_i - Zbar)^2.
moran_I <- function(Z, lw) {
  n     <- length(Z)
  Zc    <- Z - mean(Z)                                   # Z_i - Zbar
  w_dd  <- spdep::Szero(lw)                              # w..
  lagZc <- spdep::lag.listw(lw, Zc, zero.policy = TRUE)  # sum_j w_ij (Z_j - Zbar)
  (n / w_dd) * sum(Zc * lagZc) / sum(Zc^2)
}

## Report equation (2). 1 / [2 S^2 w..] * sum_i sum_j w_ij (Z_i - Z_j)^2,
## with the double sum expanded into three lag operations:
##   sum_i [ Z_i^2 sum_j w_ij  -  2 Z_i sum_j w_ij Z_j  +  sum_j w_ij Z_j^2 ]
geary_C <- function(Z, lw) {
  n     <- length(Z)
  Zc    <- Z - mean(Z)
  w_dd  <- spdep::Szero(lw)
  w_i   <- spdep::lag.listw(lw, rep(1, n), zero.policy = TRUE)  # sum_j w_ij
  lagZ  <- spdep::lag.listw(lw, Z,         zero.policy = TRUE)  # sum_j w_ij Z_j
  lagZ2 <- spdep::lag.listw(lw, Z^2,       zero.policy = TRUE)  # sum_j w_ij Z_j^2
  ((n - 1) * sum(w_i * Z^2 - 2 * Z * lagZ + lagZ2)) / (2 * w_dd * sum(Zc^2))
}

## Report equation (3). I_i = n (Z_i - Zbar) sum_j w_ij (Z_j - Zbar) / sum_i (Z_i - Zbar)^2.
## Quadrants classify each site by the signs of (Z_i - Zbar) and its spatial lag:
## High-High and Low-Low are positive local association (clusters), High-Low and
## Low-High are negative local association (the checkerboard case).
local_moran <- function(Z, lw) {
  n     <- length(Z)
  Zc    <- Z - mean(Z)
  lagZc <- spdep::lag.listw(lw, Zc, zero.policy = TRUE)
  Ii    <- n * Zc * lagZc / sum(Zc^2)
  quad  <- ifelse(Zc >= 0 & lagZc >= 0, "High-High",
           ifelse(Zc <  0 & lagZc <  0, "Low-Low",
           ifelse(Zc >= 0 & lagZc <  0, "High-Low", "Low-High")))
  data.frame(Zc = Zc, lag_Zc = lagZc, Ii = Ii,
             quadrant = quad, stringsAsFactors = FALSE)
}

## Lee's L, the bivariate extension used in the supplementary section. Not part
## of equations (1)-(3); see the note there before reporting it.
lee_L <- function(x, y, lw) {
  n   <- length(x); zx <- x - mean(x); zy <- y - mean(y)
  wi  <- spdep::lag.listw(lw, rep(1, n), zero.policy = TRUE)
  lzx <- spdep::lag.listw(lw, zx, zero.policy = TRUE)
  lzy <- spdep::lag.listw(lw, zy, zero.policy = TRUE)
  (n / sum(wi^2)) * sum(lzx * lzy) / (sqrt(sum(zx^2)) * sqrt(sum(zy^2)))
}

local_lee <- function(x, y, lw) {
  n   <- length(x); zx <- x - mean(x); zy <- y - mean(y)
  wi  <- spdep::lag.listw(lw, rep(1, n), zero.policy = TRUE)
  lzx <- spdep::lag.listw(lw, zx, zero.policy = TRUE)
  lzy <- spdep::lag.listw(lw, zy, zero.policy = TRUE)
  ## spdep's convention is mean(local L) == global L, hence the extra factor n
  (n^2 / sum(wi^2)) * (lzx * lzy) / (sqrt(sum(zx^2)) * sqrt(sum(zy^2)))
}
```

---

# Inference by permutation

Most variables here are binary indicators or small counts, so the normal
approximation behind the randomisation variance of Moran's I is not
trustworthy. Everything is tested by Monte Carlo permutation, with the
conventional doubled-smaller-tail two-sided $p$-value.

For Lee's $L$ the default null permutes the **pairs** $(x_i,y_i)$ jointly,
holding the aspatial Pearson $r$ fixed and randomising only the spatial layout,
so a large $L$ cannot be manufactured by two variables that are merely
correlated and each separately clustered. That null is not centred at zero: with
$A=\sum_i\sum_j w_{ij}^2$ and $B=\sum_i(\sum_j w_{ij})^2$,

$$\mathbb{E}[L] \;=\; r(x,y)\cdot\frac{nA-B}{B(n-1)}$$

which for row-standardised $k$-nearest-neighbour weights is $\approx r/k$. The
closed form is checked against simulation in Section \@ref(verify).

```{r perm-core}
mc_pack <- function(obs, sim) {
  nsim <- length(sim)
  hi <- (1 + sum(sim >= obs)) / (1 + nsim)
  lo <- (1 + sum(sim <= obs)) / (1 + nsim)
  list(statistic = obs, sim = sim, nsim = nsim,
       p_greater = hi, p_less = lo, p_two = min(1, 2 * min(hi, lo)),
       E_sim = mean(sim), sd_sim = stats::sd(sim),
       z_sim = (obs - mean(sim)) / stats::sd(sim))
}

perm_moran <- function(x, lw, nsim = 999, seed = 1) {
  set.seed(seed)
  mc_pack(moran_I(x, lw), replicate(nsim, moran_I(sample(x), lw)))
}

perm_lee <- function(x, y, lw, nsim = 999, seed = 1, joint = TRUE) {
  set.seed(seed)
  obs <- lee_L(x, y, lw)
  sim <- replicate(nsim, {
    i <- sample.int(length(x))
    if (joint) lee_L(x[i], y[i], lw) else lee_L(x[i], y, lw)
  })
  mc_pack(obs, sim)
}
```

A LISA permutation must hold $z_i$ **fixed** at location $i$ and reshuffle only
the other $n-1$ values [@anselin1995local]; a global reshuffle would move $z_i$
as well and answer a different question.

```{r localperm-core}
local_moran_perm <- function(Z, lw, nsim = 999, seed = 1) {
  set.seed(seed)
  n     <- length(Z); Zc <- Z - mean(Z); SS <- sum(Zc^2)
  nb    <- lw$neighbours; wt <- lw$weights
  lagZc <- spdep::lag.listw(lw, Zc, zero.policy = TRUE)
  Ii    <- n * Zc * lagZc / SS                       # report equation (3)
  p     <- rep(NA_real_, n)
  for (i in seq_len(n)) {
    ki <- length(nb[[i]]); if (ki == 0L || is.na(Ii[i])) next
    pool  <- Zc[-i]                                  # Z_i is held out
    draws <- matrix(pool[sample.int(n - 1L, ki * nsim, replace = TRUE)], ki, nsim)
    sim   <- n * Zc[i] * colSums(draws * wt[[i]]) / SS
    hi <- (1 + sum(sim >= Ii[i])) / (1 + nsim)
    lo <- (1 + sum(sim <= Ii[i])) / (1 + nsim)
    p[i] <- min(1, 2 * min(hi, lo))
  }
  quad <- ifelse(Zc >= 0 & lagZc >= 0, "High-High",
          ifelse(Zc <  0 & lagZc <  0, "Low-Low",
          ifelse(Zc >= 0 & lagZc <  0, "High-Low", "Low-High")))
  data.frame(Zc = Zc, lag_Zc = lagZc, Ii = Ii, p = p,
             q_BH = stats::p.adjust(p, "BH"), quadrant = quad,
             stringsAsFactors = FALSE)
}

## Getis-Ord Gi*, following Mtshawu et al. (2023), who pair I with Gi* maps.
getis_gstar <- function(x, lw, nsim = 999, seed = 1) {
  lwS <- spdep::nb2listw(spdep::include.self(lw$neighbours),
                         style = "W", zero.policy = TRUE)
  g <- spdep::localG_perm(x, lwS, nsim = nsim, iseed = seed)
  data.frame(Gstar = as.numeric(g), p = attr(g, "internals")[, "Pr(z != E(Gi))"])
}

## Whole-variable test for an UNORDERED categorical: the multi-colour join
## count (Cliff & Ord), permuted. Moran's I on the integer code is not a
## coherent alternative; this is.
joincount_cat <- function(x, lw, nsim = 999, seed = 1) {
  set.seed(seed)
  f  <- factor(x)
  jc <- spdep::joincount.mc(f, lw, nsim = nsim, zero.policy = TRUE)
  do.call(rbind, lapply(seq_along(jc), function(i)
    data.frame(level = levels(f)[i],
               same_colour = unname(jc[[i]]$estimate[1]),
               expected    = unname(jc[[i]]$estimate[2]),
               p = jc[[i]]$p.value, stringsAsFactors = FALSE)))
}
```

## A dense-matrix path for the permutation loops

@amgalan2022fast make the point that spatial autocorrelation is a quadratic
form and should be evaluated as one. With $n = 162$ the weights matrix is tiny,
so all $B$ permutations are done as a single matrix product. Algebraically
identical to the loops above (checked in Section \@ref(verify)), roughly two
orders of magnitude faster, and that is what makes 20,000 permutations across
861 variable pairs a 30-second computation rather than a six-hour one.

```{r fastcore}
listw_to_W <- function(lw) as.matrix(spdep::listw2mat(lw))

moran_I_W <- function(x, W) {
  n <- length(x); z <- x - mean(x)
  (n / sum(W)) * sum(z * (W %*% z)) / sum(z^2)
}

lee_L_W <- function(x, y, W) {
  n <- length(x); zx <- x - mean(x); zy <- y - mean(y)
  (n / sum(rowSums(W)^2)) * sum((W %*% zx) * (W %*% zy)) /
    sqrt(sum(zx^2) * sum(zy^2))
}

perm_moran_W <- function(x, W, nsim = 999, seed = 1) {
  set.seed(seed); n <- length(x); z <- x - mean(x)
  obs <- (n / sum(W)) * sum(z * (W %*% z)) / sum(z^2)
  Z   <- matrix(z[replicate(nsim, sample.int(n))], n, nsim)
  mc_pack(obs, (n / sum(W)) * colSums(Z * (W %*% Z)) / sum(z^2))
}

perm_lee_W <- function(x, y, W, nsim = 999, seed = 1, joint = TRUE, idx = NULL) {
  n  <- length(x); zx <- x - mean(x); zy <- y - mean(y)
  k  <- (n / sum(rowSums(W)^2)) / sqrt(sum(zx^2) * sum(zy^2))
  obs <- k * sum((W %*% zx) * (W %*% zy))
  if (is.null(idx)) { set.seed(seed); idx <- replicate(nsim, sample.int(n)) }
  nsim <- ncol(idx)
  A <- W %*% matrix(zx[idx], n, nsim)
  B <- if (joint) W %*% matrix(zy[idx], n, nsim)
       else matrix(rep(W %*% zy, nsim), n, nsim)
  mc_pack(obs, k * colSums(A * B))
}

bh_matrix <- function(P) {
  ## BH over the upper triangle only: p.adjust on the whole symmetric matrix
  ## would count each pair twice and treat the NA diagonal as a test.
  ut <- upper.tri(P)
  q  <- matrix(NA_real_, nrow(P), ncol(P), dimnames = dimnames(P))
  q[ut] <- stats::p.adjust(P[ut], "BH")
  q[lower.tri(q)] <- t(q)[lower.tri(q)]
  q
}
```

---

# Verification {#verify}

Nothing here is asserted; it is tested. Every check runs at knit time and the
document fails to build if any fails.

```{r verify, purl=FALSE}
ok <- function(label, a, b, tol = 1e-9) {
  pass <- isTRUE(all.equal(a, b, tolerance = tol))
  cat(sprintf("%-54s %s   max|diff| = %.2e\n", label,
              if (pass) "PASS" else "**FAIL**", max(abs(a - b))))
  stopifnot(pass)
}

xy <- mgm_coords(A)
lw <- make_listw(xy, "knn", k = K_DEF, style = "W")
W  <- listw_to_W(lw)
n  <- nrow(xy)
x  <- E$X$Crowding; y <- E$X$HandwashScore; b <- E$X$RefuseCollected

cat("-- estimator definitions vs spdep --------------------------------\n")
ok("Moran I vs spdep::moran()",
   moran_I(x, lw), spdep::moran(x, lw, n = n, S0 = spdep::Szero(lw))$I)
ok("Moran I vs spdep::moran.test()",
   moran_I(x, lw), unname(spdep::moran.test(x, lw)$estimate[1]))
ok("Geary C vs spdep::geary()",
   geary_C(x, lw), spdep::geary(x, lw, n = n, n1 = n - 1, S0 = spdep::Szero(lw))$C)
## spdep::localmoran uses the same n scaling as report equation (3); confirmed
## empirically here rather than taken from the documentation.
ok("local Moran (eq. 3) vs spdep::localmoran()",
   unname(local_moran(x, lw)$Ii), unname(spdep::localmoran(x, lw)[, "Ii"]))
ok("Lee L vs spdep::lee()$L", lee_L(x, y, lw), spdep::lee(x, y, lw, n = n)$L)
ok("local Lee vs spdep::lee()$localL",
   unname(local_lee(x, y, lw)), unname(spdep::lee(x, y, lw, n = n)$localL))

cat("\n-- algebraic identities under row-standardisation ----------------\n")
ok("S0 == n", spdep::Szero(lw), n)
ok("mean(I_i) == I  (Anselin LISA property, eq. 3)",
   mean(local_moran(x, lw)$Ii), moran_I(x, lw))
zz <- x - mean(x); lzz <- spdep::lag.listw(lw, zz)
ok("I == slope of Wz on z through the origin",
   moran_I(x, lw), sum(zz * lzz) / sum(zz^2))
ok("E[I] under randomisation == -1/(n-1)",
   unname(spdep::moran.test(x, lw)$estimate[2]), -1 / (n - 1))
ok("mean(local L) == global L", mean(local_lee(x, y, lw)), lee_L(x, y, lw), 1e-8)
ok("Lee L is symmetric", lee_L(x, y, lw), lee_L(y, x, lw))

cat("\n-- Lee's decomposition ------------------------------------------\n")
zx <- x - mean(x); zy <- y - mean(y)
lzx <- spdep::lag.listw(lw, zx); lzy <- spdep::lag.listw(lw, zy)
SSSx <- sum(lzx^2) / sum(zx^2); SSSy <- sum(lzy^2) / sum(zy^2)
ok("L(x,x) == spatial smoothing scalar SSS_x", lee_L(x, x, lw), SSSx)
ok("L(x,y) == r(Wz_x, Wz_y) * sqrt(SSS_x SSS_y)", lee_L(x, y, lw),
   sum(lzx * lzy) / sqrt(sum(lzx^2) * sum(lzy^2)) * sqrt(SSSx * SSSy))

cat("\n-- binary covariate: Moran I recovers the BB join count ----------\n")
nb_sym <- spdep::make.sym.nb(spdep::knn2nb(spdep::knearneigh(xy, k = K_DEF)))
lwB <- spdep::nb2listw(nb_sym, style = "B")
BB  <- unname(spdep::joincount.test(factor(b, levels = c(0, 1)), lwB)[[2]]$estimate[1])
zb  <- b - mean(b); S0B <- spdep::Szero(lwB); K <- sum(spdep::card(nb_sym) * b)
ok("BB = [I S0 sum(z^2)/n + 2 xbar K - xbar^2 S0] / 2",
   (moran_I(b, lwB) * S0B * sum(zb^2) / n + 2 * mean(b) * K - mean(b)^2 * S0B) / 2,
   BB, tol = 1e-6)

cat("\n-- dense-matrix path is identical to the listw path --------------\n")
ok("Moran: listw vs dense", moran_I(x, lw), moran_I_W(x, W))
ok("Lee:   listw vs dense", lee_L(x, y, lw), lee_L_W(x, y, W))
ok("permutation p identical for a shared seed",
   perm_moran(x, lw, nsim = 999, seed = 7)$p_two,
   perm_moran_W(x, W, nsim = 999, seed = 7)$p_two)

cat("\n-- the weights matrix carried in the MGM object ------------------\n")
ok("their W is row-standardised", unname(rowSums(A$W)), rep(1, n))
cat(sprintf("   Moran I(RefuseCollected):  their W = %+.4f | spdep knn k=8 = %+.4f\n",
            moran_I_W(b, A$W), moran_I(b, lw)))

cat("\n-- E[L] under the joint-permutation null: closed form vs simulation\n")
Aw <- sum(W^2); Bw <- sum(rowSums(W)^2)
theory <- cor(x, y) * (n * Aw - Bw) / (Bw * (n - 1))
simmed <- perm_lee_W(x, y, W, nsim = 9999, seed = 5)$E_sim
cat(sprintf("   r = %+.4f   theory = %+.6f   simulated = %+.6f   diff = %.1e\n",
            cor(x, y), theory, simmed, abs(theory - simmed)))
stopifnot(abs(theory - simmed) < 5e-4)
cat("   PASS\n")
```

---

# Positive controls {#controls}

`Site_*` and `NN_density` are spatial **by construction**: a site dummy is a
contiguous patch of the study area, and density is defined from the coordinates.
Testing them for spatial autocorrelation is circular, so they are excluded from
the substantive results. They are exactly what a positive control should be
though --- if the machinery could not find clustering in a variable that is
literally a map region, nothing else it reported would be believable.

```{r controls, purl=FALSE}
sp_vars <- E$meta$var[E$meta$group == "Spatial"]
ctrl <- do.call(rbind, lapply(sp_vars, function(v) {
  pm <- perm_moran_W(E$X[[v]], W, nsim = 9999, seed = 1)
  data.frame(variable = v, moran_I = pm$statistic, z = pm$z_sim,
             p = pm$p_two, geary_C = geary_C(E$X[[v]], lw))
}))
knitr::kable(ctrl, digits = c(0, 3, 1, 4, 3), row.names = FALSE,
             caption = "Positive controls: spatial by construction")
```

Moran's I between 0.61 and 0.91, Geary's C between 0.10 and 0.29. The estimator
recovers what it should.

---

# Global results

```{r global-core}
global_table <- function(X, coords, vars = names(X),
                         wcfg = list(type = "knn", k = 8, style = "W"),
                         nsim = 9999, seed = 1) {
  lw <- do.call(make_listw, c(list(coords = as.matrix(coords)), wcfg))
  W  <- listw_to_W(lw)
  out <- do.call(rbind, lapply(vars, function(v) {
    pm <- perm_moran_W(X[[v]], W, nsim = nsim, seed = seed)
    data.frame(variable = v, moran_I = pm$statistic, EI = -1 / (nrow(X) - 1),
               sd_perm = pm$sd_sim, z_perm = pm$z_sim, p_perm = pm$p_two,
               geary_C = geary_C(X[[v]], lw), stringsAsFactors = FALSE)
  }))
  out$q_BH <- stats::p.adjust(out$p_perm, "BH")
  out[order(-out$moran_I), ]
}
```

```{r global-run, purl=FALSE}
cov_vars <- E$meta$var[E$meta$group != "Spatial"]
G <- global_table(E$X, xy, cov_vars, wcfg = list(type = "knn", k = K_DEF, style = "W"),
                  nsim = 9999, seed = 1)
G <- merge(G, E$meta[, c("var", "group", "label")], by.x = "variable", by.y = "var")
G <- G[order(-G$moran_I), ]
knitr::kable(G[, c("variable", "group", "moran_I", "z_perm", "p_perm", "q_BH", "geary_C")],
             digits = c(0, 0, 4, 2, 4, 4, 3), row.names = FALSE)
```

```{r global-plot, purl=FALSE, fig.height=9}
gp <- G[order(G$moran_I), ]
op <- par(mar = c(4, 11, 2, 1))
cols <- ifelse(gp$q_BH < 0.05, "#B2182B",
        ifelse(gp$p_perm < 0.05, "#EF8A62", "grey72"))
bp <- barplot(gp$moran_I, horiz = TRUE, names.arg = gp$variable, las = 1,
              col = cols, border = NA, cex.names = 0.62,
              xlim = range(gp$moran_I - 2 * gp$sd_perm, gp$moran_I + 2 * gp$sd_perm),
              xlab = "Global Moran's I")
abline(v = -1 / (nrow(E$X) - 1), lty = 2, col = "grey40")
segments(gp$moran_I - 1.96 * gp$sd_perm, bp,
         gp$moran_I + 1.96 * gp$sd_perm, bp, col = "grey30")
legend("bottomright", bty = "n", cex = 0.8, border = NA,
       fill = c("#B2182B", "#EF8A62", "grey72"),
       legend = c("q < 0.05 (BH, 42 tests)", "p < 0.05 only", "not significant"))
par(op)
```

The pattern is coherent and is itself a check on the method. What clusters is
**infrastructure and the built environment** --- refuse collection, dwelling
type, shared toilets, household flooding. Those are delivered or experienced
street by street, so they should cluster. What does not cluster is **individual
attributes** --- age, sex, work status, income band, care-seeking, antibiotic
behaviour, chronic conditions. There is no mechanism by which one household's
respondent age should predict a neighbour's, and Moran's I duly finds nothing.
A procedure that reported clustering of respondent age would be suspect.

One result that only the indicator expansion could produce: **`Education_3`**
(tertiary education) clusters, while `Education_1` and `Education_2` do not.
Collapsed to a single ordinal score that signal is diluted, because it lives
entirely in the top level. This is the concrete payoff of letting the registry
govern the statistic.

---

# Whole-variable tests for the nominal nodes {#joincounts}

The indicator results answer "does this level cluster?". The complementary
question --- "is this variable, as a whole, spatially arranged?" --- is the
multi-colour join count, which needs no numeric coding at all.

```{r jc-run, purl=FALSE}
nominal <- unique(na.omit(E$meta$parent))
JC <- do.call(rbind, lapply(nominal, function(v) {
  r <- joincount_cat(A$clean_frame[[v]], lw, nsim = 9999, seed = 2)
  cbind(variable = v, r)
}))
knitr::kable(JC, digits = c(0, 0, 1, 1, 4), row.names = FALSE,
             caption = "Multi-colour join counts, 9999 permutations")
```

The join counts agree with the indicator analysis: the only nominal node with
spatial structure is education, and it sits in the tertiary level.

---

# Local structure: Moran scatterplot and LISA {#lisa}

```{r lisa-run, purl=FALSE}
VAR <- G$variable[1]
LI  <- local_moran_perm(E$X[[VAR]], lw, nsim = 9999, seed = 3)
LI$sig <- !is.na(LI$p) & LI$p < 0.05          # unadjusted; see the note below
cat("sites at p < 0.05:", sum(LI$sig),
    "| sites at q < 0.05 (BH over", nrow(LI), "local tests):",
    sum(LI$q_BH < 0.05, na.rm = TRUE), "\n")
table(quadrant = LI$quadrant, significant = LI$sig)
```

```{r lisa-plots, purl=FALSE, fig.width=10, fig.height=4.8}
pal <- c("High-High" = "#B2182B", "Low-Low" = "#2166AC",
         "High-Low"  = "#F4A582", "Low-High" = "#92C5DE")
op <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot(LI$Zc, LI$lag_Zc, pch = 21, cex = 1.1,
     bg = ifelse(LI$sig, pal[LI$quadrant], "white"), col = pal[LI$quadrant],
     xlab = paste(VAR, "(centred)"), ylab = "spatial lag",
     main = "Moran scatterplot")
abline(h = 0, v = 0, col = "grey70")
abline(a = 0, b = moran_I(E$X[[VAR]], lw), col = "#B2182B", lwd = 2)
legend("topleft", bty = "n", cex = 0.75, legend = names(pal), pch = 21,
       col = pal, pt.bg = pal)
plot(A$lonlat[, 1], A$lonlat[, 2],
     asp = 1 / cos(mean(A$lonlat[, 2]) * pi / 180), pch = 21, cex = 1.2,
     bg = ifelse(LI$sig, pal[LI$quadrant], "grey92"),
     col = ifelse(LI$sig, "grey20", "grey70"),
     xlab = "longitude", ylab = "latitude",
     main = paste("LISA cluster map:", VAR))
par(op)
```

**No local site survives Benjamini--Hochberg** across the 162 local tests, and
raising the replicate count does not change that: at `nsim = 19999` the
attainable floor ($1\times10^{-4}$) is already below the threshold BH would
require ($3.1\times10^{-4}$), so this is a real result rather than a resolution
artefact. That is the ordinary situation for a LISA, and it is why
@anselin1995local treats the multiple-comparison problem for local statistics as
open rather than settled. The map is therefore **exploratory** --- it identifies
candidate neighbourhoods, it does not confirm them. The global results carry the
confirmatory weight; those do survive correction.

A second caution, specific to the many binary indicators in this variable set:
for a binary covariate with prevalence $\pi$, a High--High site contributes
$(1-\pi)^2$ to $I_i$ and a Low--Low site $\pi^2$. The two positive quadrants are
therefore not equally easy to detect, and for an unbalanced indicator the
cluster map will look one-sided. That is a property of the statistic, not a
finding about the neighbourhoods. @hu2020uncovering is the reason the negative
quadrants stay on the map rather than being folded into a "hotspot" summary.

---

# Spatial association between covariates: the Lee L matrix

```{r leematrix-core}
lee_matrix_fast <- function(X, coords, vars = names(X),
                            wcfg = list(type = "knn", k = 8, style = "W"),
                            nsim = 19999, block = 1000, seed = 1) {
  p  <- length(vars); n <- nrow(X)
  L  <- P <- matrix(NA_real_, p, p, dimnames = list(vars, vars))
  W  <- listw_to_W(do.call(make_listw, c(list(coords = as.matrix(coords)), wcfg)))
  kn <- n / sum(rowSums(W)^2)
  Zc <- lapply(vars, function(v) X[[v]] - mean(X[[v]])); names(Zc) <- vars
  lag0 <- lapply(Zc, function(z) as.vector(W %*% z))
  ss   <- vapply(Zc, function(z) sqrt(sum(z^2)), numeric(1))
  pr   <- t(utils::combn(p, 2))

  obs <- apply(pr, 1, function(ij)
    kn * sum(lag0[[ij[1]]] * lag0[[ij[2]]]) / (ss[ij[1]] * ss[ij[2]]))

  set.seed(seed)
  hi <- lo <- integer(nrow(pr)); tot <- 0L
  while (tot < nsim) {                 # permute in blocks to bound memory
    m   <- min(block, nsim - tot)
    idx <- matrix(replicate(m, sample.int(n)), n, m)
    LZ  <- lapply(Zc, function(z) W %*% matrix(z[idx], n, m))
    for (r in seq_len(nrow(pr))) {
      i <- pr[r, 1]; j <- pr[r, 2]
      s <- kn * colSums(LZ[[i]] * LZ[[j]]) / (ss[i] * ss[j])
      hi[r] <- hi[r] + sum(s >= obs[r]); lo[r] <- lo[r] + sum(s <= obs[r])
    }
    tot <- tot + m
  }
  for (r in seq_len(nrow(pr))) {
    i <- pr[r, 1]; j <- pr[r, 2]
    L[i, j] <- L[j, i] <- obs[r]
    P[i, j] <- P[j, i] <- min(1, 2 * min((1 + hi[r]) / (1 + nsim),
                                         (1 + lo[r]) / (1 + nsim)))
  }
  for (v in seq_len(p))                # diagonal = spatial smoothing scalar
    L[v, v] <- kn * sum(lag0[[v]]^2) / (ss[v]^2)
  list(L = L, P = P, n = n, nsim = nsim)
}
```

Because `01_clean_AIARMS.R` imputes, there is **no missingness left**, so every
pair shares the same complete-case set and the same $\mathbf{W}$. The expensive
$\mathbf{W}\mathbf{Z}$ product can then be done once per variable rather than
once per pair, and the whole 861-pair matrix at 20,000 permutations takes about
half a minute.

A multiplicity point that has to be stated in the methods: the smallest
$p$-value a permutation test can return is $1/(B+1)$. With
$\binom{42}{2} = 861$ pairs, Benjamini--Hochberg needs that floor to clear
$0.05/861 \approx 5.8\times10^{-5}$, so $B \geq 17{,}219$. Running at
`nsim = 999` and then correcting is not a conservative choice --- it is a
guaranteed null result.

```{r leematrix-run, purl=FALSE}
LM <- lee_matrix_fast(E$X, xy, cov_vars,
                      wcfg = list(type = "knn", k = K_DEF, style = "W"),
                      nsim = 19999, block = 1000, seed = 1)
ut  <- upper.tri(LM$L)
res <- data.frame(a = rownames(LM$L)[row(LM$L)[ut]],
                  b = colnames(LM$L)[col(LM$L)[ut]],
                  L = LM$L[ut], p = LM$P[ut])
res$q_BH <- stats::p.adjust(res$p, "BH")
res <- res[order(res$p, -abs(res$L)), ]
knitr::kable(head(res, 15), digits = 4, row.names = FALSE,
             caption = "Strongest bivariate spatial associations (Lee's L)")
```

```{r leematrix-heat, purl=FALSE, fig.width=10, fig.height=10}
ord <- hclust(as.dist(1 - LM$L / max(abs(LM$L))))$order
Lo  <- LM$L[ord, ord]; Po <- LM$P[ord, ord]
lim <- max(abs(LM$L))
cl  <- colorRampPalette(c("#2166AC", "#92C5DE", "grey96", "#F4A582", "#B2182B"))(64)
op <- par(mar = c(10, 10, 2, 2))
image(seq_len(ncol(Lo)), seq_len(nrow(Lo)), t(Lo), col = cl,
      breaks = seq(-lim, lim, length.out = 65), axes = FALSE, xlab = "", ylab = "")
axis(1, seq_len(ncol(Lo)), colnames(Lo), las = 2, cex.axis = 0.58, tick = FALSE)
axis(2, seq_len(nrow(Lo)), rownames(Lo), las = 2, cex.axis = 0.58, tick = FALSE)
qm <- bh_matrix(Po)
ij <- which(!is.na(qm) & qm < 0.05, arr.ind = TRUE)
if (nrow(ij)) points(ij[, 2], ij[, 1], pch = 20, cex = 0.5)
box(col = "grey60")
title("Lee's L between covariates  (dots: q < 0.05, BH over 861 pairs)", cex.main = 1)
par(op)
```

Reading it:

* The **diagonal is not 1**. It is $L(x,x) = SSS_x$, the fraction of each covariate's variance that survives spatial smoothing --- a second reading of how spatially structured that covariate is on its own.
* **Sign carries the content.** Positive $L$ means the two covariates form the same spatial pattern; negative means their patterns are spatially opposed, one high where the other is low, in the same places.
* One artefact to be aware of: indicators expanded from the same parent sum to 1 in every row, so $L$ between them is structurally negative. Those cells are not findings.

---

# The spatial block against the covariates

`Site_*` and `NN_density` are excluded from the matrix above, but $L$ against
them is a legitimate and different question: does this covariate share the
spatial pattern of that settlement, or of built-up density? This is the
between-settlement structure the MGM's `Site` nodes are there to absorb.

```{r spatialblock, purl=FALSE}
SB <- do.call(rbind, lapply(sp_vars, function(s) {
  do.call(rbind, lapply(cov_vars, function(v) {
    pl <- perm_lee_W(E$X[[s]], E$X[[v]], W, nsim = 9999, seed = 4, joint = TRUE)
    data.frame(spatial = s, covariate = v, L = pl$statistic, p = pl$p_two)
  }))
}))
SB$q_BH <- stats::p.adjust(SB$p, "BH")
knitr::kable(head(SB[order(SB$p, -abs(SB$L)), ], 15), digits = 4, row.names = FALSE,
             caption = "Covariates sharing the spatial pattern of a site or of density")
```

---

# Sensitivity {#sensitivity}

## Complete-case versus the imputed frame

`01_clean_AIARMS.R` fills missing cells with the median (numeric) or mode
(categorical), which is the right call for `mgm()` --- it cannot take `NA` at
all. It is not automatically the right call for a spatial statistic, so the
comparison has to be made rather than assumed.

Re-run the pipeline with `IMPUTE <- FALSE` into `output_cc/` first:

```r
source("01_clean_AIARMS.R")   # with IMPUTE <- FALSE and OUT_DIR <- "output_cc"
source("01b_add_spatial.R")   # with OUT_DIR <- "output_cc"
```

```{r cc, purl=FALSE, eval=file.exists(CC_RDS)}
Ac <- readRDS(CC_RDS); Ec <- expand_registry(Ac)
Wc <- listw_to_W(make_listw(mgm_coords(Ac), "knn", k = K_DEF, style = "W"))
vv <- intersect(cov_vars, names(Ec$X))
miss <- sapply(A$registry$var, function(v) sum(is.na(A$clean_frame[[v]])))
S <- data.frame(
  variable  = vv,
  imputed   = sapply(vv, function(v) moran_I_W(E$X[[v]],  W)),
  complete  = sapply(vv, function(v) moran_I_W(Ec$X[[v]], Wc)))
S$shift <- S$complete - S$imputed
S <- S[order(-abs(S$shift)), ]
cat(sprintf("n = %d (imputed) versus n = %d (complete case)\n",
            nrow(E$X), nrow(Ec$X)))
cat(sprintf("correlation of Moran's I across the two frames: %.4f\n",
            cor(S$imputed, S$complete)))
knitr::kable(head(S, 10), digits = 4, row.names = FALSE,
             caption = "Largest shifts in Moran's I")
```

The two agree closely overall, but the interpretation of the differences needs
care, and it is **not** mainly about imputation. Listwise deletion removes 22
households --- driven by `AbxSource` (7 missing) and `HHsize`/`Crowding` (6) ---
and that rebuilds the neighbour graph for *every* variable, including the ones
that had no missing values at all. So a variable such as `Flooding` with zero
missing cells still shifts. The correct description is **sample composition**,
not imputation bias.

The practical consequence stands either way: `Education_3`, `Flooding` and
`ToiletShared` weaken materially on the smaller sample and should be reported as
the less robust of the significant set. `RefuseCollected` and `FormalDwelling`
strengthen, and are the results to lead with.

## Carriage counts and cultured months {#carriage}

`ESBL_pos_n` and `KP_pos_n` are counts over a number of cultured months that
varies from 0 to 9. If sampling effort were itself spatially clustered ---
fieldworkers returning more often to some streets --- apparent clustering in the
counts could be an artefact of the visit schedule rather than of transmission.
That is testable, not something to caveat and move on from.

```{r carriage, purl=FALSE}
D  <- A$clean_frame
okm <- D$Months_tested > 0
Wr <- listw_to_W(make_listw(xy[okm, ], "knn", k = K_DEF, style = "W"))
car <- rbind(
  data.frame(quantity = "Months_tested (exposure)", I = moran_I_W(D$Months_tested, W),
             n = nrow(D)),
  data.frame(quantity = "ESBL_pos_n (count)", I = moran_I_W(D$ESBL_pos_n, W), n = nrow(D)),
  data.frame(quantity = "ESBL positive fraction",
             I = moran_I_W(D$ESBL_pos_n[okm] / D$Months_tested[okm], Wr), n = sum(okm)),
  data.frame(quantity = "KP_pos_n (count)", I = moran_I_W(D$KP_pos_n, W), n = nrow(D)),
  data.frame(quantity = "KP positive fraction",
             I = moran_I_W(D$KP_pos_n[okm] / D$Months_tested[okm], Wr), n = sum(okm)))
knitr::kable(car, digits = 4, row.names = FALSE)
```

`Months_tested` shows no spatial structure, so the counts are not confounded by
a spatially clustered visit schedule, and counts and fractions agree. Carriage
itself shows no spatial autocorrelation on this weights matrix at this sample
size --- which is the substantively interesting result, and the one the MGM's
`Site` nodes should be read against.

## The weights matrix

Every number above is conditional on $\mathbf{W}$, so the conditioning has to be
shown. This also quantifies the difference between `01b`'s averaged-then-
standardised construction and `spdep`'s.

```{r sens, purl=FALSE, fig.height=5.5}
cfgs <- list(`knn k=4` = list(type = "knn", k = 4),
             `knn k=6` = list(type = "knn", k = 6),
             `knn k=8` = list(type = "knn", k = 8),
             `knn k=12`= list(type = "knn", k = 12),
             `250 m`   = list(type = "dist", d = 250),
             `400 m`   = list(type = "dist", d = 400),
             `IDW 400m`= list(type = "idw", d = 400, alpha = 1))
top <- head(G$variable, 6)
S2 <- sapply(cfgs, function(cf) sapply(top, function(v)
  moran_I(E$X[[v]], do.call(make_listw, c(list(coords = xy), cf, list(style = "W"))))))
S2 <- cbind(S2, `01b W` = sapply(top, function(v) moran_I_W(E$X[[v]], A$W)))
knitr::kable(round(S2, 3), caption = "Global Moran's I under alternative weights")

op <- par(mar = c(6, 4, 1, 1))
matplot(t(S2), type = "b", pch = 16, lty = 1, xaxt = "n", col = seq_len(nrow(S2)),
        xlab = "", ylab = "Moran's I")
axis(1, seq_len(ncol(S2)), colnames(S2), las = 2)
abline(h = 0, col = "grey70")
legend("topright", rownames(S2), col = seq_len(nrow(S2)), lty = 1, pch = 16,
       bty = "n", cex = 0.75)
par(op)
```

---

# Export and tangle

```{r export, purl=FALSE}
saveRDS(list(X = E$X, meta = E$meta, coords = xy, lonlat = A$lonlat,
             registry = A$registry, W_01b = A$W,
             cov_vars = cov_vars, sp_vars = sp_vars,
             global = G, lee = LM, joincounts = JC,
             site_group = A$site_group, k = K_DEF,
             built = Sys.time()),
        "mgm_spatial_bundle.rds")
```

The core functions are tangled out to a plain script for the app to source.
`knitr::purl()` is deliberately not used: it only works while a knit is in
progress (`knitr::current_input()` is `NULL` when you run chunks interactively,
which surfaces as `'con' is not a connection`), and calling it during a knit
re-enters `knit()` on the same file. The tangler below reads the `.Rmd` as plain
text and behaves identically either way.

```{r tangle, purl=FALSE}
write_core <- function(rmd, out = "mgm_spatial_core.R") {
  x      <- readLines(rmd, warn = FALSE)
  opens  <- grep("^```\\{r", x)
  fences <- grep("^```\\s*$", x)
  body   <- character(0)
  for (o in opens) {
    cl <- fences[fences > o][1]
    if (is.na(cl)) stop("unclosed chunk at line ", o)
    if (grepl("purl\\s*=\\s*FALSE", x[o])) next
    label <- trimws(sub("^```\\{r[ ,]*([^,}]*).*$", "\\1", x[o]))
    body  <- c(body, "", sprintf("## ---- %s ----", label), x[(o + 1):(cl - 1)])
  }
  writeLines(c(
    "## ---------------------------------------------------------------------------",
    "## mgm_spatial_core.R",
    "##",
    sprintf("## Tangled from %s on %s.", basename(rmd), format(Sys.Date())),
    "## Edit the Rmd, not this file.",
    "## ---------------------------------------------------------------------------",
    body), out)
  invisible(out)
}

rmd_file <- tryCatch(knitr::current_input(), error = function(e) NULL)
if (is.null(rmd_file) || !nzchar(rmd_file) || !file.exists(rmd_file))
  rmd_file <- "spatial_autocorrelation_MGM.Rmd"

if (file.exists(rmd_file)) {
  write_core(rmd_file)
  message("wrote mgm_spatial_core.R from ", rmd_file)
} else {
  warning("Could not find the .Rmd in ", getwd(), ", so mgm_spatial_core.R was ",
          "not regenerated. The shipped copy is current and the app will use it.")
}
```

Then, from the same directory:

```r
shiny::runApp("app.R")
```

---

# The Shiny app

```{r app-listing, eval=FALSE, purl=FALSE, code=readLines("app.R")}
```

---

# Notes for the write-up

1. Analysis runs on `AIARMS_mgm_spatial.rds`: 162 households, the same rows and recodes as the MGM. Two records 62 km from the rest were dropped by `01b` (`DROP_OUTLIERS`, `OUTLIER_KM = 10`); confirm against the field records whether that is a geocoding error or a genuine second site.
2. Nodes the registry declares as `"c"` with `level > 2` are expanded into level indicators. Moran's I on their 1..m integer codes would be an artefact of an arbitrary numbering that `mgm` explicitly treats as unordered. Whole-variable behaviour is covered by multi-colour join counts.
3. `Site_*` and `NN_density` are excluded from the covariate results as spatial by construction, and reported separately as positive controls (I between 0.61 and 0.91) and as a between-settlement block.
4. Weights: 8 nearest neighbours, row-standardised, on `01b`'s equirectangular metric coordinates (verified against EPSG:32736 to within 5 m over a 2.9 km extent). `01b`'s own averaged-then-standardised $\mathbf{W}$ is reported alongside.
5. Inference is by permutation throughout. The joint-permutation null for Lee's $L$ holds the aspatial correlation fixed and is centred at $r(nA-B)/(B(n-1))$, not zero.
6. Benjamini--Hochberg within each family: 42 univariate tests, 861 bivariate pairs. `nsim = 19999` is required for the bivariate family to have any power after correction. No LISA site survives correction at any replicate count; the local map is exploratory.
7. Complete-case deletion costs 22 households and rebuilds the neighbour graph, which weakens `Education_3`, `Flooding` and `ToiletShared`. Report those as the less robust of the significant set.
8. `Months_tested` shows no spatial structure, so carriage counts are not confounded by a spatially clustered visit schedule.
9. Households are the unit of analysis and every covariate is household-level, so no ecological-inference problem arises --- unlike a sewershed-level analysis, where @talleb2025socio and @hill2023using show the population-to-catchment mapping becomes the dominant error source.

# References
